### 2015-05-26 missing.Rmd
missing

---
title: "Pairwise-complete correlation considered dangerous"
author: "B. W. Lewis"
date: "5/25/2015"
output: html_document
---

R includes facilities for handling missing values for all native data types. Perhaps counterintuitively, marking a value as missing conveys the information that the value is not known. Donald Rumsfeld might call it a "known unknown^[https://goo.gl/gGYcJe]." Upon encountering a missing value, we can deal with it by simply omitting it, imputing it somehow, or through several other possible approaches. R does a good job of making our choice of missing value approach explicit.

The `cov` and `cor` functions in the R programming language include several approaches for dealing with missing data. The `use="pairwise.complete.obs"` option is particularly confusing, and can easily lead to faulty comparisons. This note explains and warns against its use.

Consider the following really simple example:
```{r}
(x = matrix(c(-2,-1,0,1,2,1.5,2,0,1,2,NA,NA,0,1,2),5))
```

The functions `cov(x)` computes the symmetric covariance matrix `V` with entries defined by the pairwise covariance of columns of `x`,
```{r,eval=FALSE}
V[i,j] = cov(x[,i], x[,j])
```
where `i=j=c(1,2,3)` in this example. The function `cor(x)` similarly computes the symmetric correlation matrix with entries defined by pairwise correlation of the columns of `x`. For example:
```{r}
cov(x)
cor(x)
```
Due to missing values in the third column of `x` we know that we don't know the covariance between `x[,3]` and anything else. Thanks to an arguably questionable^[The `cov2cor` always puts ones along the diagonal; that choice is valid only if all unknowns may assume bounded numeric values.] choice in R's `cov2cor` function, we see that `cor(x[,3],x[,3])=1`, but we don't know the correlation between `x[,3]` and the other columns.

The `use="complete"` option is one reasonable alternative way to deal with missing values. It simply removes rows of the matrix `x` with one or more missing observations. Since the third through fifth rows of our example matrix
are all identical we expect perfect correlation across the board, and indeed:
```{r}
cor(x, use="complete")
```
is the same in our example as omitting the first two rows because they contain missing values:
```{r}
cor(x[3:5,])
```

The `use="pairwise.complete.obs"` is a far less reasonable way to deal with missing values. When specified, R computes correlations for each pair of columns using vectors formed by omitting rows with missing values on a case by case basis. Thus each column vector may vary depending on it's pairing, resulting in correlation values that are not even comparable. Consider our simple example again:
```{r}
cor(x, use="pairwise.complete.obs")
```
By this bizarre measurement, the correlation of `x[,1]` and `x[,2]` is zero (as we saw above in the first example), and yet `cor` claims that `x[,3]` is perfectly correlated with _both_ `x[,1]` and `x[,2]`. In other words, the result is nonsense. What's going on here is that the reported correlations are not comparable because they are computed against different vectors: all of `x[,1]` and `x[,2]` are compared to each other, but only parts of `x[,1]` and `x[,2]` are compared to `x[,3]`.

The bad result is obvious for our small example. But the danger here is in large matrices with lots of missing values, where it may be impossible to use the `pairwise` option in any meaningful way.

## Recommendations

If you want to run correlations on lots of vectors with missing values, consider using the `use="complete"` option.

If you really don't want to do that for some reason, consider imputing the missing values. The simplest available replaces missing values in each column with the mean of the non-missing values in the respective column:
```{r}
m = mean(na.omit(x[,3]))
xi = x
xi[is.na(x)] = m
cor(xi)
```

Sometimes, it might make more sense to use a piecewise constant interpolant, sometimes referred to as "last
observation carry forward" especially when dealing with time series and ordered data. In yet other cases a known default value (perhaps from a much larger population than the one under study) might be more appropriate.

A more expensive approach bootstraps the missing values from the non-missing ones, either per column or from all the values if you expect them to be similarly distributed. You can decide by examining the empirical distributions of the non-missing data in each column.
```{r}
i = is.na(x[,3])
N = sum(i)
b = replicate(500, {x[i,3] = sample(x[!i,3], size=N, replace=TRUE);cor(x[,1:2],x[,3])})

# Average imputed values of cor(x[,1],x[,3]) and cor(x[,2],x[,3])
apply(b,1,mean)
# Standard deviation of imputed values of cor(x[,1],x[,3]) and cor(x[,2],x[,3])
apply(b,1,sd)
```

If you have lots of observations, consider partitioning them with a basic clustering algorithm first and then imputing the missing values from their respective cluster cohorts.

## References and packages

There are of course many excellent R packages and references on missing data. I recommend consulting the following packages and references:

* http://www.stat.columbia.edu/~gelman/arm/missing.pdf
* http://cran.r-project.org/web/packages/mi/
* http://gking.harvard.edu/amelia

### 2015-05-26 concordance-notes.R

@@ -1,0 +1,31 @@
+# Set up an example matrix x with sampled iid rows and known covariance structure
+v = qr.Q(qr(matrix(rnorm(100),10)))
+d = exp(-(1:10))
+sigma = v %*% diag(d) %*% t(v)
+library("mvtnorm")
+x = rmvnorm(n=500,mean=1:10,sigma=sigma)
+
+# Generate some row samples
+set.seed(1)
+i = lapply(rep(seq(from=2,to=500,by=1),1), function(j) list(j=j,idx=sample(500,j)))
+
+# I think that 'concordance' is trying to quantify this
+# kind of relationship:
+nrm = "2"
+ncx = norm(cov(x),nrm)
+a = t(sapply(i, function(j) c(j$j,norm(cov(x[j$idx,]),nrm))))
+#plot(a[,1], abs(a[,2] - ncx)/ncx, main="rel difference")
+
+# or maybe this ratio relationship:
+plot(a[,1], abs(a[,2]/norm(cov(x),nrm)), main="ratio")
+
+
+# A somewhat crude bound for this is:
+s = svd(x)
+alpha = t(sapply(i, function(j) c(j$j,(499/(j$j-1))*norm(crossprod(s$u[j$idx,]),nrm))))
+
+lines(alpha[,1],alpha[,2], col=2,lwd=2)
+
+# Note that althought ratio looks like 1/x, it's not. You can see this by plotting:
+# plot(alpha[,1], 1/alpha[,2])
+# That funky shape is from the || U_k^T U_k|| term (see the LaTeX notes).

### 2015-05-26 missing.Rmd
missing

@@
 ```{r,eval=FALSE}
-V[i,j] = cov(x[,i], x[,j])
+V[i,j] = crossprod(x[,i] - mean(x[,i]), x[,j] - mean(x[,j])) / nrow(x)
 ```

### 2015-05-26 missing.Rmd
missing

--- 2015-05-26 missing.Rmd
@@
 
-The `cov` and `cor` functions in the R programming language include several approaches for dealing with missing data.
+The `cov` and `cor` functions in the R programming language include several options for dealing with missing data.
 The `use="pairwise.complete.obs"` option is particularly confusing, and can easily lead to faulty comparisons.
@@
 
-The functions `cov(x)` computes the symmetric covariance matrix `V` with entries defined by the pairwise covariance of columns of `x`,
-```{r,eval=FALSE}
-V[i,j] = cov(x[,i], x[,j])
-```
-where `i=j=c(1,2,3)` in this example. The function `cor(x)` similarly computes the symmetric correlation matrix with entries defined by pairwise correlation of the columns of `x`. For example:
+The functions `V=cov(x)` computes the symmetric covariance matrix `V` with entries defined by the pairwise covariance of columns of `x`,
+$$
+\mathsf{V}[i,j] = covariance (\mathsf{x}[,i], \,\,\,\mathsf{x}[,j]),
+$$
+where $i=j=1,2,3$ in this example. The function `cor(x)` similarly computes the symmetric correlation matrix with entries defined by pairwise correlation of the columns of `x`. For example:
 ```{r}

### 2015-05-26 missing.Rmd
missing

@@
 If you really don't want to do that for some reason, consider imputing the missing values.
-The simplest available replaces missing values in each column with the mean of the non-missing values in the respective column:
+The simplest available method replaces missing values in each column with the mean of the non-missing values in the respective column:
 ```{r}
@@
 ```
+This can be done really efficiently when "centering" a matrix by simply replacing missing values of the centered matrix with zero.
 

### 2015-05-26 missing.Rmd
missing

@@
 
-Consider the following really simple example:
+Consider the following tiny example:
 ```{r}
@@
 
-The `use="complete"` option is one reasonable alternative way to deal with missing values.
-It simply removes rows of the matrix `x` with one or more missing observations.
-Since the third through fifth rows of our example matrix
+The `use="complete"` option is one way to deal with missing values. It simply removes rows of the matrix `x` with missing observations. Since the third through fifth rows of our example matrix
 are all identical we expect perfect correlation across the board, and indeed:
@@
 ```
-is the same in our example as omitting the first two rows because they contain missing values:
-```{r}
-cor(x[3:5,])
-```
+Reasonable people might question this approach.
+Deleting two observations has a huge effect on the correlation between `x[,1]` and `x[,2]` in this example because those columns exhibit large variance.
+The result really says that we should collect more observations!
 
-The `use="pairwise.complete.obs"` is a far less reasonable way to deal with missing values.
-When specified, R computes correlations for each pair of columns using vectors formed by omitting rows with missing values on a case by case basis.
+## Unknown knowns
+The `use="pairwise.complete.obs"` is a less reasonable way to deal with missing values.
+When specified, R computes correlations for each pair of columns using vectors formed by omitting rows with missing values on a _pairwise_ basis.
 Thus each column vector may vary depending on it's pairing, resulting in correlation values that are not even comparable.
@@
 In other words, the result is nonsense.
+As Rumsfeld might say, we've converted known unknowns into unknown knowns.
+
 What's going on here is that the reported correlations are not comparable because they are computed against different vectors: all of `x[,1]` and `x[,2]` are compared to each other, but only parts of `x[,1]` and `x[,2]` are compared to `x[,3]`.
 
-The bad result is obvious for our small example. But the danger here is in large matrices with lots of missing values, where it may be impossible to use the `pairwise` option in any meaningful way.
+The bad result is obvious for our small example. But the danger here is in large matrices with lots of missing values, where it may be impossible to use the `pairwise` option in a meaningful way.
 
@@
 
-If you want to run correlations on lots of vectors with missing values, consider using the `use="complete"` option.
+If you want to run correlations on lots of vectors with missing values, consider simply using the R default of `use="everything"` and propagating missing values into the correlation matrix.
+This makes it clear what you don't know.
 
-If you really don't want to do that for some reason, consider imputing the missing values.
+If you really don't want to do that, consider imputing the missing values.
 The simplest available method replaces missing values in each column with the mean of the non-missing values in the respective column:

### 2015-05-26 missing.Rmd
missing

@@
 ## Unknown knowns
-The `use="pairwise.complete.obs"` is a less reasonable way to deal with missing values.
+The `use="pairwise.complete.obs"` is an even less reasonable way to deal with missing values.
 When specified, R computes correlations for each pair of columns using vectors formed by omitting rows with missing values on a _pairwise_ basis.
@@
 
-Sometimes, it might make more sense to use a piecewise constant interpolant, sometimes referred to as "last
+Sometimes it might make more sense to use a piecewise constant interpolant, referred to as "last
 observation carry forward" especially when dealing with time series and ordered data.

### 2015-05-26 missing.Rmd
missing

@@
 Reasonable people might question this approach.
-Deleting two observations has a huge effect on the correlation between `x[,1]` and `x[,2]` in this example because those columns exhibit large variance.
+Deleting two observations has a huge effect on the correlation between `x[,1]` and `x[,2]` in this example mostly because of the large change in `x[,1]`.
 The result really says that we should collect more observations!
@@
 
-If you really don't want to do that, consider imputing the missing values.
-The simplest available method replaces missing values in each column with the mean of the non-missing values in the respective column:
+If you really don't want to do that, consider imputing the missing values. The simplest method replaces missing values in each column with the mean of the non-missing values in the respective column:
 ```{r}
@@
 
-A more expensive approach bootstraps the missing values from the non-missing ones, either per column or from all the values if you expect them to be similarly distributed.
-You can decide by examining the empirical distributions of the non-missing data in each column.
+Another basic approach bootstraps the missing values from the non-missing ones:
 ```{r}
@@
 If you have lots of observations, consider partitioning them with a basic clustering algorithm first and then imputing the missing values from their respective cluster cohorts.
+Or consider a matching method or a regression-based imputation method.
+See the references below for many more details.
 

### 2015-05-26 missing.Rmd
Updated missing example

@@
 ---
+
+This note warns about potentially misleading results when using the
+`use=pairwise.complete.obs` and `use=complete` options in R's `cor`
+and `cov` functions. The note illustrates potential issues with
+a very simple pathological example.
+
+## Known unknowns
 
@@
 ```
-Due to missing values in the third column of `x` we know that we don't know the covariance between `x[,3]` and anything else.
-Thanks to an arguably questionable^[The `cov2cor` always puts ones along the diagonal; that choice is valid only if all unknowns may assume bounded numeric values.] choice in R's `cov2cor` function, we see that `cor(x[,3],x[,3])=1`, but we don't know the correlation between `x[,3]` and the other columns.
+Due to missing values in the third column of `x` we know that we don't know the
+covariance between `x[,3]` and anything else. Thanks to an arguably
+questionable^[The `cov2cor` function used by `cor` always puts ones along the
+diagonal of a correlation matrix; that choice is valid only if all unknowns may
+assume bounded and valid numeric values which is actually pretty reasonable. But worse and weirder than that, R is inconsistent: `cor(x[,3],x[,3])` returns `NA` (!!!!)] choice in R's
+`cov2cor` function, we see that `cor(x[,3],x[,3])=1`, but we don't know the
+correlation between `x[,3]` and the other columns.
 

### 2015-05-27 missing.Rmd
Updated missing note

@@
 diagonal of a correlation matrix; that choice is valid only if all unknowns may
-assume bounded and valid numeric values which is actually pretty reasonable. But worse and weirder than that, R is inconsistent: `cor(x[,3],x[,3])` returns `NA` (!!!!)] choice in R's
-`cov2cor` function, we see that `cor(x[,3],x[,3])=1`, but we don't know the
+assume bounded and valid numeric values which is actually pretty reasonable. But in a rare example of inconsistency in R `cor(x[,3],x[,3])` returns `NA`. Yikes!)] choice in R's
+`cov2cor` function, R reports that 
+the correlation of `x[,3]` with itself is one, but we don't know the
 correlation between `x[,3]` and the other columns.

### 2015-05-27 missing.Rmd
references

@@
 This note warns about potentially misleading results when using the
-`use=pairwise.complete.obs` and `use=complete` options in R's `cor`
-and `cov` functions. The note illustrates potential issues with
-a very simple pathological example.
+`use=pairwise.complete.obs` and related options in R's `cor`
+and `cov` functions. The note illustrates potential issues using
+a very simple pathological example and lists alternative approaches
+to handling missing data and references about them.
 
@@
 * http://gking.harvard.edu/amelia
+* http://cran.r-project.org/web/views/OfficialStatistics.html  (Look for the "Imputation" section in this task view for a list of R packages.)

### 2015-05-27 missing.Rmd
references

@@
 `use=pairwise.complete.obs` and related options in R's `cor`
-and `cov` functions. The note illustrates potential issues using
-a very simple pathological example and lists alternative approaches
-to handling missing data and references about them.
+and `cov` functions. Pitfalls are illustrated using
+a very simple pathological example followed by a brief list
+of alternative ways to deal with missing data and some references about them.
 
@@
 
-R includes facilities for handling missing values for all native data types.
+R includes excellent facilities for handling missing values for all native data types.
 Perhaps counterintuitively, marking a value as missing conveys the information that the value is not known.

### 2015-05-30 missing.Rmd
typo

@@
 
-The `use="complete"` option is one way to deal with missing values. It simply removes rows of the matrix `x` with missing observations. Since the third through fifth rows of our example matrix
+The `use="complete"` option is one way to deal with missing values.
+It simply removes rows of the matrix `x` with missing observations.
+Since the columns of the  third through fifth rows of our example matrix
 are all identical we expect perfect correlation across the board, and indeed:

### 2015-07-21 threshold/draft.Rnw
Added threshold project

%%
%% Instructions for processing this document into a PDF document
%%
%% FIRST NOTE, you need the 'biclust' package installed in R.
%% Some of the examples will come from data sets in this package.
%% You can install it from the R console with, for example:
%%
%% install.packages("biclust",repos="http://cran.case.edu")
%%
%% Now you can prepare the PDF of the paper with any of:
%%
%% From a shell:
%% R CMD Sweave --pdf draft.Rnw
%%
%% From an R prompt:
%% Sweave("draft.Rnw")
%% tools::texi2pdf("draft.tex")
%%
%% From RStudio:
%% click on the little "Compile PDF" button
%%
%% Note that you should probably run the above a few times because
%% LaTeX generally takes a few runs to get reference numbers right.
%%
\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{definition}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{example}[1][Example]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{remark}[1][Remark]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcounter{algorithmctr}
\newenvironment{algorithm}{
   \refstepcounter{algorithmctr}
   \bigskip\noindent
   \textbf{Algorithm \thealgorithmctr\\}
}
{\bigskip}
\numberwithin{algorithmctr}{section}

\newcommand{\qed}{\nobreak \ifvmode \relax \else
      \ifdim\lastskip<1.5em \hskip-\lastskip
      \hskip1.5em plus0em minus0.5em \fi \nobreak
      \vrule height0.75em width0.5em depth0.25em\fi}

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{B. Lewis, A. Poliakov}

\date{}

\begin{document}
\SweaveOpts{concordance=TRUE}
\maketitle

\begin{abstract}

Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance, and other fields
can produce correlation matrices too large to explicitly form and tracticably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.

\end{abstract}

\section{Introduction}\label{intro}

XXX add discussion of applications (especially genomics?) and previous work,
then on to the preliminaries:
\[\]

Consider a real-valued data matrix $A$ consisting of $m$ observations of $n$
column vectors.  Denote the columns of $A$ as $a_j\in\mathcal{R}^m$ for
$j=1,2,\ldots,n$, $A=[a_1, a_2, \ldots, a_n]\in\mathcal{R}^{m\times n}$, and
let $k=\mathrm{rank}(A)$.  Assume that the mean of each column $a_j$ is zero
and that each column has unit norm $\|a_j\|^2 = 1$. Here and below, $\|\cdot\|$
denotes the Euclidean vector norm. Then the sample correlation matrix
$\mathrm{cor}(A)=A^TA$.  Note that under these assumptions the sample
correlation and covariance matrices are the same.  Section \ref{examples} below
illustrates how to relax the unit norm and zero mean assumptions in practice.

Let $A=U\Sigma V^T$ be the singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $U^TU = I$, $V\in\mathcal{R}^{n\times k}$,
$V^TV=I$, and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  Denote
the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$. The span
of the columns of $V$ form an orthonormal basis of the range of the correlation
matrix $A^TA$.

Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix $A$
(including possibly the case $i=j$). Denote the correlation of these two vectors
by $\mathrm{cor}(a_i,a_j) = a_j^T a_i$.  The following simple lemma
from Rafiei\cite{raf} establishes a relationship between correlation and
Euclidean distance between two vectors.

\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{cordist}
\mathrm{cor}(a_i,a_j) = a_i^Ta_j = 1 - \|a_i - a_j\|^2/2.
\end{equation}
In particular, for a given correlation threshold $t$, $0<t<1$, pairs
of vectors $a_i$ and $a_j$ such that $\|a_i - a_j\|^2 > 2(1-t)$ lie
outside the correlation threshold.
\end{lemma}
\begin{proof} (XXX perhaps omit this)
\begin{eqnarray*}
\|a_i - a_j\|^2 &=& (a_i - a_j)^T(a_i - a_j) \\
&=& a_i^T a_i + a_j^T a_j - 2a_i^T a_j \\
&=& \|a_i\|^2 + \|a_j\|^2 - 2a_i^T a_j \\
&=& 2 - 2a_i^T a_j.
\end{eqnarray*}
The lemma follows by solving for $a_i^Ta_j$.
\end{proof}

Equation \ref{cordist} equates the problem of finding pairs of highly
correlated column vectors from the matrix $A$ with a problem of finding paris
of vectors sufficiently close together. For example, correlation values larger
than 0.99 may only be associated with column vectors from the matrix $A$ whose
Euclidean distance is less than $\sqrt{0.02}$.  Vectors farther apart will not
meet the correlation threshold.

The following lemma expresses the Euclidean distance between any two columns of the
matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 |v_{1i} - v_{1j}|^2 + 
\sigma_2^2 |v_{2i} - v_{2j}|^2 + \cdots + 
\sigma_k^2 |v_{ki} - v_{kj}|^2,
\end{equation}
where $v_{kj}$ denotes the $jth$ position in the vector $v_k$ (that is, the
$[k,j]$ entry in the matrix $V$).
\end{lemma}

%%\begin{proof}
%%\begin{align*}
%%\|a_i - a_j\|^2 &= (a_i - a_j)^T(a_i - a_j) \\
%%&= a_i^Ta_i + a_j^T a_j - 2a_i^Ta_j \\
%%&= \sigma_1^2 v_{1i}^2 + \sigma_2^2 v_{2i}^2 + \cdots + \sigma_k^2 v_{ki}^2 \\
%%&\phantom{=} + \sigma_1^2 v_{1j}^2 + \sigma_2^2 v_{2j}^2 + \cdots + \sigma_k^2 v_{kj}^2 \\
%%&\phantom{=} -2(\sigma_1^2 v_{1i}v_{1j} + \sigma_2^2 v_{2i}v_{2j} + \cdots + \sigma_k^2 v_{ki}v_{kj}) \\
%%&= \sigma_1^2 (v_{1i}^2 + v_{1j}^2 - 2v_{1i}v_{1j}) \\
%%&\phantom{=} + \sigma_2^2 (v_{2i}^2 + v_{2j}^2 - 2v_{2i}v_{2j}) \\
%%&\phantom{=} + \cdots \\
%%&\phantom{=} + \sigma_k^2 (v_{ki}^2 + v_{kj}^2 - 2v_{ki}v_{kj}) \\
%%&= \sigma_1^2 |v_{1i} - v_{1j}|^2
%% + \sigma_2^2 |v_{2i} - v_{2j}|^2
%% + \cdots
%% + \sigma_k^2 |v_{ki} - v_{kj}|^2.
%%\end{align*}
%%\end{proof}

Note that each term in the sum is nonnegative and the weights $\sigma_j^2$,
$j=1,2,\ldots,k$ are nonincreasing and defined by the covariance structure of
the data.  Given a correlation threshold $t$, $0<t<1$, Equations \ref{cordist}
and \ref{vdist} together suggest that pairs of vectors can be excluded from
consideration whenever their projected distance is too large.  For example if
$\sigma_1^2 |v_{1i} - v_{1j}|^2 \ge 2(1-t)$, then we can conclude from just two
scalar values that the column vectors $a_i$ and $a_j$ do not meet the
correlation threshold $t$.

In practice, many pairs of vectors may be ruled out in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.  Approaches that
cheaply omit pairs of vectors from consideration are called \emph{pruning
methods}, and a variety pruning methods have been described elsewhere for the
thresholded correlation problem; see, for example\cite{zf,mnl} and the
references therein.  The approach described here prunes along an intuitive path
of decreasing variance defined by the data and computed by the SVD.

\section{Efficient pruning using the truncated SVD}

Let $t$ be a given correlation threshold $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation that orders then entries of
$v_1$ in increasing order.  It follows from Equations \ref{cordist} and
\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |v_{1i} - v_{1j}|^2
\ge 2(1-t)$ then the column vectors $a_j$ and $a_i$ do not meet the correlation
threshold.  Let $\ell$ be the longest run of successive points in $P v_1$
within the interval $\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1}
displays the entries of $v_1$ for the ``BicatYeast'' example from Section
\ref{examples} in increasing order.  The lines in the plot illustrate the
interval associated with the correlation threshold $t=0.99$. The quantity
$\ell$ is obtained by rolling the interval across all the points in the plot
and counting the maximum number of points that lie within the interval.  With
respect to the ordering $P$, $\ell$ represents the biggest index difference
that paris of indices corresponding to correlated vectors can exhibit.

\begin{figure}[!ht]
\label{fig1}
\begin{center}
<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
library(biclust)
data(BicatYeast)
A=t(BicatYeast)
s=svd(A)
i=order(s$v[,1])
plot(s$v[i,1], ylab=expression(Pi * v[1]), xlab="Sorted index", pch='.')
t=0.99
abline(h=c(-0.05,-0.05+sqrt(2*(1-t))/s$d[1]),col=2,lwd=0.5)
@
\caption{Example plot of the ordered entries of $v_1$ for the BicatYeast example from Section \ref{examples}. The red lines illustrate the interval corresponding threshold $t=0.99$. Points farther apart than this interval correspond to column vectors that do not meet the correlation threshold.}
\end{center}
\end{figure}

Let $D^i\in\mathcal{R}^{n-i, n}$, $i=1,2,\ldots,n-1$, be the $i$th order finite
difference matrix with diagonal entries $-1$, $i$th super-diagonal entries $1$,
and zeros elsewhere.  Consider just adjacent points of $P v_1$.  Then $D^1P
v_1$ consists of the differences of adjacent entries in $P v_1$.  Entries
such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$, where the exponentiation is
applied element-wise, correspond to pairs of vectors that do not meet
the correlation threshold. This is unlikely to rule out many pairs of vectors
in most problems. For example, the minimum distance between the points shown in
the example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma$, and no
pruning at all occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
pairs of vectors outside the correlation threshold. For example, including
only one more term finds that only 116 out of 418 pairs of adjacent vectors
in the example shown in Figure \ref{fig1} satisfy
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 \le 2(1-t),
\]
where exponentiation is applied element-wise to the vectors.
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including five terms reduces the total to only 11 pairs of vectors.
In general, using $p\le k$ terms boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 \mathrm{diag}(\Sigma_{1:p}) \le 2(1-t),
\end{equation}
where 
$V_{1:p}$ and $\Sigma_{1:p}$ denote the first $p$ columns of the
matrices $V$ and $\Sigma$, respectively,
$(D^1PV_{1:p})^2$ denotes element-wise exponentiation, 
and $\mathrm{diag}(\Sigma)$ denotes the diagonal of the matrix $\Sigma$.

Pairs of vectors beyond the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=1,2,\ldots,\ell$ in the above expressions.
Following this procedure, we can produce a well-pruned candidate set of
possible vector pairs that meet a given correlation threshold using $\ell$
matrix vector products with matrices of order $n \times p$, where $p$ is chosen
in practice such that $p<<k$. The steps of this procedure are outlined in the
following algorithm:

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$, maximum
subspace dimension $p<<n$.
\begin{enumerate}
\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\Sigma_{1:p}$, $V_{1:p} = [v_1, v_2, \ldots, v_p]$.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 \mathrm{diag}(\Sigma_{1:p}) \le 2(1-t)
\right\},
\]
where $e_{\tilde{j}}\in\mathcal{R}^n$ consists of $1$ in position $\tilde{j}$ and
zeros otherwise.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
step 1. Continue to increase $p$ in this way until the number of candidate
pairs stops decreasing.
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

Algorithm \ref{prune} by construction guarantees that no pruned vector pair
meets the given correlation threshold; that is, the pruning can not produce any
``false negatives.''  The algorithm typically prunes the vast majority of pairs
of vectors outside the correlation threshold with truncated singular value
decompositions of only relatively low dimension $p$.  The choice of $p$ is a
balance between work in computing a truncated SVD and work required by the
$\ell$ matrix vector products of order $n\times p$ in step 4.

\section{Efficient truncated SVD for general matrices}

We have so far assumed that the mean of each column vector of $A$ is zero and
the columns are scaled to have norm 1, but we really want a method that works
with general matrices $A$. We've also implicitly assumed that we can
efficiently compute a relatively low-dimensional truncated SVD of potentially
large data matrices required by step 1 of Algorithm \ref{prune}.  Fortunately,
both needs are met by one algorithm.

The augmented, implicitly restarted Lanczos bidiagonalization (IRLBA) method of
Baglama and Reichel \cite{br} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
The R implementation includes efficient methods for restarting the algorithm
to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
efficient methods for computing the SVD of scaled and centered matrices.

\section{Numerical experiments}\label{examples}

%%<<eval=TRUE,echo=FALSE>>=
%%library(irlba)
%%
%%print("R code here")
%%@

\section{Conclusion}\label{conclusion}

%% XXX switch to bibtex

\begin{thebibliography}{99}
\bibitem{br}
Baglama, J. and Reichel, L., Augmented Implicitly Restarted Lanczos Bidiagonalization Methods, SIAM J. Sci. Comput. 2005.
\bibitem{mnl}
Mueen, A., Nath, S., and Liu, J., Fast approximate correlation for
massive time-series data, in Proceedings of the 2010 ACM SIGMOD International
Conference on Management of data, 2010, pp. 171-182.
\bibitem{raf}
Rafiei, D., On similarity-based queries for time series data,
in Data Engineering Proceedings., IEEE, 1999.
\bibitem{zf}
Zhang, J. and Feigenbaum, J, Finding Highly Correlated Pairs Efficiently with Powerful Pruning,
in Proceedings of ACM Conference on Information and Knowledge Management (CIKM), 2006, pp. 152-161.
\bibitem{irlba-R}
\url{https://cran.r-project.org/web/packages/irlba/index.html}
\bibitem{irlba-P}
\url{https://github.com/bwlewis/irlbpy}
\bibitem{irlba-M}
\url{http://www.math.uri.edu/~jbaglama/software/irlba.m}
\end{thebibliography}
\end{document}

### 2015-07-21 threshold/threshold_notes.R

@@ -1,0 +1,55 @@
+library(s4vd)    # for the lung data
+library(biclust) # for the EisenYeast and BicatYeast data
+library(zoo)
+#data(lung200)
+#A = t(lung200)
+#data(EisenYeast)
+#A = t(EisenYeast)
+data(BicatYeast)
+A = t(BicatYeast)
+
+C = cor(A) # for reference
+
+# Explicitly center and scale the data matrix
+A = sweep(A,2,apply(A,2,mean))
+A = sweep(A,2,sqrt(apply(A,2,crossprod)),FUN=`/`)
+s = svd(A)
+
+t = 0.99 # correlation threshold
+
+# permutation ordering of the first basis vector
+i = order(s$v[,1])
+
+# Naive and inefficient way to find the longest run of 1-d values within
+# the correlation tolerance (IMPROVE ?).
+width = 1000
+limit = sqrt(2*(1-t))/s$d[1]   # 1-d tolerance
+group_sizes = rollapply(s$v[i,1],width=width, FUN=function(x) sum(x - x[1] < limit), align="left", partial=TRUE)
+max_group_size = max(group_sizes)
+# if max_group_size=width, then we mis-estimated. Increase width and try again until
+# max_group_size < width to be sure that we've found the maximum.
+
+cat("Maximum group size",max_group_size,"\n")
+
+
+# This plot shows each vector projected down to a point and ordered in
+# increasing order...
+plot(s$v[i,1], main="1-d ordered", ylab="v")
+# ...and the interval beyond which we can say vectors corresponding to those
+# points won't be correlated above the threshold value, here shown at zero.
+abline(h=c(0,sqrt(2*(1-t))/s$d[1]),col=2)
+
+
+
+
+# This plot shows which sequentially-adjacent points (lag=1) fall above/below
+# the threshold and are therefore correlated or not. Each point in this plot
+# represents the difference between a *pair* of ordered points (a pair of
+# projected vectors).
+k = 5    # subspace dimension
+d = diff(s$v[i,1:k,drop=FALSE],lag=1)^2 %*% s$d[1:k]^2
+ylim = c(0,max(max(d), 1.01*2*(1-t)))
+plot(d,ylim=ylim)
+abline(h=2*(1 - t),col=2,lwd=1)
+j = which(d < 2*(1-t))
+points(j,d[j],pch=19,col="#ff000044")

### 2015-07-21 threshold/draft.Rnw
a few typos

@@
 $V^TV=I$, and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
-diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  Denote
-the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
+diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$. 
+For later convenience, we write $s=\mathrm{diag}(\Sigma)$ to be the
+diagonal vector of the matrix $\Sigma$.
+Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
 $v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
@@
 respect to the ordering $P$, $\ell$ represents the biggest index difference
-that paris of indices corresponding to correlated vectors can exhibit.
+that paris of indices corresponding to correlated vectors above the threshold
+can exhibit.
 
@@
 \begin{equation}\label{proj}
-(D^1 P V_{1:p} )^2 \mathrm{diag}(\Sigma_{1:p}) \le 2(1-t),
+(D^1 P V_{1:p} )^2 s_{1:p}) \le 2(1-t),
 \end{equation}
 where 
-$V_{1:p}$ and $\Sigma_{1:p}$ denote the first $p$ columns of the
-matrices $V$ and $\Sigma$, respectively,
-$(D^1PV_{1:p})^2$ denotes element-wise exponentiation, 
-and $\mathrm{diag}(\Sigma)$ denotes the diagonal of the matrix $\Sigma$.
+$V_{1:p}$ denotes the first $p$ columns of the
+matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values,
+and $(D^1PV_{1:p})^2$ denotes element-wise exponentiation.
 
@@
 \begin{enumerate}
-\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\Sigma_{1:p}$, $V_{1:p} = [v_1, v_2, \ldots, v_p]$.
+\item Compute the truncated SVD $AV_{1:p} = U_{1:p}s_{1:p}$, $V_{1:p} = [v_1, v_2, \ldots, v_p]$.
 \item Compute permutation $P$ that orders points in $v_1$.
@@
 (\tilde{j},\tilde{j}+\tilde{i}) : 
-e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 \mathrm{diag}(\Sigma_{1:p}) \le 2(1-t)
+e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p} \le 2(1-t)
 \right\},

### 2015-07-21 threshold/draft.Rnw
a few typos
Notation fixes

@@
 $V^TV=I$, and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
-diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$. 
-For later convenience, we write $s=\mathrm{diag}(\Sigma)$ to be the
-diagonal vector of the matrix $\Sigma$.
-Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
+diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
+later convenience, we write $s=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$ to be
+the vector of the first $p\le k$ singular values along the diagonal of
+$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
 $v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
@@
 \begin{equation}\label{proj}
-(D^1 P V_{1:p} )^2 s_{1:p}) \le 2(1-t),
+(D^1 P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t),
 \end{equation}
@@
 matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values,
-and $(D^1PV_{1:p})^2$ denotes element-wise exponentiation.
+and exponentiation is applied element-wise.
 
@@
 \begin{enumerate}
-\item Compute the truncated SVD $AV_{1:p} = U_{1:p}s_{1:p}$, $V_{1:p} = [v_1, v_2, \ldots, v_p]$.
+\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
+where $V_{1:p} = [v_1, v_2, \ldots, v_p]$
+and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
 \item Compute permutation $P$ that orders points in $v_1$.
@@
 (\tilde{j},\tilde{j}+\tilde{i}) : 
-e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p} \le 2(1-t)
+e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
 \right\},

### 2015-07-21 threshold/draft.Rnw
authors

@@
 \title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
-\author{B. Lewis, A. Poliakov}
+\author{Some dudes}
 
@@
 $\mathrm{cor}(A)=A^TA$.  Note that under these assumptions the sample
-correlation and covariance matrices are the same.  Section \ref{examples} below
+correlation and covariance matrices are the same.  Section \ref{irlba} below
 illustrates how to relax the unit norm and zero mean assumptions in practice.
@@
 
-\section{Efficient truncated SVD for general matrices}
+\section{Fast truncated SVD for general matrices}\label{irlba}
 
@@
 efficiently compute a relatively low-dimensional truncated SVD of potentially
-large data matrices required by step 1 of Algorithm \ref{prune}.  Fortunately,
+large  matrices required by step 1 of Algorithm \ref{prune}.  Fortunately,
 both needs are met by one algorithm.
@@
 available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
-The R implementation includes efficient methods for restarting the algorithm
+The R implementation includes efficient methods for re-starting the algorithm
 to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and

### 2015-07-22 threshold/draft.Rnw
Added experimental/prototype implementation; a few typos in the paper

@@
 
-XXX add discussion of applications (especially genomics?) and previous work,
+XXX add discussion of applications (genomics?) and previous work,
 then on to the preliminaries:
@@
 diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
-later convenience, we write $s=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$ to be
+later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$ to be
 the vector of the first $p\le k$ singular values along the diagonal of
@@
 of vectors $a_i$ and $a_j$ such that $\|a_i - a_j\|^2 > 2(1-t)$ lie
-outside the correlation threshold.
+below the correlation threshold.
 \end{lemma}
@@
 i=order(s$v[,1])
-plot(s$v[i,1], ylab=expression(Pi * v[1]), xlab="Sorted index", pch='.')
+plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
 t=0.99
@@
 Let $D^i\in\mathcal{R}^{n-i, n}$, $i=1,2,\ldots,n-1$, be the $i$th order finite
-difference matrix with diagonal entries $-1$, $i$th super-diagonal entries $1$,
-and zeros elsewhere.  Consider just adjacent points of $P v_1$.  Then $D^1P
-v_1$ consists of the differences of adjacent entries in $P v_1$.  Entries
-such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$, where the exponentiation is
-applied element-wise, correspond to pairs of vectors that do not meet
-the correlation threshold. This is unlikely to rule out many pairs of vectors
-in most problems. For example, the minimum distance between the points shown in
+difference matrix with $-1$ along the diagonal, $1$ along the $i$th
+super-diagonal, and zeros elsewhere.  Consider just adjacent points of $P v_1$.
+Then $D^1P v_1$ consists of the differences of adjacent entries in $P v_1$.
+Entries such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$
+correspond to pairs of vectors that do not meet the correlation threshold,
+(exponentiation is applied to the vector element-wise).
+However, this observation is unlikely to rule out many pairs of vectors in
+most problems. For example, the minimum distance between the points shown in
 the example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma$, and no
@@
 Including more terms from Equation \ref{vdist} increases our ability to prune
-pairs of vectors outside the correlation threshold. For example, including
-only one more term finds that only 116 out of 418 pairs of adjacent vectors
-in the example shown in Figure \ref{fig1} satisfy
+pairs of vectors below the correlation threshold. For example, including
+only one more term finds that 72\% of adjacent vectors fall below the
+correlation threshold
+in the example shown in Figure \ref{fig1} and are pruned by identifying
+indices such that
 \[
 \sigma_1^2 (D^1 P v_1)^2 +
-\sigma_2^2 (D^1 P v_2)^2 \le 2(1-t),
+\sigma_2^2 (D^1 P v_2)^2 > 2(1-t),
 \]
@@
 Note that we use the permutation $P$ defined by the order of vector $v_1$
-throughout. Including five terms reduces the total to only 11 pairs of vectors.
-In general, using $p\le k$ terms boils down to evaluating the expression
+throughout. Including five terms prunes over 97\% of adjacent vector pairs,
+leaving only 11 candidate pairs of adjacent vectors.
+In general, using $p\le k$ terms to prune pairs of vectors below the
+correlation threshold boils down to evaluating the expression
 \begin{equation}\label{proj}
-(D^1 P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t),
+(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
 \end{equation}
@@
 matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values,
-and exponentiation is applied element-wise.
+and exponentiation is applied to vectors element-wise.
 
-Pairs of vectors beyond the correlation threshold that lie more than one index
+Pairs of vectors below the correlation threshold that lie more than one index
 difference apart relative to the permutation $P$ can be similarly pruned by
@@
 \item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
-\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$
+\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
 \[
@@
 step 1. Continue to increase $p$ in this way until the number of candidate
-pairs stops decreasing.
+pairs stops decreasing (the best this algorithm can do).
 \item Recover the non-permuted candidate column vector
@@
 ``false negatives.''  The algorithm typically prunes the vast majority of pairs
-of vectors outside the correlation threshold with truncated singular value
+of vectors below the correlation threshold with truncated singular value
 decompositions of only relatively low dimension $p$.  The choice of $p$ is a
@@
 with general matrices $A$. We've also implicitly assumed that we can
-efficiently compute a relatively low-dimensional truncated SVD of potentially
+efficiently compute a relatively low-rank truncated SVD of potentially
 large  matrices required by step 1 of Algorithm \ref{prune}.  Fortunately,
@@
 
-The augmented, implicitly restarted Lanczos bidiagonalization (IRLBA) method of
-Baglama and Reichel \cite{br} efficiently computes truncated singular value
+The augmented, implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
+of Baglama and Reichel \cite{br} efficiently computes truncated singular value
 decompositions of large dense or sparse matrices. Reference implementations are
@@
 to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
-efficient methods for computing the SVD of scaled and centered matrices.
+efficient methods for computing the SVD of scaled and centered matrices without
+explicitly forming a new matrix.
 

### 2015-07-22 threshold/tcor.R

@@ -1,0 +1,83 @@
+# EXPERIMENTAL PROTOTYPE FUNCTION
+#
+# As of 20-July, 2015 this needs the development version of irlba available
+# from GitHub: devtools::install_github("bwlewis/IRL")
+require(irlba)
+require(zoo)   # for rollapply
+
+#' Compute the thresholded correlations between columns of a matrix.
+#'
+#' Increase p to cut down the total number of candate pairs.
+#'
+#' @param A an m by n real-valued dense or sparse matrix
+#' @param t a threshold value for correlation
+#' @param p projected subspace dimension
+#'
+#' @return
+#' A two-column matrix. Each row of the matrix are the
+#' indices of candidate vectors that meet the correlation threshold.
+tcor = function(A, t=0.99, p=5)
+{
+  mu = colMeans(A)
+  s  = sqrt(apply(A,2,crossprod))
+  L  = irlba(A, p, center=mu, scale=s)
+  P  = order(L$v[,1])  # order the entries of v1 (the permutation in the paper)
+  limit = sqrt(2*(1-t))/L$d[1]
+# XXX this is stupid, find a better way ('l' in the paper):
+  max_group_size  = max(rollapply(L$v[P,1], width=nrow(A), FUN=function(x) sum(x-x[1] <= limit), align="left", partial=TRUE))
+cat("longest run of adjacent points within threshold l =",max_group_size,"\n")
+
+# this is the big union in step 4 of algorithm 2.1, combined with step 6 to
+# convert back to original indices:
+  candidates = lapply(1:max_group_size, function(i)
+  {
+    d = diff(L$v[P,1:p,drop=FALSE],lag=i)^2 %*% L$d[1:p]^2
+    j = which(d <= 2*(1-t))  # These ordered indices meet the threshold
+    # return original un-permuted column indices
+    cbind(P[j], P[j+i])
+  })
+  Reduce(rbind,candidates) # Return a two-column array of index pairs
+}
+
+# EXAMPLE USE
+
+
+library(biclust)  # for the EisenYeast and BicatYeast data
+data(BicatYeast)  # in gene by sample orientation
+A = t(BicatYeast) # to correlate genes to genes
+
+C = cor(A)                     # actual correlation matrix for reference
+C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
+
+# Compute candidate pairs with the new algorithm
+x = tcor(A, t=0.95, p=10)
+
[…]

### 2015-07-22 threshold/draft.Rnw
notes

@@
 Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$, maximum
-subspace dimension $p<<n$.
+truncated SVD rank $p<<k$.
 \begin{enumerate}
@@
 of vectors below the correlation threshold with truncated singular value
-decompositions of only relatively low dimension $p$.  The choice of $p$ is a
+decompositions of only relatively low rank $p$.  The choice of $p$ is a
 balance between work in computing a truncated SVD and work required by the
@@
 explicitly forming a new matrix.
+
+XXX scaling/centering discussion and example
 

### 2015-07-22 threshold/tcor.R

@@ -43,7 +43,7 @@ cat("longest run of adjacent points within threshold l =",max_group_size,"\n")
 
 
 library(biclust)  # for the EisenYeast and BicatYeast data
-data(BicatYeast)  # in gene by sample orientation
+data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
 A = t(BicatYeast) # to correlate genes to genes
 
 C = cor(A)                     # actual correlation matrix for reference

### 2015-07-22 threshold/draft.Rnw
notes

@@
 let $k=\mathrm{rank}(A)$.  Assume that the mean of each column $a_j$ is zero
-and that each column has unit norm $\|a_j\|^2 = 1$. Here and below, $\|\cdot\|$
+and that each column has unit norm $\|a_j\| = 1$. Here and below, $\|\cdot\|$
 denotes the Euclidean vector norm. Then the sample correlation matrix
@@
 We have so far assumed that the mean of each column vector of $A$ is zero and
-the columns are scaled to have norm 1, but we really want a method that works
+the columns are scaled to have unit norm, but we really want a method that works
 with general matrices $A$. We've also implicitly assumed that we can

### 2015-07-22 threshold/draft.Rnw
Fixed more typos and figure label

@@
 where $v_{kj}$ denotes the $jth$ position in the vector $v_k$ (that is, the
-$[k,j]$ entry in the matrix $V$).
+$[j,k]$ entry in the matrix $V$).
 \end{lemma}
@@
 cheaply omit pairs of vectors from consideration are called \emph{pruning
-methods}, and a variety pruning methods have been described elsewhere for the
-thresholded correlation problem; see, for example\cite{zf,mnl} and the
+methods}, and a variety of pruning methods have been described elsewhere for
+the thresholded correlation problem; see, for example\cite{zf,mnl} and the
 references therein.  The approach described here prunes along an intuitive path
@@
 \begin{figure}[!ht]
-\label{fig1}
 \begin{center}
@@
 @
-\caption{Example plot of the ordered entries of $v_1$ for the BicatYeast example from Section \ref{examples}.
+\caption{
+Example plot of the ordered entries of $v_1$ for the BicatYeast example from Section \ref{examples}.
 The red lines illustrate the interval corresponding threshold $t=0.99$.
-Points farther apart than this interval correspond to column vectors that do not meet the correlation threshold.}
+Points farther apart than this interval correspond to column vectors that do not meet the correlation threshold.
+\label{fig1}
+}
 \end{center}
@@
 Then $D^1P v_1$ consists of the differences of adjacent entries in $P v_1$.
-Entries such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$
-correspond to pairs of vectors that do not meet the correlation threshold,
-(exponentiation is applied to the vector element-wise).
-However, this observation is unlikely to rule out many pairs of vectors in
-most problems. For example, the minimum distance between the points shown in
-the example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma$, and no
+Entries such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$ correspond to pairs of
+vectors that do not meet the correlation threshold, where here and below
+exponentiation applied to a vector denotes element-wise exponentiation.
+However, this observation is unlikely to rule out many pairs of vectors in most
+problems. For example, the minimum distance between the points shown in the
+example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma_1$, and no
 pruning at all occurs.
@@
 \sigma_1^2 (D^1 P v_1)^2 +
-\sigma_2^2 (D^1 P v_2)^2 > 2(1-t),
+\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
 \]
-where exponentiation is applied element-wise to the vectors.
 Note that we use the permutation $P$ defined by the order of vector $v_1$
@@
 $V_{1:p}$ denotes the first $p$ columns of the
-matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values,
-and exponentiation is applied to vectors element-wise.
+matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.
 

### 2015-07-22 threshold/draft.Rnw
Fixed column scaling in tcor.R

@@
 diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
-later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$ to be
-the vector of the first $p\le k$ singular values along the diagonal of
+later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$
+to be the vector of the first $p\le k$ singular values along the diagonal of
 $\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
@@
 V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
-matrix with eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$. The span
-of the columns of $V$ form an orthonormal basis of the range of the correlation
-matrix $A^TA$.
+matrix with nonzero eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$.
+The span of the columns of $V$ form an orthonormal basis of the range of the
+correlation matrix $A^TA$.
 
@@
 Let $t$ be a given correlation threshold $0<t<1$, and let
-$P\in\mathcal{R}^{n\times n}$ be a permutation that orders then entries of
-$v_1$ in increasing order.  It follows from Equations \ref{cordist} and
+$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders then entries
+of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
 \ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |v_{1i} - v_{1j}|^2
@@
 vectors that do not meet the correlation threshold, where here and below
-exponentiation applied to a vector denotes element-wise exponentiation.
+an exponent applied to a vector denotes element-wise exponentiation.
 However, this observation is unlikely to rule out many pairs of vectors in most
@@
 Including more terms from Equation \ref{vdist} increases our ability to prune
-pairs of vectors below the correlation threshold. For example, including
-only one more term finds that 72\% of adjacent vectors fall below the
-correlation threshold
-in the example shown in Figure \ref{fig1} and are pruned by identifying
-indices such that
+pairs of vectors below the correlation threshold. For example, including only
+one more term finds that 72\% of adjacent vectors fall below the a correlation
+threshold of $0.99$ in the example shown in Figure \ref{fig1} and are pruned by
+identifying indices such that
 \[
@@
 throughout. Including five terms prunes over 97\% of adjacent vector pairs,
-leaving only 11 candidate pairs of adjacent vectors.
+leaving only 11 candidate pairs of adjacent vectors that might possibly
+meet the correlation threshold.
 In general, using $p\le k$ terms to prune pairs of vectors below the
@@
 replacing $D^1$ with $D^j$, $j=1,2,\ldots,\ell$ in the above expressions.
-Following this procedure, we can produce a well-pruned candidate set of
-possible vector pairs that meet a given correlation threshold using $\ell$
-matrix vector products with matrices of order $n \times p$, where $p$ is chosen
-in practice such that $p<<k$. The steps of this procedure are outlined in the
-following algorithm:
+Following this process, we can produce a well-pruned candidate set of possible
+vector pairs that meet a given correlation threshold using $\ell$ matrix vector
+products with matrices of order $n \times p$, where $p$ is chosen in practice
+such that $p<<k$. These steps are formalized in Algorithm \ref{prune} below.
 
 \begin{algorithm}\label{prune}
-Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$, maximum
+Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$,
 truncated SVD rank $p<<k$.
@@
 
-The augmented, implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
+The augmented implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
 of Baglama and Reichel \cite{br} efficiently computes truncated singular value
@@
 explicitly forming a new matrix.
+
+Let $x\in\mathcal{R}^n$ represent the vector of column means of the matrix $A$,
+$W\in\mathcal{R}^{n\times n}$ be a diagonal matrix with the reciprocal of the
+column norms of $A$ along the diagonal, and let $e\in\mathcal{R}^m$ be a vector
+of all ones. Then $(A - ex^T)W$ is a centered matrix with zero column means,
+and scaled to have unit column norms, and
+
+XXX
+
+IRLBA, based on the Lanczos process, is an iterative method based on
+matrix-vector products involving the matrix $A$.
 

### 2015-07-22 threshold/tcor.R

@@ -19,7 +19,7 @@ require(zoo)   # for rollapply
 tcor = function(A, t=0.99, p=5)
 {
   mu = colMeans(A)
-  s  = sqrt(apply(A,2,crossprod))
+  s  = sqrt(apply(A,2,crossprod) - nrow(A)*mu^2) # column norms of centered matrix
   L  = irlba(A, p, center=mu, scale=s)
   P  = order(L$v[,1])  # order the entries of v1 (the permutation in the paper)
   limit = sqrt(2*(1-t))/L$d[1]

### 2015-07-22 threshold/draft.Rnw
Added notes on general matrices (scaling and centering)

@@
 \end{lemma}
-\begin{proof} (XXX perhaps omit this)
+\begin{proof} (XXX probably omit this)
 \begin{eqnarray*}
@@
 
-We have so far assumed that the mean of each column vector of $A$ is zero and
-the columns are scaled to have unit norm, but we really want a method that works
-with general matrices $A$. We've also implicitly assumed that we can
-efficiently compute a relatively low-rank truncated SVD of potentially
-large  matrices required by step 1 of Algorithm \ref{prune}.  Fortunately,
-both needs are met by one algorithm.
+We have so far assumed that the mean of each column vector of the $m\times n$
+matrix $A$ is zero and the columns are scaled to have unit norm, but we really
+want a method that works with general matrices $A$. We've also implicitly
+assumed that we can efficiently compute a relatively low-rank truncated SVD of
+potentially large  matrices required by step 1 of Algorithm \ref{prune}.
+Fortunately, both needs are met by one algorithm.
 
@@
 available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
-The R implementation includes efficient methods for re-starting the algorithm
+The R implementation includes easy to use and efficient methods for restarting
 to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
-efficient methods for computing the SVD of scaled and centered matrices without
-explicitly forming a new matrix.
+for computing the SVD of scaled and centered matrices without explicitly
+forming a new matrix.
 
-Let $x\in\mathcal{R}^n$ represent the vector of column means of the matrix $A$,
-$W\in\mathcal{R}^{n\times n}$ be a diagonal matrix with the reciprocal of the
-column norms of $A$ along the diagonal, and let $e\in\mathcal{R}^m$ be a vector
-of all ones. Then $(A - ex^T)W$ is a centered matrix with zero column means,
-and scaled to have unit column norms, and
-
-XXX
-
-IRLBA, based on the Lanczos process, is an iterative method based on
-matrix-vector products involving the matrix $A$.
-
-XXX scaling/centering discussion and example
+We relax the column mean and scale assumptions without much additional
+computational or storage overhead as follows.  Let
+$x=[x_1,x_2,\ldots,x_n]\in\mathcal{R}^n$ represent the vector of column means
+of the matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector
+of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal
+matrix with diagonal entries $1/\sqrt{w_i^2 - m x_i^2}$.  Let
+$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ex^T)W$ is a centered
+matrix with zero column means, and scaled to have unit column norms, and
+\[
+\mathrm{cor}(A) = W^T (A-ex^T)^T (A-ex^T) W.
+\]
+IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
+products. The main idea behind efficient application of IRLBA to correlation
+problems replaces matrix-vector products of the form $Az$ with $AWz - ex^TWz$ in
+the iterations, implicitly working with a scaled and centered matrix without
+forming it.  This comes at the cost of storing two additional length $n$ vectors and
+computing two additional vector inner products per matrix-vector product.
 

### 2015-07-22 threshold/draft.Rnw
Added notes on general matrices (scaling and centering)

@@
 step 1. Continue to increase $p$ in this way until the number of candidate
-pairs stops decreasing (the best this algorithm can do).
+pairs is sufficiently small or stops decreasing (the best this algorithm can do).
 \item Recover the non-permuted candidate column vector
@@
 
-\section{Fast truncated SVD for general matrices}\label{irlba}
+\section{General matrices and fast truncated SVD}\label{irlba}
 
@@
 matrix $A$ is zero and the columns are scaled to have unit norm, but we really
-want a method that works with general matrices $A$. We've also implicitly
-assumed that we can efficiently compute a relatively low-rank truncated SVD of
-potentially large  matrices required by step 1 of Algorithm \ref{prune}.
-Fortunately, both needs are met by one algorithm.
+want a method that works with general matrices $A$.  We also want a way to
+efficiently compute a relatively low-rank truncated SVD of potentially large
+matrices required by step 1 of Algorithm \ref{prune}.  Fortunately, both
+desires are met by one algorithm.
 
@@
 
-We relax the column mean and scale assumptions without much additional
+We relax the column mean and scale assumptions without introducing much additional
 computational or storage overhead as follows.  Let

### 2015-07-22 threshold/draft.Rnw
Added a few more related citations;need to work this in to the intro

@@
 
-XXX add discussion of applications (genomics?) and previous work,
+XXX add discussion of applications (genomics?) and previous work (in particular
+\cite{latj} and \cite{bhpt} are related but different),
 then on to the preliminaries:
@@
 Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix $A$
-(including possibly the case $i=j$). Denote the correlation of these two vectors
-by $\mathrm{cor}(a_i,a_j) = a_j^T a_i$.  The following simple lemma
-from Rafiei\cite{raf} establishes a relationship between correlation and
-Euclidean distance between two vectors.
+(including possibly the case $i=j$). Denote the correlation of these two
+vectors by $\mathrm{cor}(a_i,a_j) = a_j^T a_i$.  The following simple lemma
+establishes a relationship between correlation and Euclidean distance between
+two vectors.
 
@@
 
-%% XXX switch to bibtex
+%% XXX switch to bibtex / make citations uniform
 
@@
 \bibitem{br}
-Baglama, J. and Reichel, L., Augmented Implicitly Restarted Lanczos Bidiagonalization Methods, SIAM J. Sci. Comput. 2005.
+Baglama, J. and Reichel, L., Augmented Implicitly Restarted Lanczos Bidiagonalization Methods, SIAM J. Sci. Comput., 2005.
+\bibitem{bhpt}
+Bair, E., Hastie, T., Paul, D.., and Tibshirani, P., Prediction by supervised principal components, Journal of the American Statistical Association 101, no. 473, 2006.
+\bibitem{latj}
+Levin, Asriel U., Todd K. Leen, and John E. Moody. Fast pruning using principal components. (1993).
 \bibitem{mnl}

### 2015-07-22 threshold/draft.Rnw
Added TAPER reference

@@
 
-XXX add discussion of applications (genomics?) and previous work (in particular
-\cite{latj} and \cite{bhpt} are related but different),
+XXX add discussion of applications (genomics?) and previous work (especially
+\cite{xstk1},\cite{xstk2} and
+related but different work that use PCA: \cite{latj} and \cite{bhpt} are related but different),
 then on to the preliminaries:
@@
 methods}, and a variety of pruning methods have been described elsewhere for
-the thresholded correlation problem; see, for example\cite{zf,mnl} and the
-references therein.  The approach described here prunes along an intuitive path
-of decreasing variance defined by the data and computed by the SVD.
+the thresholded correlation problem; see, for example\cite{xstk1,xstk2,zf,mnl}
+and the references therein.  The approach described here prunes along an
+intuitive path of decreasing variance defined by the data and computed by the
+SVD.
 
@@
 Conference on Management of data, 2010, pp. 171-182.
-\bibitem{raf}
-Rafiei, D., On similarity-based queries for time series data,
-in Data Engineering Proceedings., IEEE, 1999.
+\bibitem{xstk1}
+Xiong, Hui, et al.
+"Exploiting a support-based upper bound of Pearson's correlation coefficient for efficiently identifying strongly correlated pairs." Proceedings of the tenth ACM SIGKDD international conference on Knowledge discovery and data mining.
+ACM, 2004.
+\bibitem{xstk2}
+Xiong, Hui, et al. "TAPER: A two-step approach for all-strong-pairs correlation query in large databases." Knowledge and Data Engineering, IEEE Transactions on 18.4 (2006): 493-508.
 \bibitem{zf}

### 2015-07-27 threshold/draft.Rnw
Notation clean up

@@
 %%
-\documentclass[12pt]{article}
+\documentclass[10pt]{article}
 
@@
 \title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
-\author{Some dudes}
+\author{
+J. Baglama\\
+Department of Mathematics,\\University of Rhode Island
+\and
+B. Lewis\\
+Paradigm4, Inc.\\
+\and
+A. Poliakov\\
+Paradigm4, Inc.
+}
 
@@
 XXX add discussion of applications (genomics?) and previous work (especially
-\cite{xstk1},\cite{xstk2} and
+\cite{xstk1},\cite{xstk2},\cite{xbm} and
 related but different work that use PCA: \cite{latj} and \cite{bhpt} are related but different),
@@
 and that each column has unit norm $\|a_j\| = 1$. Here and below, $\|\cdot\|$
-denotes the Euclidean vector norm. Then the sample correlation matrix
+denotes the Euclidean vector norm. Then the Pearson sample correlation matrix
 $\mathrm{cor}(A)=A^TA$.  Note that under these assumptions the sample
@@
 Let $A=U\Sigma V^T$ be the singular value decomposition of $A$, where
-$U\in\mathcal{R}^{m\times k}$, $U^TU = I$, $V\in\mathcal{R}^{n\times k}$,
-$V^TV=I$, and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
+$U\in\mathcal{R}^{m\times k}$, $V\in\mathcal{R}^{n\times k}$,
+$U^TU = V^TV = I$,
+and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
 diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
@@
 \end{equation}
-In particular, for a given correlation threshold $t$, $0<t<1$, pairs
-of vectors $a_i$ and $a_j$ such that $\|a_i - a_j\|^2 > 2(1-t)$ lie
-below the correlation threshold.
+In particular, for a given correlation threshold $t$, $0<t<1$, if
+$\|a_i - a_j\|^2 > 2(1-t)$ then
+$\mathrm{cor}(a_i, a_j) < t$.
 \end{lemma}
-\begin{proof} (XXX probably omit this)
-\begin{eqnarray*}
-\|a_i - a_j\|^2 &=& (a_i - a_j)^T(a_i - a_j) \\
-&=& a_i^T a_i + a_j^T a_j - 2a_i^T a_j \\
-&=& \|a_i\|^2 + \|a_j\|^2 - 2a_i^T a_j \\
-&=& 2 - 2a_i^T a_j.
-\end{eqnarray*}
-The lemma follows by solving for $a_i^Ta_j$.
-\end{proof}
+%%\begin{proof}
+%%\begin{eqnarray*}
+%%\|a_i - a_j\|^2 &=& (a_i - a_j)^T(a_i - a_j) \\
+%%&=& a_i^T a_i + a_j^T a_j - 2a_i^T a_j \\
+%%&=& \|a_i\|^2 + \|a_j\|^2 - 2a_i^T a_j \\
+%%&=& 2 - 2a_i^T a_j.
+%%\end{eqnarray*}
+%%The lemma follows by solving for $a_i^Ta_j$.
+%%\end{proof}
 
@@
 than 0.99 may only be associated with column vectors from the matrix $A$ whose
-Euclidean distance is less than $\sqrt{0.02}$.  Vectors farther apart will not
+Euclidean distance is less than $\sqrt{0.02}$.  Vectors farther apart can not
 meet the correlation threshold.
 
-The following lemma expresses the Euclidean distance between any two columns of the
-matrix $A$ as a weighted sum of projected interval distances along the SVD
+The following lemma expresses the Euclidean distance between any two columns of
+the matrix $A$ as a weighted sum of projected interval distances along the SVD
 basis vectors forming the columns of $V$.
@@
 \|a_i - a_j\|^2 =
-\sigma_1^2 |v_{1i} - v_{1j}|^2 + 
-\sigma_2^2 |v_{2i} - v_{2j}|^2 + \cdots + 
-\sigma_k^2 |v_{ki} - v_{kj}|^2,
+\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 + 
+\sigma_2^2 |e_i^Tv_{2} - e_j^Tv_{2}|^2 + \cdots + 
+\sigma_k^2 |e_i^Tv_{ki} - e_j^Tv_{k}|^2,
 \end{equation}
-where $v_{kj}$ denotes the $jth$ position in the vector $v_k$ (that is, the
-$[j,k]$ entry in the matrix $V$).
 \end{lemma}
+where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
+vector consisting of $1$ in position $j$ and zeros otherwise.
+(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$; that is, the
+$[j,k]$ entry in the matrix $V$.)
 
@@
 consideration whenever their projected distance is too large.  For example if
-$\sigma_1^2 |v_{1i} - v_{1j}|^2 \ge 2(1-t)$, then we can conclude from just two
-scalar values that the column vectors $a_i$ and $a_j$ do not meet the
+$\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 \ge 2(1-t)$, then we can conclude from
+just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
 correlation threshold $t$.
@@
 methods}, and a variety of pruning methods have been described elsewhere for
-the thresholded correlation problem; see, for example\cite{xstk1,xstk2,zf,mnl}
-and the references therein.  The approach described here prunes along an
-intuitive path of decreasing variance defined by the data and computed by the
-SVD.
+the thresholded correlation problem; see, for
+example\cite{mnl,xbm,xstk1,xstk2,zf} and the references therein.  The approach
+described here prunes along an intuitive path of decreasing variance defined by
+the data and computed by the SVD.
 
@@
 of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
-\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |v_{1i} - v_{1j}|^2
-\ge 2(1-t)$ then the column vectors $a_j$ and $a_i$ do not meet the correlation
-threshold.  Let $\ell$ be the longest run of successive points in $P v_1$
-within the interval $\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1}
-displays the entries of $v_1$ for the ``BicatYeast'' example from Section
-\ref{examples} in increasing order.  The lines in the plot illustrate the
-interval associated with the correlation threshold $t=0.99$. The quantity
-$\ell$ is obtained by rolling the interval across all the points in the plot
-and counting the maximum number of points that lie within the interval.  With
-respect to the ordering $P$, $\ell$ represents the biggest index difference
-that paris of indices corresponding to correlated vectors above the threshold
-can exhibit.
+\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |e_i^Tv_{1} -
+e_j^Tv_{1j}|^2 \ge 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
+longest run of successive points in $P v_1$ within the interval
+$\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1} displays the entries
+of $v_1$ for the small ``BicatYeast'' example (where $A\in\mathcal{R}^{70\times
+419}$) from Section \ref{examples} in increasing order.  The lines in the plot
+illustrate the interval associated with the correlation threshold $t=0.95$. The
+quantity $\ell$ is obtained by rolling the interval across all the points in
+the plot and counting the maximum number of points that lie within the
+interval.  With respect to the ordering $P$, $\ell$ represents the biggest
+index difference that paris of indices corresponding to correlated vectors
+above the threshold can exhibit.
 
@@
 plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
-t=0.99
+t=0.95
 abline(h=c(-0.05,-0.05+sqrt(2*(1-t))/s$d[1]),col=2,lwd=0.5)
@@
 \caption{
-Example plot of the ordered entries of $v_1$ for the BicatYeast example from Section \ref{examples}.
-The red lines illustrate the interval corresponding threshold $t=0.99$.
-Points farther apart than this interval correspond to column vectors that do not meet the correlation threshold.
+Example plot of the ordered entries of $v_1$ for the BicatYeast example from
+Section \ref{examples}. The red lines illustrate the interval corresponding
+threshold $t=0.95$ placed at an arbitrary location on the vertical axis. Points
+farther apart than this interval correspond to column vectors that do not meet
+the correlation threshold.
 \label{fig1}
@@
 an exponent applied to a vector denotes element-wise exponentiation.
-However, this observation is unlikely to rule out many pairs of vectors in most
-problems. For example, the minimum distance between the points shown in the
-example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma_1$, and no
-pruning at all occurs.
+
+Even a 1-d projected interval distance threshold may prune many possible
+non-adjacent vector pairs with respect to the ordering defined by $P$ as
+illustrated by the example in Figure \ref{fig1}.  However, this observation is
+unlikely to rule out many adjacent pairs of vectors in most problems. For
+example, the minimum distance between the points shown in the example in Figure
+\ref{fig1} is less that $\sqrt{2(1-t)}/\sigma_1$, and no pruning at all occurs.
 
 Including more terms from Equation \ref{vdist} increases our ability to prune
-pairs of vectors below the correlation threshold. For example, including only
-one more term finds that 72\% of adjacent vectors fall below the a correlation
-threshold of $0.99$ in the example shown in Figure \ref{fig1} and are pruned by
-identifying indices such that
+adjacent pairs of vectors with respect to the ordering defined by $P$
+below the correlation threshold. For example,
+including only one more term finds that 49\% of adjacent vectors fall below the
+a correlation threshold of $0.95$ in the example shown in Figure \ref{fig1} and
+are pruned by identifying indices such that
 \[
@@
 Note that we use the permutation $P$ defined by the order of vector $v_1$
-throughout. Including five terms prunes over 97\% of adjacent vector pairs,
-leaving only 11 candidate pairs of adjacent vectors that might possibly
+throughout. Including ten terms prunes over 96\% of adjacent vector pairs,
+leaving only 15 candidate pairs of adjacent vectors that might possibly
 meet the correlation threshold.
@@
 products with matrices of order $n \times p$, where $p$ is chosen in practice
-such that $p<<k$. These steps are formalized in Algorithm \ref{prune} below.
+such that $p<<k$. Setting $p=10$ for the example shown in Figure
+\ref{fig1} prunes all but $431$ possible vector pairs out of a total
+$419(419-1)/2 = 87,571$ with $\ell=138$.
+
+These steps are formalized in Algorithm \ref{prune} below. The algorithm
+proceeds in two main parts: steps 1--6 prune pairs of vectors down to a small
+candidate set of vector pairs that may meet the correlation threshold; step 7
+computes the correlation values and applies the threshold across the reduced
+set of vector pairs.
 
 \begin{algorithm}\label{prune}
-Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$,
+Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
+zero mean and unit norm, correlation threshold $t$,
 truncated SVD rank $p<<k$.
@@
 \item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
-where $V_{1:p} = [v_1, v_2, \ldots, v_p]$
+where\\
+$V_{1:p}=[v_1, v_2, \ldots, v_p]$
 and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
@@
 \item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
-\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
+\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
 \[
@@
 e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
-\right\},
+\right\}.
 \]
-where $e_{\tilde{j}}\in\mathcal{R}^n$ consists of $1$ in position $\tilde{j}$ and
-zeros otherwise.
 \item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
@@
 Algorithm \ref{prune} by construction guarantees that no pruned vector pair
-meets the given correlation threshold; that is, the pruning can not produce any
-``false negatives.''  The algorithm typically prunes the vast majority of pairs
+exceeds the given correlation threshold; that is, the pruning can not produce
+any false negatives.  The algorithm typically prunes the vast majority of pairs
 of vectors below the correlation threshold with truncated singular value
-decompositions of only relatively low rank $p$.  The choice of $p$ is a
-balance between work in computing a truncated SVD and work required by the
-$\ell$ matrix vector products of order $n\times p$ in step 4.
+decompositions of only relatively low rank $p$.  The choice of $p$ is a balance
+between work in computing a truncated SVD and work required by the $\ell$
+matrix vector products of order $n\times p$ in step 4.
 
@@
 computational or storage overhead as follows.  Let
-$x=[x_1,x_2,\ldots,x_n]\in\mathcal{R}^n$ represent the vector of column means
+$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
 of the matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector
 of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal
-matrix with diagonal entries $1/\sqrt{w_i^2 - m x_i^2}$.  Let
-$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ex^T)W$ is a centered
+matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
+$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
 matrix with zero column means, and scaled to have unit column norms, and
 \[
-\mathrm{cor}(A) = W^T (A-ex^T)^T (A-ex^T) W.
+\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
 \]
-IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
-products. The main idea behind efficient application of IRLBA to correlation
-problems replaces matrix-vector products of the form $Az$ with $AWz - ex^TWz$ in
+The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
+products. The main idea behind efficient application of IRLB to correlation
+problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
 the iterations, implicitly working with a scaled and centered matrix without
@@
 Conference on Management of data, 2010, pp. 171-182.
+\bibitem{xbm}
+Xiong, Hui, Mark Brodie, and Sheng Ma. "Top-cop: Mining top-k strongly correlated pairs in large databases." Data Mining, 2006. ICDM'06. Sixth International Conference on. IEEE, 2006.
 \bibitem{xstk1}

### 2015-07-27 threshold/threshold_notes.R

@@ -15,7 +15,7 @@ A = sweep(A,2,apply(A,2,mean))
 A = sweep(A,2,sqrt(apply(A,2,crossprod)),FUN=`/`)
 s = svd(A)
 
-t = 0.99 # correlation threshold
+t = 0.95 # correlation threshold
 
 # permutation ordering of the first basis vector
 i = order(s$v[,1])
@@ -46,10 +46,11 @@ abline(h=c(0,sqrt(2*(1-t))/s$d[1]),col=2)
 # the threshold and are therefore correlated or not. Each point in this plot
 # represents the difference between a *pair* of ordered points (a pair of
 # projected vectors).
-k = 5    # subspace dimension
+k = 10    # subspace dimension
 d = diff(s$v[i,1:k,drop=FALSE],lag=1)^2 %*% s$d[1:k]^2
 ylim = c(0,max(max(d), 1.01*2*(1-t)))
 plot(d,ylim=ylim)
 abline(h=2*(1 - t),col=2,lwd=1)
 j = which(d < 2*(1-t))
 points(j,d[j],pch=19,col="#ff000044")
+print(length(j))

### 2015-07-27 threshold/draft.Rnw
Better references

@@
 
+%%<<eval=TRUE, echo=FALSE>>=
+%%Sys.setenv(TEXINPUTS=getwd(),
+%%           BIBINPUTS=getwd(),
+%%           BSTINPUTS=getwd())
+%%@
+
 \begin{abstract}
@@
 Real-world problems in genomics, machine learning, finance, and other fields
-can produce correlation matrices too large to explicitly form and tracticably
+can produce correlation matrices too large to explicitly form and tractably
 compute. Often, only values corresponding to highly-correlated vectors are of
@@
 
-XXX add discussion of applications (genomics?) and previous work (especially
-\cite{xstk1},\cite{xstk2},\cite{xbm} and
-related but different work that use PCA: \cite{latj} and \cite{bhpt} are related but different),
-then on to the preliminaries:
-\[\]
+XXX
+\\
 
@@
 the thresholded correlation problem; see, for
-example\cite{mnl,xbm,xstk1,xstk2,zf} and the references therein.  The approach
-described here prunes along an intuitive path of decreasing variance defined by
-the data and computed by the SVD.
+example\cite{prune1, prune2, prune3, prune4}
+and the references therein.  The approach described here prunes along an
+intuitive path of decreasing variance defined by the data and computed by the
+SVD.
 
@@
 difference apart relative to the permutation $P$ can be similarly pruned by
-replacing $D^1$ with $D^j$, $j=1,2,\ldots,\ell$ in the above expressions.
+replacing $D^1$ with $D^j$, $j=2,3,\ldots,\ell$ in the above expressions.
 Following this process, we can produce a well-pruned candidate set of possible
@@
 The augmented implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
-of Baglama and Reichel \cite{br} efficiently computes truncated singular value
+of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value
 decompositions of large dense or sparse matrices. Reference implementations are
-available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
-The R implementation includes easy to use and efficient methods for restarting
-to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
-for computing the SVD of scaled and centered matrices without explicitly
-forming a new matrix.
+available for R\cite{irlbar}, Matlab\cite{irlbam}, and Python\cite{irlbap}.
 
@@
 $e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
-matrix with zero column means, and scaled to have unit column norms, and
+matrix with zero column means and scaled to have unit column norms, and
 \[
@@
 forming it.  This comes at the cost of storing two additional length $n$ vectors and
-computing two additional vector inner products per matrix-vector product.
+at most computing two additional vector inner products per matrix-vector product.
 
@@
 
-%% XXX switch to bibtex / make citations uniform
+\begin{thebibliography}{99}
 
-\begin{thebibliography}{99}
-\bibitem{br}
-Baglama, J. and Reichel, L., Augmented Implicitly Restarted Lanczos Bidiagonalization Methods, SIAM J. Sci. Comput., 2005.
-\bibitem{bhpt}
-Bair, E., Hastie, T., Paul, D.., and Tibshirani, P., Prediction by supervised principal components, Journal of the American Statistical Association 101, no. 473, 2006.
-\bibitem{latj}
+%% Related use of SVD in other methods
+\bibitem{spca}
+Bair, Eric, et al. Prediction by supervised principal components. Journal of the American Statistical Association 101.473 (2006).
+
+%% irlba
+\bibitem{irlba}
+Baglama, James, and Lothar Reichel. Augmented implicitly restarted Lanczos bidiagonalization methods. SIAM Journal on Scientific Computing 27.1 (2005): 19-42.
+\bibitem{irlbam}
+Baglama, James,  {\url{http://www.math.uri.edu/~jbaglama/software/irlba.m}}.
+
+%% Applications in machine learning
+\bibitem{ml1}
+Hall, Mark A. Correlation-based feature selection for machine learning. Diss. The University of Waikato, 1999.
+
+%% IRLBA implementations
+\bibitem{irlbap}
+Kane, Michael and Lewis, Bryan,  \url{https://github.com/bwlewis/irlbpy}.
+
+%% Related use of SVD in neural networks
+\bibitem{neural}
 Levin, Asriel U., Todd K. Leen, and John E. Moody. Fast pruning using principal components. (1993).
-\bibitem{mnl}
-Mueen, A., Nath, S., and Liu, J., Fast approximate correlation for
-massive time-series data, in Proceedings of the 2010 ACM SIGMOD International
-Conference on Management of data, 2010, pp. 171-182.
-\bibitem{xbm}
-Xiong, Hui, Mark Brodie, and Sheng Ma. "Top-cop: Mining top-k strongly correlated pairs in large databases." Data Mining, 2006. ICDM'06. Sixth International Conference on. IEEE, 2006.
-\bibitem{xstk1}
-Xiong, Hui, et al.
-"Exploiting a support-based upper bound of Pearson's correlation coefficient for efficiently identifying strongly correlated pairs." Proceedings of the tenth ACM SIGKDD international conference on Knowledge discovery and data mining.
-ACM, 2004.
-\bibitem{xstk2}
-Xiong, Hui, et al. "TAPER: A two-step approach for all-strong-pairs correlation query in large databases." Knowledge and Data Engineering, IEEE Transactions on 18.4 (2006): 493-508.
-\bibitem{zf}
-Zhang, J. and Feigenbaum, J, Finding Highly Correlated Pairs Efficiently with Powerful Pruning,
-in Proceedings of ACM Conference on Information and Knowledge Management (CIKM), 2006, pp. 152-161.
-\bibitem{irlba-R}
-\url{https://cran.r-project.org/web/packages/irlba/index.html}
-\bibitem{irlba-P}
-\url{https://github.com/bwlewis/irlbpy}
-\bibitem{irlba-M}
-\url{http://www.math.uri.edu/~jbaglama/software/irlba.m}
+
+%% IRLBA implementations
+\bibitem{irlbar}
+Lewis, Bryan,  \url{https://cran.r-project.org/web/packages/irlba}.
+
+%% Approximate correlation
+\bibitem{approxcor}
+Mueen, Abdullah, Suman Nath, and Jie Liu. Fast approximate correlation for massive time-series data. Proceedings of the 2010 ACM SIGMOD International Conference on Management of data. ACM, 2010.
+
+%% Applications in genomics
+\bibitem{genomics}
+Song, Lin, Peter Langfelder, and Steve Horvath. Comparison of co-expression measures: mutual information, correlation, and model based indices. BMC bioinformatics 13.1 (2012): 328.
+
+%% Other correlation pruning methods
+\bibitem{prune1}
+Xiong, Hui, et al. TAPER: A two-step approach for all-strong-pairs correlation query in large databases. Knowledge and Data Engineering, IEEE Transactions on 18.4 (2006): 493-508.
+\bibitem{prune2}
+Xiong, Hui, Mark Brodie, and Sheng Ma. Top-cop: Mining top-k strongly correlated pairs in large databases. Data Mining, 2006. ICDM'06. Sixth International Conference on. IEEE, 2006.
+
+%% Applications in machine learning
+\bibitem{ml2}
+Yu, Lei, and Huan Liu. Feature selection for high-dimensional data: A fast correlation-based filter solution. ICML. Vol. 3. 2003.
+
+%% Other correlation pruning methods
+\bibitem{prune3}
+Zhang, Jian, and Joan Feigenbaum.
+Finding highly correlated pairs efficiently with powerful pruning.
+Proceedings of the 15th ACM international conference on Information and knowledge management.
+ACM, 2006.
+\bibitem{prune4}
+Zhu, Shiwei, et al. Scaling up top-K cosine similarity search. Data \& Knowledge Engineering 70.1 (2011): 60-83.
+
 \end{thebibliography}

### 2015-07-27 threshold/draft.Rnw
references

@@
 
-XXX
-\\
+Finding highly-correlated pairs among a large set of vectors is an important
+part of many applications. For instance, subsets of highly-correlated vectors
+of gene expression values may be used in the discovery networks of genes
+relevant to particular biological processes\cite{genomics}. Identification of
+highly-correlated pairs may be used in feature selection algorithms for machine
+learning applications\cite{ml1, ml2}. The problem is also important to time
+series and image processing applications, see for example\cite{timeseries,
+svd-similarity}.
 
@@
 the thresholded correlation problem; see, for
-example\cite{prune1, prune2, prune3, prune4}
+example\cite{prune1, prune2, prune3}
 and the references therein.  The approach described here prunes along an
@@
 
-%% Related use of SVD in neural networks
-\bibitem{neural}
-Levin, Asriel U., Todd K. Leen, and John E. Moody. Fast pruning using principal components. (1993).
-
 %% IRLBA implementations
@@
 %% Approximate correlation
-\bibitem{approxcor}
+\bibitem{timeseries}
 Mueen, Abdullah, Suman Nath, and Jie Liu. Fast approximate correlation for massive time-series data. Proceedings of the 2010 ACM SIGMOD International Conference on Management of data. ACM, 2010.
@@
 Song, Lin, Peter Langfelder, and Steve Horvath. Comparison of co-expression measures: mutual information, correlation, and model based indices. BMC bioinformatics 13.1 (2012): 328.
+
+%% SVD similarity -- extremely similar to the approach in this paper!
+%% Applied to images
+\bibitem{svd-similarity}
+Wu, Daniel, et al. Efficient retrieval for browsing large image databases. Proceedings of the fifth international conference on Information and knowledge management. ACM, 1996.
 
@@
 ACM, 2006.
-\bibitem{prune4}
-Zhu, Shiwei, et al. Scaling up top-K cosine similarity search. Data \& Knowledge Engineering 70.1 (2011): 60-83.
 

### 2015-07-27 threshold/draft.Rnw
New intro

@@
 \author{
-J. Baglama\\
+James Baglama\\
 Department of Mathematics,\\University of Rhode Island
 \and
-B. Lewis\\
+Bryan Lewis\\
 Paradigm4, Inc.\\
 \and
-A. Poliakov\\
+Alex Poliakov\\
 Paradigm4, Inc.
@@
 
-%%<<eval=TRUE, echo=FALSE>>=
-%%Sys.setenv(TEXINPUTS=getwd(),
-%%           BIBINPUTS=getwd(),
-%%           BSTINPUTS=getwd())
-%%@
-
 \begin{abstract}
@@
 above a specified threshold is important to many practical applications.
-Real-world problems in genomics, machine learning, finance, and other fields
-can produce correlation matrices too large to explicitly form and tractably
-compute. Often, only values corresponding to highly-correlated vectors are of
-interest, and those values typically make up a small fraction of the overall
-correlation matrix. We present a method based on the singular value
-decomposition (SVD) and its relationship to the data covariance structure that
-can efficiently compute thresholded subsets of very large correlation matrices.
+Real-world problems in genomics, machine learning, financial and other time
+series and other applications can produce correlation matrices
+too large to explicitly form and tractably compute. Often, only values
+corresponding to highly-correlated vectors are of interest, and those values
+typically make up a small fraction of the overall correlation matrix. We
+present a method based on the singular value decomposition (SVD) and its
+relationship to the data covariance structure that can efficiently compute
+thresholded subsets of very large correlation matrices.
 
@@
 part of many applications. For instance, subsets of highly-correlated vectors
-of gene expression values may be used in the discovery networks of genes
-relevant to particular biological processes\cite{genomics}. Identification of
-highly-correlated pairs may be used in feature selection algorithms for machine
-learning applications\cite{ml1, ml2}. The problem is also important to time
-series and image processing applications, see for example\cite{timeseries,
-svd-similarity}.
+of gene expression values may be used in the discovery of networks of genes
+relevant to particular biological processes\cite{genomics}.
+Identification of highly-correlated pairs may be used in feature selection
+algorithms for machine learning applications\cite{ml1, ml2}. It's also
+important to many time series and image processing applications, see for
+example\cite{timeseries, svd-similarity} and references therein.
+
+The number of correlation coefficients grows quadratically with the number of
+vectors. Simply computing all pairs of correlation coefficients and then
+filtering out coefficients below a given threshold may not be computationally
+feasable for high-dimensional data arising in modern genomics and other
+applications. Considerable attention has been devoted to methods called
+\emph{pruning methods} that cheaply prune away all but the most
+highly-correlated pairs of vectors.  A variety of pruning methods have been
+described for the thresholded correlation problem; see, for
+example\cite{prune1, prune2} and the references therein. Related methods
+approximate a set containing the most highly-correlated pairs\cite{prune3,
+timeseries}. A method of Wu, et al.\cite{svd-similarity}, conceptually very
+similar to our approach, uses the singular value decomposition (SVD) to find
+pairs of close vectors with respect to a distance metric (instead of
+correlation).  Our method described below prunes along an intuitive path of
+decreasing variance defined by the data and computed by the SVD.
 
@@
 correlation threshold $t$.
-
-In practice, many pairs of vectors may be ruled out in this fashion by
+In practice, many pairs of vectors may be pruned in this fashion by
 inspecting distances in low-dimensional subspaces, substantially reducing the
-complexity of computing thresholded correlation matrices.  Approaches that
-cheaply omit pairs of vectors from consideration are called \emph{pruning
-methods}, and a variety of pruning methods have been described elsewhere for
-the thresholded correlation problem; see, for
-example\cite{prune1, prune2, prune3}
-and the references therein.  The approach described here prunes along an
-intuitive path of decreasing variance defined by the data and computed by the
-SVD.
+complexity of computing thresholded correlation matrices.
 

### 2015-07-27 threshold/draft.Rnw
Fix inequalities

@@
 consideration whenever their projected distance is too large.  For example if
-$\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 \ge 2(1-t)$, then we can conclude from
+$\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 > 2(1-t)$, then we can conclude from
 just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
@@
 \ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |e_i^Tv_{1} -
-e_j^Tv_{1j}|^2 \ge 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
+e_j^Tv_{1j}|^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
 longest run of successive points in $P v_1$ within the interval
@@
 unlikely to rule out many adjacent pairs of vectors in most problems.  For
-example, the minimum distance between the points shown in the example in Figure
-\ref{fig1} is less that $\sqrt{2(1-t)}/\sigma_1$, and no pruning at all occurs.
+instance, the minimum distance between the points shown in the example in
+Figure \ref{fig1} is less than $\sqrt{2(1-t)}/\sigma_1$, and no pruning occurs.
 
 Including more terms from Equation \ref{vdist} increases our ability to prune
-adjacent pairs of vectors with respect to the ordering defined by $P$
-below the correlation threshold. For example,
+adjacent pairs of vectors
+below the correlation threshold
+with respect to the ordering defined by $P$.
+For example,
 including only one more term finds that 49\% of adjacent vectors fall below the

### 2015-07-27 threshold/draft.Rnw
Commented out spca reference; not needed probably

@@
 proceeds in two main parts: steps 1--6 prune pairs of vectors down to a small
-candidate set of vector pairs that may meet the correlation threshold; step 7
+candidate set that may meet the correlation threshold; step 7
 computes the correlation values and applies the threshold across the reduced
@@
 %% Related use of SVD in other methods
-\bibitem{spca}
-Bair, Eric, et al. Prediction by supervised principal components. Journal of the American Statistical Association 101.473 (2006).
+%\bibitem{spca}
+%Bair, Eric, et al. Prediction by supervised principal components. Journal of the American Statistical Association 101.473 (2006).
 

### 2015-07-27 threshold/draft.Rnw
Minor update

@@
 of gene expression values may be used in the discovery of networks of genes
-relevant to particular biological processes\cite{genomics}.
-Identification of highly-correlated pairs may be used in feature selection
-algorithms for machine learning applications\cite{ml1, ml2}. It's also
-important to many time series and image processing applications, see for
-example\cite{timeseries, svd-similarity} and references therein.
+relevant to particular biological processes\cite{genomics}.  Identification of
+highly-correlated pairs may be used in feature selection algorithms for machine
+learning applications\cite{ml1, ml2} and are also important to time series
+and image processing applications\cite{timeseries, svd-similarity}.
 

### 2015-07-28 threshold/draft.Rnw
Typos; some more detail about other methods

@@
 highly-correlated pairs of vectors.  A variety of pruning methods have been
-described for the thresholded correlation problem; see, for
-example\cite{prune1, prune2} and the references therein. Related methods
-approximate a set containing the most highly-correlated pairs\cite{prune3,
-timeseries}. A method of Wu, et al.\cite{svd-similarity}, conceptually very
-similar to our approach, uses the singular value decomposition (SVD) to find
-pairs of close vectors with respect to a distance metric (instead of
-correlation).  Our method described below prunes along an intuitive path of
-decreasing variance defined by the data and computed by the SVD.
+described for the thresholded correlation problem.  For example the TAPER and
+related algorithms of Xiong et al. base pruning rules on the frequency of
+mutually co-ocurring observations (rows) among vector pairs; see \cite{prune1,
+prune2} and the references therein. Related methods only approximate a set
+containing the most highly-correlated pairs using hashing
+algorithms\cite{prune3}. A number of other methods arrive at approximate sets
+of the most highly-correlated paris using dimensionality-reduction techniques
+like the discrete Fourier transform\cite{timeseries}. The method of Wu, et
+al.\cite{svd-similarity}, conceptually very similar to our approach, uses the
+singular value decomposition (SVD) to find all pairs of close vectors with
+respect to a distance metric (instead of correlation).  Our method finds all of
+the most highly-correlated vector pairs with respect to a given threshold by
+pruning along an intuitive path of decreasing variance defined by the data and
+computed by the SVD.
 
@@
 $\mathrm{cor}(A)=A^TA$.  Note that under these assumptions the sample
-correlation and covariance matrices are the same.  Section \ref{irlba} below
+correlation and covariance matrices are the same.  Section \ref{irlba}
 illustrates how to relax the unit norm and zero mean assumptions in practice.
@@
 \begin{lemma}
+\label{distlemma}
 Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
@@
 
-Equation \ref{cordist} equates the problem of finding pairs of highly
+Lemma \ref{distlemma} equates the problem of finding pairs of highly
 correlated column vectors from the matrix $A$ with a problem of finding paris
@@
 
-\section{Efficient pruning using the truncated SVD}
+\section{Efficient pruning using truncated SVD}
 
@@
 of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
-\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |e_i^Tv_{1} -
-e_j^Tv_{1j}|^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
+\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |e_i^Tv_1 -
+e_j^Tv_1|^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
 longest run of successive points in $P v_1$ within the interval
@@
 replacing $D^1$ with $D^j$, $j=2,3,\ldots,\ell$ in the above expressions.
-Following this process, we can produce a well-pruned candidate set of possible
-vector pairs that meet a given correlation threshold using $\ell$ matrix vector
-products with matrices of order $n \times p$, where $p$ is chosen in practice
-such that $p<<k$. Setting $p=10$ for the example shown in Figure
-\ref{fig1} prunes all but $431$ possible vector pairs out of a total
+Following this process, we can produce a well-pruned candidate set of vector
+pairs that contains all pairs that meet a given correlation threshold using
+$\ell$ matrix vector products with matrices of order $n \times p$, where $p$ is
+chosen in practice such that $p<<k$. Setting $p=10$ for the example shown in
+Figure \ref{fig1} prunes all but $431$ possible vector pairs out of a total
 $419(419-1)/2 = 87,571$ with $\ell=138$.

### 2015-07-28 threshold/tcor.R
rollapply didn't work as expected, replaced with even less efficient code to find ell

@@ -3,7 +3,6 @@
 # As of 20-July, 2015 this needs the development version of irlba available
 # from GitHub: devtools::install_github("bwlewis/IRL")
 require(irlba)
-require(zoo)   # for rollapply
 
 #' Compute the thresholded correlations between columns of a matrix.
 #'
@@ -24,7 +23,9 @@ tcor = function(A, t=0.99, p=5)
   P  = order(L$v[,1])  # order the entries of v1 (the permutation in the paper)
   limit = sqrt(2*(1-t))/L$d[1]
 # XXX this is stupid, find a better way ('l' in the paper):
-  max_group_size  = max(rollapply(L$v[P,1], width=nrow(A), FUN=function(x) sum(x-x[1] <= limit), align="left", partial=TRUE))
+  v = L$v[P,1]
+  max_group_size = max(vapply(1:length(v), function(i) {x=v[-(1:i)];sum(x-x[1] <= limit)}, 1))
+
 cat("longest run of adjacent points within threshold l =",max_group_size,"\n")
 
 # this is the big union in step 4 of algorithm 2.1, combined with step 6 to
@@ -50,11 +51,12 @@ C = cor(A)                     # actual correlation matrix for reference
 C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
 
 # Compute candidate pairs with the new algorithm
-x = tcor(A, t=0.95, p=10)
+t = 0.95
+x = tcor(A, t=t, p=10)
 
 # Compare
 # First note, there are 115 pairs with better than 0.95 correlation:
-sum(C>0.95)
+sum(C>t)
 
 # and our algorithm with p=10 returns a list of 431 possible candidates:
 dim(x)
@@ -71,7 +73,7 @@ x[i,] = x[i,2:1]
 candidates = apply(x,1,paste,collapse=",")
 
 # Now the actual list of correlated pairs as strings (already ordered):
-actual = apply(which(C>0.95, arr.ind=TRUE),1,paste,collapse=",")
+actual = apply(which(C>t, arr.ind=TRUE),1,paste,collapse=",")
 
 # Now check that all the actual correlated pairs are accounted for in the
 # candidates:

### 2015-07-29 threshold/threshold_notes.R
Removed old notes

@@ -1,56 +1,0 @@
-library(s4vd)    # for the lung data
-library(biclust) # for the EisenYeast and BicatYeast data
-library(zoo)
-#data(lung200)
-#A = t(lung200)
-#data(EisenYeast)
-#A = t(EisenYeast)
-data(BicatYeast)
-A = t(BicatYeast)
-
-C = cor(A) # for reference
-
-# Explicitly center and scale the data matrix
-A = sweep(A,2,apply(A,2,mean))
-A = sweep(A,2,sqrt(apply(A,2,crossprod)),FUN=`/`)
-s = svd(A)
-
-t = 0.95 # correlation threshold
-
-# permutation ordering of the first basis vector
-i = order(s$v[,1])
-
-# Naive and inefficient way to find the longest run of 1-d values within
-# the correlation tolerance (IMPROVE ?).
-width = 1000
-limit = sqrt(2*(1-t))/s$d[1]   # 1-d tolerance
-group_sizes = rollapply(s$v[i,1],width=width, FUN=function(x) sum(x - x[1] < limit), align="left", partial=TRUE)
-max_group_size = max(group_sizes)
-# if max_group_size=width, then we mis-estimated. Increase width and try again until
-# max_group_size < width to be sure that we've found the maximum.
-
-cat("Maximum group size",max_group_size,"\n")
-
-
-# This plot shows each vector projected down to a point and ordered in
-# increasing order...
-plot(s$v[i,1], main="1-d ordered", ylab="v")
-# ...and the interval beyond which we can say vectors corresponding to those
-# points won't be correlated above the threshold value, here shown at zero.
-abline(h=c(0,sqrt(2*(1-t))/s$d[1]),col=2)
-
-
-
-
-# This plot shows which sequentially-adjacent points (lag=1) fall above/below
-# the threshold and are therefore correlated or not. Each point in this plot
-# represents the difference between a *pair* of ordered points (a pair of
-# projected vectors).
-k = 10    # subspace dimension
-d = diff(s$v[i,1:k,drop=FALSE],lag=1)^2 %*% s$d[1:k]^2
-ylim = c(0,max(max(d), 1.01*2*(1-t)))
-plot(d,ylim=ylim)
-abline(h=2*(1 - t),col=2,lwd=1)
-j = which(d < 2*(1-t))
-points(j,d[j],pch=19,col="#ff000044")
-print(length(j))

### 2015-07-30 threshold/example.R
changed around example R files

@@ -1,0 +1,40 @@
+# EXAMPLE USE of tcor
+source("tcor.R")
+
+
+library(biclust)  # for the EisenYeast and BicatYeast data
+data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
+A = t(BicatYeast) # to correlate genes to genes
+
+C = cor(A)                     # actual correlation matrix for reference
+C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
+
+# Compute candidate pairs with the new algorithm
+t = 0.95
+x = tcor(A, t=t, p=10)
+
+# Compare
+# First note, there are 115 pairs with better than 0.95 correlation:
+sum(C>t)
+
+# and our algorithm with p=10 returns a list of 431 possible candidates:
+dim(x)
+
+# We need to verify that x at least contains the 115 good pairs! Here is
+# one cheesy way to tell. First note that the entries of x are in somewhat
+# arbitrary order defined by the first singular vector. For example, the
+# pair (15,3) might be listed but not the equivalent redundant pair (3,15).
+# We account for that and build the following list of candidate
+# pairs in order so that the first coordinate is less than the second:
+i = which(x[,2] < x[,1])
+x[i,] = x[i,2:1]
+# Express the coordinates as strings
+ans = apply(x,1,paste,collapse=",")
+
+# Now the actual list of correlated pairs as strings (already ordered):
+actual = apply(which(C>t, arr.ind=TRUE),1,paste,collapse=",")
+
+# Now check that all the actual correlated pairs are accounted for in the
+# candidates:
+length(intersect(actual, ans$candidates))
+# It's 115, so all the actual correlated pairs were found (plus some extra)!

### 2015-07-30 threshold/tcor.R

@@ -1,4 +1,5 @@
 # EXPERIMENTAL PROTOTYPE FUNCTION
+# For an example use of this, see the file example.R.
 #
 # As of 20-July, 2015 this needs the development version of irlba available
 # from GitHub: devtools::install_github("bwlewis/IRL")
@@ -13,8 +14,12 @@ require(irlba)
 #' @param p projected subspace dimension
 #'
 #' @return
-#' A two-column matrix. Each row of the matrix are the
-#' indices of candidate vectors that meet the correlation threshold.
+#' A list with entries:
+#' candidates: A two-column matrix. Each row of the matrix are the
+#'             indices of candidate vectors that meet the correlation threshold.
+#' dnum: a vector of length equal to the longest run of adjacent points
+#'       in v_1 meeting the threshold that contains the number of candidate
+#'       entries making the threshold cut at each index difference.
 tcor = function(A, t=0.99, p=5)
 {
   mu = colMeans(A)
@@ -22,64 +27,23 @@ tcor = function(A, t=0.99, p=5)
   L  = irlba(A, p, center=mu, scale=s)
   P  = order(L$v[,1])  # order the entries of v1 (the permutation in the paper)
   limit = sqrt(2*(1-t))/L$d[1]
-# XXX this is stupid, find a better way ('l' in the paper):
+# XXX this is stupid, find a better way ('l' or LaTeX '\ell' in the paper):
   v = L$v[P,1]
-  max_group_size = max(vapply(1:length(v), function(i) {x=v[-(1:i)];sum(x-x[1] <= limit)}, 1))
+  ell = max(vapply(1:length(v), function(i) {x=v[-(1:i)];sum(x-x[1] <= limit)}, 1))
 
-cat("longest run of adjacent points within threshold l =",max_group_size,"\n")
+cat("longest run of adjacent points within threshold l =",ell,"\n")
 
+  dnum = rep(0,ell)
 # this is the big union in step 4 of algorithm 2.1, combined with step 6 to
 # convert back to original indices:
-  candidates = lapply(1:max_group_size, function(i)
+  candidates = lapply(1:ell, function(i)
   {
     d = diff(L$v[P,1:p,drop=FALSE],lag=i)^2 %*% L$d[1:p]^2
     j = which(d <= 2*(1-t))  # These ordered indices meet the threshold
+    dnum[i] = length(j)      # Record the number of them
     # return original un-permuted column indices
     cbind(P[j], P[j+i])
   })
-  Reduce(rbind,candidates) # Return a two-column array of index pairs
[…]

### 2015-07-30 threshold/tcor.R
typo in tcor.R dnum

@@ -40,7 +40,7 @@ cat("longest run of adjacent points within threshold l =",ell,"\n")
   {
     d = diff(L$v[P,1:p,drop=FALSE],lag=i)^2 %*% L$d[1:p]^2
     j = which(d <= 2*(1-t))  # These ordered indices meet the threshold
-    dnum[i] = length(j)      # Record the number of them
+    dnum[i] <<- length(j)    # Record the number of them
     # return original un-permuted column indices
     cbind(P[j], P[j+i])
   })

### 2015-07-30 threshold/experiments.R
Added interesting experiment about projected dimension size

@@ -1,0 +1,24 @@
+source("tcor.R")  # thresholded correlation prototype function
+
+library(biclust)  # for the EisenYeast and BicatYeast example data
+# A tiny example dataset
+data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
+A = t(BicatYeast) # to correlate genes to genes
+
+t = 0.95
+
+C = cor(A)                     # actual correlation matrix for reference
+C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
+
+# Compare solutions in several dimensions
+x2  = tcor(A,t,p=2)
+x4  = tcor(A,t,p=4)
+x8  = tcor(A,t,p=8)
+x32 = tcor(A,t,p=32)
+
+
+par(mfrow=c(2,2))
+plot(x2$dnum,ylab="",xlab="index difference",main=paste("2-d projection, num. candidates =",nrow(x2$candidates)))
+plot(x4$dnum,ylab="",xlab="index difference",main=paste("4-d projection, num. candidates =",nrow(x4$candidates)))
+plot(x8$dnum,ylab="",xlab="index difference",main=paste("8-d projection, num. candidates =",nrow(x8$candidates)))
+plot(x32$dnum,ylab="",xlab="index difference",main=paste("32-d projection, num. candidates =",nrow(x32$candidates)))

### 2015-07-31 threshold/draft.Rnw
Added the distribution of cov(a_i, a_j) with normal assumptions.

@@
 
+\section{The distribution of $||a_i - a_j||^2$}
+
+\begin{lemma}
+\label{asymdistm1lemma}
+Let $m=1$, $X_i$ and $X_j$ be distributed as $\mathcal{N}(0, \Delta_{ij})$,
+with $\Delta_{ii} = 1$.  Then
+\begin{equation}
+\label{eqasymdist1}
+|| X_i - X_j ||^2 \sim R + \frac{\Delta_{ij} + 1}{2} S + 
+  \frac{\Delta_{ij} - 1}{2} T
+\end{equation}
+where $R$ is distributed as $\chi_2^2$ and $S$ and $T$ are $\chi_1^2$.
+\end{lemma}
+\begin{proof}
+\begin{equation*}
+||X_i - X_j||^2 = X_i^2 - 2 X_i X_j + X_j^2
+\end{equation*}
+$X_i^2$ and $X_j^2$ are clearly $\chi_1^2$. To derive the distribution
+of the cross-product term realize
+\begin{equation*}
+X_i X_j = \frac{1}{4} \left(X_i + X_j\right)^2 - 
+  \frac{1}{4} \left(X_i - X_j\right)^2.
+\end{equation*}
+$X_i + X_j$ and $X_i - X_j$ are both distributed as normal and they are 
+independent
+since $\mathbf{E}X_1^2 = \mathbf{E}X_2^2$. Therefore, their squares
+are $\chi_1^2$ and the leading coefficients in Equation \ref{eqasymdist1}
+are the respective variances.
+\end{proof}
+
+\begin{lemma}
+\label{asymdistlemma}
+Let $a_i$ and $a_j$ be any two columns of the matrix 
+$B \in \mathcal{R}^{m \times n}$, sampled from 
+$\mathcal{N}(0, \Delta/m)$ where $\Delta_{ii} = 1$ for $1 \leq i \leq n$.
+Then
+\begin{equation*}
+||a_i - a_j||^2 \xrightarrow[m]{D} \mathcal{N} \left(2 + \Delta_{ij},  
+  \frac{5 + \Delta_{ij}^2}{m} \right) 
+\end{equation*}
+\end{lemma}
+This result follows by applying the central limit theorem to
+the result in Lemma \ref{asymdistm1lemma} where $m$ is appropriately large.
+
 \begin{thebibliography}{99}

### 2015-08-02 threshold/draft.Rnw
Minor correction.
Fixed the cross-product coefficient.
Fixed the variance.
typo

--- 2015-07-28 threshold/draft.Rnw
@@
 unlikely to rule out many adjacent pairs of vectors in most problems.  For
-instance, the minimum distance between the points shown in the example in
+instance, the maximum distance between adjacent points shown in the example in
 Figure \ref{fig1} is less than $\sqrt{2(1-t)}/\sigma_1$, and no pruning occurs.

### 2015-08-02 threshold/experiments.R

@@ -14,7 +14,12 @@ C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
 x2  = tcor(A,t,p=2)
 x4  = tcor(A,t,p=4)
 x8  = tcor(A,t,p=8)
+x20 = tcor(A,t,p=20)
+x24 = tcor(A,t,p=24)
 x32 = tcor(A,t,p=32)
+
+d = function(i) diag(s$v[,1:i] %*% diag(s$d[1:i]^2) %*% t(s$v[,1:i]))
+
 
 
 par(mfrow=c(2,2))

### 2015-08-02 threshold/draft.Rnw

--- 2015-07-31 threshold/draft.Rnw
@@
 \label{eqasymdist1}
-|| X_i - X_j ||^2 \sim R + \frac{\Delta_{ij} + 1}{2} S + 
-  \frac{\Delta_{ij} - 1}{2} T
+|| X_i - X_j ||^2 \sim R + \left(\Delta_{ij} + 1 \right) S + 
+  \left(\Delta_{ij} - 1 \right) T
 \end{equation}
@@
 \begin{equation*}
-||a_i - a_j||^2 \xrightarrow[m]{D} \mathcal{N} \left(2 + \Delta_{ij},  
-  \frac{5 + \Delta_{ij}^2}{m} \right) 
+||a_i - a_j||^2 \xrightarrow[m]{D} \mathcal{N} \left(2 - 2 \Delta_{ij},  
+  \frac{6 + 4\Delta_{ij}^2}{m} \right) 
 \end{equation*}

### 2015-08-02 threshold/draft.Rnw
restart in Alg. 2.1

@@
 Department of Mathematics,\\University of Rhode Island
+\and
+Michael Kane\\
+Yale University
 \and
@@
 \begin{enumerate}
-\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
+\item Compute the truncated SVD
+$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
 where\\
@@
 \]
-\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
-step 1. Continue to increase $p$ in this way until the number of candidate
+\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and
+restart IRLBA to compute 
+$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
+Continue to increase $p$ in this way until the number of candidate
 pairs is sufficiently small or stops decreasing (the best this algorithm can do).
@@
 
-We relax the column mean and scale assumptions without introducing much additional
-computational or storage overhead as follows.  Let
+We can relax the column mean and scale assumptions without introducing much
+additional computational or storage overhead as follows.  Let
 $z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
-of the matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector
-of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal
-matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
+of the $m\times n$ matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
+be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
+a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
 $e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
@@
 %%@
-
-\section{Conclusion}\label{conclusion}
 
@@
 the result in Lemma \ref{asymdistm1lemma} where $m$ is appropriately large.
+
+\section{Conclusion}\label{conclusion}
 

### 2015-08-02 threshold/experiments.R

@@ -26,4 +26,4 @@ par(mfrow=c(2,2))
 plot(x2$dnum,ylab="",xlab="index difference",main=paste("2-d projection, num. candidates =",nrow(x2$candidates)))
 plot(x4$dnum,ylab="",xlab="index difference",main=paste("4-d projection, num. candidates =",nrow(x4$candidates)))
 plot(x8$dnum,ylab="",xlab="index difference",main=paste("8-d projection, num. candidates =",nrow(x8$candidates)))
-plot(x32$dnum,ylab="",xlab="index difference",main=paste("32-d projection, num. candidates =",nrow(x32$candidates)))
+plot(x24$dnum,ylab="",xlab="index difference",main=paste("24-d projection, num. candidates =",nrow(x24$candidates)))

### 2015-08-02 threshold/draft.Rnw
restart in Alg. 2.1

@@
 algorithms\cite{prune3}. A number of other methods arrive at approximate sets
-of the most highly-correlated paris using dimensionality-reduction techniques
+of the most highly-correlated pairs using dimensionality-reduction techniques
 like the discrete Fourier transform\cite{timeseries}. The method of Wu, et
@@
 Lemma \ref{distlemma} equates the problem of finding pairs of highly
-correlated column vectors from the matrix $A$ with a problem of finding paris
+correlated column vectors from the matrix $A$ with a problem of finding pairs
 of vectors sufficiently close together. For example, correlation values larger
@@
 interval.  With respect to the ordering $P$, $\ell$ represents the biggest
-index difference that paris of indices corresponding to correlated vectors
+index difference that pairs of indices corresponding to correlated vectors
 above the threshold can exhibit.

### 2015-08-03 threshold/draft.Rnw
Removed absolute value in squared terms where appropriate

@@
 \|a_i - a_j\|^2 =
-\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 + 
-\sigma_2^2 |e_i^Tv_{2} - e_j^Tv_{2}|^2 + \cdots + 
-\sigma_k^2 |e_i^Tv_{ki} - e_j^Tv_{k}|^2,
+\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 + 
+\sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
+\sigma_k^2 (e_i^Tv_{ki} - e_j^Tv_{k})^2,
 \end{equation}
@@
 consideration whenever their projected distance is too large.  For example if
-$\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 > 2(1-t)$, then we can conclude from
+$\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 > 2(1-t)$, then we can conclude from
 just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
@@
 of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
-\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |e_i^Tv_1 -
-e_j^Tv_1|^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
+\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 (e_i^Tv_1 -
+e_j^Tv_1)^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
 longest run of successive points in $P v_1$ within the interval

### 2015-08-03 threshold/experiments.R

@@ -14,16 +14,29 @@ C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
 x2  = tcor(A,t,p=2)
 x4  = tcor(A,t,p=4)
 x8  = tcor(A,t,p=8)
-x20 = tcor(A,t,p=20)
 x24 = tcor(A,t,p=24)
-x32 = tcor(A,t,p=32)
 
 d = function(i) diag(s$v[,1:i] %*% diag(s$d[1:i]^2) %*% t(s$v[,1:i]))
 
 
 
-par(mfrow=c(2,2))
+p = par(mfrow=c(2,2))
 plot(x2$dnum,ylab="",xlab="index difference",main=paste("2-d projection, num. candidates =",nrow(x2$candidates)))
 plot(x4$dnum,ylab="",xlab="index difference",main=paste("4-d projection, num. candidates =",nrow(x4$candidates)))
 plot(x8$dnum,ylab="",xlab="index difference",main=paste("8-d projection, num. candidates =",nrow(x8$candidates)))
 plot(x24$dnum,ylab="",xlab="index difference",main=paste("24-d projection, num. candidates =",nrow(x24$candidates)))
+par(p)
+
+
+# Shows how conservative the sqrt(2*(1-t))/d[1] bound is
+#A = matrix(rnorm(100*500),nrow=100)
+#C = cor(A)
+#C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
+#t = 0.80*max(C)
+#i = which(C>=t,arr.ind=TRUE)
+#A = scale(A,center=TRUE,scale=sqrt(apply(A,2,crossprod)))
+#s = svd(A)
+#lim = sqrt(2*(1-t))/s$d[1]
+#x = abs(s$v[i[,1],1] - s$v[i[,2],1])
+#plot(x, ylim=c(0,lim*1.2))
+#abline(h=lim,col=2)

### 2015-09-26 threshold/draft.Rnw
typos

@@
 \sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
-\sigma_k^2 (e_i^Tv_{ki} - e_j^Tv_{k})^2,
+\sigma_k^2 (e_i^Tv_{k} - e_j^Tv_{k})^2,
 \end{equation}
@@
 e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
-\right\}.
+\right\},
 \]
+where $\tilde{j}=1,2,\ldots,n-\tilde{i}$.
 \item If there are judged to be  ``too many'' candidate pairs, increase $p$ and
@@
 %%@
+
+\section{NOTES}
+
+A full/naive thresholded correlation requires $O(n^2 m)$ flops. Let $p$ be the
+selected IRLBA dimension. Then the proposed method requires $O(\ell  n  p)$
+flops, where $\ell$ is the longest run of ordered entries in $v_1$ that meet
+the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$ (not counting
+from flops due to IRLBA and the post-processing step of evaluating the
+correlation of all the candidate pairs).  One problem with the method as
+written is that $\ell$ can a large fraction of n.
+
+Unnecessarily large values of $\ell$ are due to the loose bound
+$\sqrt{2(1-t)}/\sigma_1$ used in step 3 of Algorithm\ref{prune} which is only
+sharp when $k=1$. General problems may not need to run through differencing
+steps up to $\ell$ in step 4 of the algorithm. We can further improve the
+efficiency of the algorithm by dynamically reducing $\ell$ in step 4 as
+follows.
+
+Equation \ref{vdist} expresses the Euclidean distance between any two columns
+of the matrix $A$ in terms of singular components. Write the smallest squared
+gap between entries of a particular singular vector $v_h$ as:
+\[
+\delta_h = \min_{i,j} (e_i^T v_h - e_j^T v_h)^2.
+\]
+Then note that
+\begin{equation}\label{mindist}
+\|a_i - a_j\|^2 \ge \sigma_1^2 \delta_1 + \sigma_2^2 \delta_2 + \cdots + \sigma_k^2\delta_k,
+\end{equation}
+for any indices $i,j$.
+
+Fix $1\le \tilde{i}\le \ell$ and let $\nu \in R^p$ be the vector of minimum values in
+each column of the $n\times p$ matrix $D^{\tilde{i}}PV_{1:p}$ from step 4 of
+the algorithm.
+It follows from Equation \ref{mindist} that
+\[
+e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \ge \nu^Ts_{1:p}
+\]
+for all $\tilde{j}=1,2,\ldots,n$.
+
+XXX need monotonicity of $\nu^Ts_{1:p}$ XXX
 

### 2015-09-26 threshold/example.R

@@ -26,6 +26,9 @@ dim(x)
 # pair (15,3) might be listed but not the equivalent redundant pair (3,15).
 # We account for that and build the following list of candidate
 # pairs in order so that the first coordinate is less than the second:
+check = function(x,C)
+{
+x = x[[1]]
 i = which(x[,2] < x[,1])
 x[i,] = x[i,2:1]
 # Express the coordinates as strings
@@ -36,5 +39,7 @@ actual = apply(which(C>t, arr.ind=TRUE),1,paste,collapse=",")
 
 # Now check that all the actual correlated pairs are accounted for in the
 # candidates:
-length(intersect(actual, ans$candidates))
+print(length(actual))
+print(length(intersect(actual, ans)))
 # It's 115, so all the actual correlated pairs were found (plus some extra)!
+}

### 2015-09-26 threshold/tcor.R

@@ -20,13 +20,18 @@ require(irlba)
 #' dnum: a vector of length equal to the longest run of adjacent points
 #'       in v_1 meeting the threshold that contains the number of candidate
 #'       entries making the threshold cut at each index difference.
-tcor = function(A, t=0.99, p=5)
+tcor = function(A, t=0.99, p=5, reduce=1)
 {
   mu = colMeans(A)
   s  = sqrt(apply(A,2,crossprod) - nrow(A)*mu^2) # column norms of centered matrix
   L  = irlba(A, p, center=mu, scale=s)
   P  = order(L$v[,1])  # order the entries of v1 (the permutation in the paper)
   limit = sqrt(2*(1-t))/L$d[1]
+# XXX EXPERIMENTAL reduce limit by fraction of variance in 1-d
+#if(reduce) limit = limit*L$d[1]^2/ncol(A) 
+limit = limit*reduce
+# sum of squared singular values of centered/scaled A = ncol(A)
+
 # XXX this is stupid, find a better way ('l' or LaTeX '\ell' in the paper):
   v = L$v[P,1]
   ell = max(vapply(1:length(v), function(i) {x=v[-(1:i)];sum(x-x[1] <= limit)}, 1))

### 2015-11-20 threshold/draft.Rnw
Paper update
Minor re-wording of text, fix typos, etc. I am now thinking of avoiding discussion of probabalistic implications of reducint ell, in favor of a simple, crisp and concise result.

@@
 Real-world problems in genomics, machine learning, financial and other time
-series and other applications can produce correlation matrices
-too large to explicitly form and tractably compute. Often, only values
-corresponding to highly-correlated vectors are of interest, and those values
-typically make up a small fraction of the overall correlation matrix. We
-present a method based on the singular value decomposition (SVD) and its
-relationship to the data covariance structure that can efficiently compute
-thresholded subsets of very large correlation matrices.
+series and other applications can produce correlation matrices too large to
+explicitly form and tractably compute. Often, only values corresponding to
+highly-correlated vectors are of interest, and those values typically make up a
+small fraction of the overall correlation matrix. We present a method based on
+the singular value decomposition (SVD) and its relationship to the data
+covariance structure that can efficiently compute thresholded subsets of very
+large correlation matrices.
 
@@
 highly-correlated pairs may be used in feature selection algorithms for machine
-learning applications\cite{ml1, ml2} and are also important to time series
-and image processing applications\cite{timeseries, svd-similarity}.
+learning applications\cite{ml1, ml2} and are also important to time series and
+image processing applications\cite{timeseries, svd-similarity}.
 
@@
 filtering out coefficients below a given threshold may not be computationally
-feasable for high-dimensional data arising in modern genomics and other
-applications. Considerable attention has been devoted to methods called
-\emph{pruning methods} that cheaply prune away all but the most
-highly-correlated pairs of vectors.  A variety of pruning methods have been
-described for the thresholded correlation problem.  For example the TAPER and
-related algorithms of Xiong et al. base pruning rules on the frequency of
-mutually co-ocurring observations (rows) among vector pairs; see \cite{prune1,
-prune2} and the references therein. Related methods only approximate a set
-containing the most highly-correlated pairs using hashing
-algorithms\cite{prune3}. A number of other methods arrive at approximate sets
-of the most highly-correlated pairs using dimensionality-reduction techniques
-like the discrete Fourier transform\cite{timeseries}. The method of Wu, et
-al.\cite{svd-similarity}, conceptually very similar to our approach, uses the
-singular value decomposition (SVD) to find all pairs of close vectors with
-respect to a distance metric (instead of correlation).  Our method finds all of
-the most highly-correlated vector pairs with respect to a given threshold by
-pruning along an intuitive path of decreasing variance defined by the data and
-computed by the SVD.
+feasable for high-dimensional data in modern genomics and other applications.
+Considerable attention has been devoted to methods called \emph{pruning
+methods} that cheaply prune away all but the most highly-correlated pairs of
+vectors.  For example the TAPER and related algorithms of Xiong et al. base
+pruning rules on the frequency of mutually co-ocurring observations (rows)
+among vector pairs; see \cite{prune1, prune2} and the references therein.
+Related methods only approximate a set containing the most highly-correlated
+pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
+at approximate sets of the most highly-correlated pairs using
+dimensionality-reduction techniques like the discrete Fourier
+transform\cite{timeseries}. The method of Wu, et al.\cite{svd-similarity},
+conceptually very similar to our approach, uses the singular value
+decomposition (SVD) to find all pairs of close vectors with respect to a
+distance metric (instead of correlation).  Our method finds all of the most
+highly-correlated vector pairs with respect to a given threshold by pruning
+along an intuitive path of decreasing variance defined by the data and computed
+by the SVD.
 
@@
 
-Let $A=U\Sigma V^T$ be the singular value decomposition of $A$, where
+Let $A=U\Sigma V^T$ be a singular value decomposition of $A$, where
 $U\in\mathcal{R}^{m\times k}$, $V\in\mathcal{R}^{n\times k}$,
@@
 
-Lemma \ref{distlemma} equates the problem of finding pairs of highly
-correlated column vectors from the matrix $A$ with a problem of finding pairs
-of vectors sufficiently close together. For example, correlation values larger
-than 0.99 may only be associated with column vectors from the matrix $A$ whose
-Euclidean distance is less than $\sqrt{0.02}$.  Vectors farther apart can not
-meet the correlation threshold.
+Lemma \ref{distlemma} equates the problem of finding pairs of highly correlated
+column vectors from the matrix $A$ with a problem of finding pairs of vectors
+sufficiently close together. For example, correlation values larger than 0.99
+may only be associated with column vectors from the matrix $A$ whose Euclidean
+distance is less than $\sqrt{0.02}$.  Vectors farther apart can not meet the
+correlation threshold.
 
@@
 vector consisting of $1$ in position $j$ and zeros otherwise.
-(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$; that is, the
+(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$--that is, the
 $[j,k]$ entry in the matrix $V$.)
@@
 $\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 > 2(1-t)$, then we can conclude from
-just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
+just a few scalar values that the column vectors $a_i$ and $a_j$ do not meet the
 correlation threshold $t$.
@@
 
-Let $t$ be a given correlation threshold $0<t<1$, and let
-$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders then entries
-of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
-\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 (e_i^Tv_1 -
-e_j^Tv_1)^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
-longest run of successive points in $P v_1$ within the interval
-$\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1} displays the entries
-of $v_1$ for the small ``BicatYeast'' example (where $A\in\mathcal{R}^{70\times
-419}$) from Section \ref{examples} in increasing order.  The lines in the plot
-illustrate the interval associated with the correlation threshold $t=0.95$. The
-quantity $\ell$ is obtained by rolling the interval across all the points in
-the plot and counting the maximum number of points that lie within the
-interval.  With respect to the ordering $P$, $\ell$ represents the biggest
-index difference that pairs of indices corresponding to correlated vectors
-above the threshold can exhibit.
+Let $t$ be a given correlation threshold, $0<t<1$, and let
+$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders the entries
+of $v_1$ in increasing order.  For example, Figure \ref{fig1} displays the
+ordered entries of $Pv_1$ for the small ``BicatYeast'' example (where
+$A\in\mathcal{R}^{70\times 419}$) from Section \ref{examples}.  The lines in
+the plot illustrate the interval associated with the correlation threshold
+$t=0.95$ placed at an arbitrary vertical axis location.
 
@@
 
+Let $\ell$ be the longest run of successive points in $P v_1$ within the
+interval $\sqrt{2(1-t)}/\sigma_1$.  The quantity $\ell$ may, for example, be
+obtained by rolling the interval over all the points in Figure \ref{fig1} and
+counting the maximum number of points that lie within the interval.  With
+respect to the ordering $P$, $\ell$ represents the biggest index difference
+that pairs of indices corresponding to correlated vectors above the threshold
+can exhibit.
+
 Pairs of vectors below the correlation threshold that lie more than one index
@@
 where $\tilde{j}=1,2,\ldots,n-\tilde{i}$.
-\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and
-restart IRLBA to compute 
-$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
+\item If there are judged to be  ``too many'' candidate pairs, increase $p$,
+compute $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
 Continue to increase $p$ in this way until the number of candidate
@@
 efficiently compute a relatively low-rank truncated SVD of potentially large
-matrices required by step 1 of Algorithm \ref{prune}.  Fortunately, both
-desires are met by one algorithm.
+matrices required by step 1 of Algorithm \ref{prune}.  And the truncated SVD
+method should be able to cheaply restart to compute additional singular vectors
+as required in step 5 of Algorithm \ref{prune}. Fortunately all desires are met
+by one algorithm.
 
@@
 We can relax the column mean and scale assumptions without introducing much
-additional computational or storage overhead as follows.  Let
+additional computational or storage overhead as follows. Assume $A$ is an
+$m\times n$ real-valued matrix without constant-valued columns. Let
 $z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
@@
 at most computing two additional vector inner products per matrix-vector product.
+The R implementation\cite{irlbar} of IRLBA includes arguments for centering and
+scaling the input matrix.
 
 \section{Numerical experiments}\label{examples}
+
+A full/naive thresholded correlation computation requires $O(n^2 m)$ flops. Let
+$p$ be the selected IRLBA dimension. Then the proposed method requires
+$O(\ell n p)$ flops, where $\ell$ is the longest run of ordered entries in $v_1$
+that
+meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
+plus the flops required by IRLBA and the post-processing step of evaluating the
+correlation of all the candidate pairs.
 
@@
 
-\section{NOTES}
+\section{NOTES not for publication}
 

### 2015-11-20 threshold/draft.Rnw
Removed some words

@@
 highly-correlated pairs may be used in feature selection algorithms for machine
-learning applications\cite{ml1, ml2} and are also important to time series and
+learning applications\cite{ml1, ml2} and are important to time series and
 image processing applications\cite{timeseries, svd-similarity}.
@@
 feasable for high-dimensional data in modern genomics and other applications.
-Considerable attention has been devoted to methods called \emph{pruning
+Considerable attention has been devoted to \emph{pruning
 methods} that cheaply prune away all but the most highly-correlated pairs of
-vectors.  For example the TAPER and related algorithms of Xiong et al. base
-pruning rules on the frequency of mutually co-ocurring observations (rows)
+vectors.  For example the TAPER and related algorithms of Xiong et al. develop
+pruning rules based on the frequency of mutually co-ocurring observations (rows)
 among vector pairs; see \cite{prune1, prune2} and the references therein.
-Related methods only approximate a set containing the most highly-correlated
+Related methods approximate a set containing the most highly-correlated
 pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
@@
 $\mathrm{cor}(A)=A^TA$.  Note that under these assumptions the sample
-correlation and covariance matrices are the same.  Section \ref{irlba}
-illustrates how to relax the unit norm and zero mean assumptions in practice.
+correlation and covariance matrices are the same up to a constant multiple.
+Section \ref{irlba} illustrates how to relax the unit norm and zero mean
+assumptions in practice.
 
@@
 may only be associated with column vectors from the matrix $A$ whose Euclidean
-distance is less than $\sqrt{0.02}$.  Vectors farther apart can not meet the
-correlation threshold.
+distance is less than $\sqrt{0.02}$.
 
@@
 \begin{lemma}
-Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
-$A$. Then
+Let $a_i$ and $a_j$
+be any two columns of the matrix
+$A$, $1\le i,j\le n$. Then
 \begin{equation}\label{vdist}
@@
 instance, the maximum distance between adjacent points shown in the example in
-Figure \ref{fig1} is less than $\sqrt{2(1-t)}/\sigma_1$, and no pruning occurs.
+Figure \ref{fig1} is less than $\sqrt{2(1-t)}/\sigma_1$, and no pruning of
+adjacent pairs of vectors with respect to the ordering $P$ occurs.
 
@@
 
-Algorithm \ref{prune} by construction guarantees that no pruned vector pair
-exceeds the given correlation threshold; that is, the pruning can not produce
-any false negatives.  The algorithm typically prunes the vast majority of pairs
-of vectors below the correlation threshold with truncated singular value
-decompositions of only relatively low rank $p$.  The choice of $p$ is a balance
-between work in computing a truncated SVD and work required by the $\ell$
-matrix vector products of order $n\times p$ in step 4.
+Algorithm \ref{prune} guarantees that no pruned vector pair exceeds the given
+correlation threshold--the pruning does not produce false negatives.
+The algorithm typically prunes the vast majority of pairs of vectors below the
+correlation threshold with truncated singular value decompositions of only
+relatively low rank $p$.  The choice of $p$ represents a balance between work in
+computing a truncated SVD and work required by the $\ell$ matrix vector
+products of order $n\times p$ in step 4.
 
@@
 
-The augmented implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
+The augmented implicitly restarted Lanczos bidiagonalization algorithm\break(IRLBA)
 of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value

### 2015-11-20 threshold/example.R
Updated example tcor.R implementation

@@ -1,7 +1,6 @@
 # EXAMPLE USE of tcor
 source("tcor.R")
 
-
 library(biclust)  # for the EisenYeast and BicatYeast data
 data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
 A = t(BicatYeast) # to correlate genes to genes
@@ -14,32 +13,5 @@ t = 0.95
 x = tcor(A, t=t, p=10)
 
 # Compare
-# First note, there are 115 pairs with better than 0.95 correlation:
-sum(C>t)
-
-# and our algorithm with p=10 returns a list of 431 possible candidates:
+sum(C > t)
 dim(x)
-
-# We need to verify that x at least contains the 115 good pairs! Here is
-# one cheesy way to tell. First note that the entries of x are in somewhat
-# arbitrary order defined by the first singular vector. For example, the
-# pair (15,3) might be listed but not the equivalent redundant pair (3,15).
-# We account for that and build the following list of candidate
-# pairs in order so that the first coordinate is less than the second:
-check = function(x,C)
-{
-x = x[[1]]
-i = which(x[,2] < x[,1])
-x[i,] = x[i,2:1]
-# Express the coordinates as strings
-ans = apply(x,1,paste,collapse=",")
-
-# Now the actual list of correlated pairs as strings (already ordered):
-actual = apply(which(C>t, arr.ind=TRUE),1,paste,collapse=",")
-
-# Now check that all the actual correlated pairs are accounted for in the
-# candidates:
-print(length(actual))
-print(length(intersect(actual, ans)))
-# It's 115, so all the actual correlated pairs were found (plus some extra)!
-}

### 2015-11-20 threshold/tcor.R

@@ -1,10 +1,5 @@
-# EXPERIMENTAL PROTOTYPE FUNCTION
-# For an example use of this, see the file example.R.
-#
-# As of 20-July, 2015 this needs the development version of irlba available
-# from GitHub: devtools::install_github("bwlewis/IRL")
 require(irlba)
-
+require(parallel)
 #' Compute the thresholded correlations between columns of a matrix.
 #'
 #' Increase p to cut down the total number of candate pairs.
@@ -13,42 +8,31 @@ require(irlba)
 #' @param t a threshold value for correlation
 #' @param p projected subspace dimension
 #'
-#' @return
-#' A list with entries:
-#' candidates: A two-column matrix. Each row of the matrix are the
-#'             indices of candidate vectors that meet the correlation threshold.
-#' dnum: a vector of length equal to the longest run of adjacent points
-#'       in v_1 meeting the threshold that contains the number of candidate
-#'       entries making the threshold cut at each index difference.
-tcor = function(A, t=0.99, p=5, reduce=1)
+#' @return A two-column matrix. Each row of the matrix are the
+#' column indices of vectors that meet the correlation threshold.
+tcor = function(A, t=0.99, p=5)
 {
   mu = colMeans(A)
   s  = sqrt(apply(A,2,crossprod) - nrow(A)*mu^2) # column norms of centered matrix
   L  = irlba(A, p, center=mu, scale=s)
   P  = order(L$v[,1])  # order the entries of v1 (the permutation in the paper)
   limit = sqrt(2*(1-t))/L$d[1]
-# XXX EXPERIMENTAL reduce limit by fraction of variance in 1-d
-#if(reduce) limit = limit*L$d[1]^2/ncol(A) 
-limit = limit*reduce
-# sum of squared singular values of centered/scaled A = ncol(A)
 
 # XXX this is stupid, find a better way ('l' or LaTeX '\ell' in the paper):
   v = L$v[P,1]
   ell = max(vapply(1:length(v), function(i) {x=v[-(1:i)];sum(x-x[1] <= limit)}, 1))
 
-cat("longest run of adjacent points within threshold l =",ell,"\n")
-
-  dnum = rep(0,ell)
-# this is the big union in step 4 of algorithm 2.1, combined with step 6 to
-# convert back to original indices:
-  candidates = lapply(1:ell, function(i)
+# This is the big union in step 4 of algorithm 2.1, combined with step 6 to
[…]

### 2015-11-24 threshold/tcor.R
Updated example algorithm

@@ -2,14 +2,16 @@ require(irlba)
 require(parallel)
 #' Compute the thresholded correlations between columns of a matrix.
 #'
-#' Increase p to cut down the total number of candate pairs.
+#' Increase p to cut down the total number of candidate pairs evaluated,
+#' at the expense of costlier truncated SVDs.
 #'
 #' @param A an m by n real-valued dense or sparse matrix
 #' @param t a threshold value for correlation
 #' @param p projected subspace dimension
 #'
-#' @return A two-column matrix. Each row of the matrix are the
-#' column indices of vectors that meet the correlation threshold.
+#' @return A three-column matrix. The first two columns contain
+#' indices of vectors meeting the correlation threshold \code{t},
+#' and the third column contains the corresponding correlation value.
 tcor = function(A, t=0.99, p=5)
 {
   mu = colMeans(A)
@@ -24,15 +26,19 @@ tcor = function(A, t=0.99, p=5)
 
 # This is the big union in step 4 of algorithm 2.1, combined with step 6 to
 # convert back to original indices, and step 7 to evaluate the candiadtes.
-# Each step from 1 to \ell is independent of the others; the steps may be run
-# in parallel.
+# Each step from 1 to \ell is independent of the others; the steps can run
+# in parallel:
   indices = mclapply(1:ell, function(i)
   {
     d = diff(L$v[P,1:p,drop=FALSE],lag=i)^2 %*% L$d[1:p]^2
     j = which(d <= 2*(1-t))  # These ordered indices meet the threshold
     # return original un-permuted column indices that meet threshold (step 7)
-    if(length(j) > 0) j = j[vapply(j, function(k) cor(A[,P[k]], A[,P[k+i]]) >= t, TRUE)]
-    cbind(P[j], P[j+i])
+    if(length(j) == 0) return()
+    v = vapply(j, function(k) cor(A[, P[k]], A[, P[k+i]]), 1)
+    h = v >= t
+    j = j[h]
+    v = v[h]
+    cbind(i=P[j], j=P[j+i], cor=v)
   })
   Reduce(rbind,indices)
 }

### 2015-11-25 threshold/tcor.R
updated reference implementation

@@ -20,9 +20,21 @@ tcor = function(A, t=0.99, p=5)
   P  = order(L$v[,1])  # order the entries of v1 (the permutation in the paper)
   limit = sqrt(2*(1-t))/L$d[1]
 
-# XXX this is stupid, find a better way ('l' or LaTeX '\ell' in the paper):
   v = L$v[P,1]
-  ell = max(vapply(1:length(v), function(i) {x=v[-(1:i)];sum(x-x[1] <= limit)}, 1))
+# linear time longest run search (A. Poliakov):
+  lower = 1
+  ell = 1
+  for(upper in 2:length(v))
+  {
+    if(v[upper] - v[lower] <= limit)
+    {
+      ell = max(ell, upper - lower + 1)
+    } else
+    {
+      while(lower < upper && v[upper] - v[lower] > limit) lower = lower + 1
+    }
+  }
+
 
 # This is the big union in step 4 of algorithm 2.1, combined with step 6 to
 # convert back to original indices, and step 7 to evaluate the candiadtes.

### 2015-11-25 threshold/tcor.R
minor linting

@@ -15,10 +15,10 @@ require(parallel)
 tcor = function(A, t=0.99, p=5)
 {
   mu = colMeans(A)
-  s  = sqrt(apply(A,2,crossprod) - nrow(A)*mu^2) # column norms of centered matrix
+  s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2) # column norms of centered matrix
   L  = irlba(A, p, center=mu, scale=s)
-  P  = order(L$v[,1])  # order the entries of v1 (the permutation in the paper)
-  limit = sqrt(2*(1-t))/L$d[1]
+  P  = order(L$v[, 1])  # order the entries of v1 (the permutation in the paper)
+  limit = sqrt(2 * (1 - t)) / L$d[1]
 
   v = L$v[P,1]
 # linear time longest run search (A. Poliakov):
@@ -35,22 +35,21 @@ tcor = function(A, t=0.99, p=5)
     }
   }
 
-
 # This is the big union in step 4 of algorithm 2.1, combined with step 6 to
 # convert back to original indices, and step 7 to evaluate the candiadtes.
 # Each step from 1 to \ell is independent of the others; the steps can run
 # in parallel:
   indices = mclapply(1:ell, function(i)
   {
-    d = diff(L$v[P,1:p,drop=FALSE],lag=i)^2 %*% L$d[1:p]^2
-    j = which(d <= 2*(1-t))  # These ordered indices meet the threshold
+    d = diff(L$v[P, 1:p, drop=FALSE], lag=i) ^ 2 %*% L$d[1:p] ^ 2
+    j = which(d <= 2 * (1 - t))  # These ordered indices meet the threshold
     # return original un-permuted column indices that meet threshold (step 7)
     if(length(j) == 0) return()
     v = vapply(j, function(k) cor(A[, P[k]], A[, P[k+i]]), 1)
     h = v >= t
     j = j[h]
     v = v[h]
-    cbind(i=P[j], j=P[j+i], cor=v)
+    cbind(i=P[j], j=P[j + i], cor=v)
   })
-  Reduce(rbind,indices)
+  Reduce(rbind, indices)
 }

### 2015-11-25 threshold/tcor.R
Added informative retun values to reference implementation

@@ -9,9 +9,17 @@ require(parallel)
 #' @param t a threshold value for correlation
 #' @param p projected subspace dimension
 #'
-#' @return A three-column matrix. The first two columns contain
-#' indices of vectors meeting the correlation threshold \code{t},
-#' and the third column contains the corresponding correlation value.
+#' @return A list with three elements:
+#' \enumerate{
+#'   \item \code{cor} A three-column matrix such that the  first two columns contain
+#'         indices of vectors meeting the correlation threshold \code{t},
+#'         and the third column contains the corresponding correlation value.
+#'   \item \code{longest_run} The largest number of successive entries in the
+#'     ordered first singular vector within a projected distance defined by the
+#'     correlation threshold.
+#'   \item \code{candidates} The total number of _possible_ vectors that meet
+#'     the correlation thrshold identified by the algorithm.
+#' }
 tcor = function(A, t=0.99, p=5)
 {
   mu = colMeans(A)
@@ -43,13 +51,22 @@ tcor = function(A, t=0.99, p=5)
   {
     d = diff(L$v[P, 1:p, drop=FALSE], lag=i) ^ 2 %*% L$d[1:p] ^ 2
     j = which(d <= 2 * (1 - t))  # These ordered indices meet the threshold
+    n = length(j)
     # return original un-permuted column indices that meet threshold (step 7)
-    if(length(j) == 0) return()
+    # including the number of possible candidates
+    if(n == 0)
+    {
+      ans = vector("list", 2)
+      names(ans) = c("idx", "n")
+      ans$n = n
+      return(ans)
+    }
     v = vapply(j, function(k) cor(A[, P[k]], A[, P[k+i]]), 1)
     h = v >= t
     j = j[h]
     v = v[h]
-    cbind(i=P[j], j=P[j + i], cor=v)
+    return(list(idx=cbind(i=P[j], j=P[j + i], cor=v), n=n))
   })
-  Reduce(rbind, indices)
+  list(cor=Reduce(rbind, Map(function(x) x$idx, indices)), longest_run=ell,
+         candidates=Reduce(sum, Map(function(x) x$n, indices)))
 }

### 2015-11-25 threshold/experiments.R
Removed old experiments

@@ -1,42 +1,0 @@
-source("tcor.R")  # thresholded correlation prototype function
-
-library(biclust)  # for the EisenYeast and BicatYeast example data
-# A tiny example dataset
-data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
-A = t(BicatYeast) # to correlate genes to genes
-
-t = 0.95
-
-C = cor(A)                     # actual correlation matrix for reference
-C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
-
-# Compare solutions in several dimensions
-x2  = tcor(A,t,p=2)
-x4  = tcor(A,t,p=4)
-x8  = tcor(A,t,p=8)
-x24 = tcor(A,t,p=24)
-
-d = function(i) diag(s$v[,1:i] %*% diag(s$d[1:i]^2) %*% t(s$v[,1:i]))
-
-
-
-p = par(mfrow=c(2,2))
-plot(x2$dnum,ylab="",xlab="index difference",main=paste("2-d projection, num. candidates =",nrow(x2$candidates)))
-plot(x4$dnum,ylab="",xlab="index difference",main=paste("4-d projection, num. candidates =",nrow(x4$candidates)))
-plot(x8$dnum,ylab="",xlab="index difference",main=paste("8-d projection, num. candidates =",nrow(x8$candidates)))
-plot(x24$dnum,ylab="",xlab="index difference",main=paste("24-d projection, num. candidates =",nrow(x24$candidates)))
-par(p)
-
-
-# Shows how conservative the sqrt(2*(1-t))/d[1] bound is
-#A = matrix(rnorm(100*500),nrow=100)
-#C = cor(A)
-#C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
-#t = 0.80*max(C)
-#i = which(C>=t,arr.ind=TRUE)
-#A = scale(A,center=TRUE,scale=sqrt(apply(A,2,crossprod)))
-#s = svd(A)
-#lim = sqrt(2*(1-t))/s$d[1]
-#x = abs(s$v[i[,1],1] - s$v[i[,2],1])
-#plot(x, ylim=c(0,lim*1.2))
-#abline(h=lim,col=2)

### 2015-11-25 threshold/example.R
reference implementation tweaks

@@ -2,16 +2,24 @@
 source("tcor.R")
 
 library(biclust)  # for the EisenYeast and BicatYeast data
-data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
-A = t(BicatYeast) # to correlate genes to genes
+#data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
+#A = t(BicatYeast) # to correlate genes to genes
+data(EisenYeast)  # slightly larger example
+A = t(EisenYeast) # to correlate genes to genes
 
+t = 0.95
+
+t1 = proc.time()
 C = cor(A)                     # actual correlation matrix for reference
 C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
+C = C[C>t]
+print(proc.time() - t1)
 
+t1 = proc.time()
 # Compute candidate pairs with the new algorithm
-t = 0.95
 x = tcor(A, t=t, p=10)
+print(proc.time() - t1)
 
 # Compare
-sum(C > t)
-dim(x)
+print(sum(C > t))
+print(nrow(x$cor))

### 2015-11-25 threshold/tcor.R

@@ -1,5 +1,6 @@
 require(irlba)
-require(parallel)
+require(doParallel)
+
 #' Compute the thresholded correlations between columns of a matrix.
 #'
 #' Increase p to cut down the total number of candidate pairs evaluated,
@@ -8,23 +9,30 @@ require(parallel)
 #' @param A an m by n real-valued dense or sparse matrix
 #' @param t a threshold value for correlation
 #' @param p projected subspace dimension
+#' @param cores number of CPU cores to use in explicitly parallel code sections
+#'  (note that available BLAS libraries may implicitly use multiple threads)
+#' @param additional arguments passed to \code{\link{irlba}}
 #'
 #' @return A list with three elements:
 #' \enumerate{
-#'   \item \code{cor} A three-column matrix such that the  first two columns contain
+#'   \item \code{indices} A three-column matrix such that the  first two columns contain
 #'         indices of vectors meeting the correlation threshold \code{t},
 #'         and the third column contains the corresponding correlation value.
 #'   \item \code{longest_run} The largest number of successive entries in the
 #'     ordered first singular vector within a projected distance defined by the
 #'     correlation threshold.
-#'   \item \code{candidates} The total number of _possible_ vectors that meet
-#'     the correlation thrshold identified by the algorithm.
+#'   \item \code{n} The total number of _possible_ vectors that meet
+#'     the correlation threshold identified by the algorithm.
+#'   \item \code{total_time} Total run time.
 #' }
-tcor = function(A, t=0.99, p=5)
+tcor = function(A, t=0.99, p=5, cores=detectCores(), ...)
 {
+  t0 = proc.time()
   mu = colMeans(A)
-  s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2) # column norms of centered matrix
-  L  = irlba(A, p, center=mu, scale=s)
+  s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2) # col norms of centered matrix
+  if(any(s < 10*.Machine$double.eps)) stop("the standard deviation is zero for some columns")
+  L  = irlba(A, p, center=mu, scale=s, ...)
+  t1 = (proc.time() - t0)[[3]]
   P  = order(L$v[, 1])  # order the entries of v1 (the permutation in the paper)
[…]

### 2015-11-25 threshold/tcor.R
reference implementation tweaks

@@ -35,7 +35,6 @@ tcor = function(A, t=0.99, p=5, cores=detectCores(), ...)
   t1 = (proc.time() - t0)[[3]]
   P  = order(L$v[, 1])  # order the entries of v1 (the permutation in the paper)
   limit = sqrt(2 * (1 - t)) / L$d[1]
-
   v = L$v[P,1]
 # linear time longest run search (A. Poliakov):
   lower = 1

### 2015-11-27 threshold/draft.Rnw
minor tweaks to algo, working on efficiency now...

@@
 along an intuitive path of decreasing variance defined by the data and computed
-by the SVD.
+by a truncated SVD.
 
@@
 relatively low rank $p$.  The choice of $p$ represents a balance between work in
-computing a truncated SVD and work required by the $\ell$ matrix vector
-products of order $n\times p$ in step 4.
+computing a truncated SVD plus work required by step 4 and filtering the
+thresholded values in step 7.
 

### 2015-11-27 threshold/example.R

@@ -2,10 +2,10 @@
 source("tcor.R")
 
 library(biclust)  # for the EisenYeast and BicatYeast data
-#data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
-#A = t(BicatYeast) # to correlate genes to genes
-data(EisenYeast)  # slightly larger example
-A = t(EisenYeast) # to correlate genes to genes
+data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
+A = t(BicatYeast) # to correlate genes to genes
+#data(EisenYeast)  # slightly larger example
+#A = t(EisenYeast) # to correlate genes to genes
 
 t = 0.95
 

### 2015-11-27 threshold/tcor.R

@@ -25,8 +25,9 @@ require(doParallel)
 #'     the correlation threshold identified by the algorithm.
 #'   \item \code{total_time} Total run time.
 #' }
-tcor = function(A, t=0.99, p=5, cores=detectCores(), ...)
+tcor = function(A, t=0.99, p=10, cores=detectCores(), ...)
 {
+  if(ncol(A) < p) p = max(1, floor(ncol(A) / 2 - 1))
   t0 = proc.time()
   mu = colMeans(A)
   s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2) # col norms of centered matrix

### 2015-11-27 threshold/tcor.R
Minor optimization

@@ -1,5 +1,27 @@
 require(irlba)
-require(doParallel)
+require(doMC)
+
+#' linear time longest run search (A. Poliakov), find the longest
+#' run of values in the vector within the specified distance
+#' @param v a vector with entries ordered in increasing order
+#' @param limit distance interval
+#' @return run length
+longrun = function(v, limit)
+{
+  lower = 1
+  ell = 1
+  for(upper in 2:length(v))
+  {
+    if(v[upper] - v[lower] <= limit)
+    {
+      ell = max(ell, upper - lower + 1)
+    } else
+    {
+      while(lower < upper && v[upper] - v[lower] > limit) lower = lower + 1
+    }
+  }
+  ell
+}
 
 #' Compute the thresholded correlations between columns of a matrix.
 #'
@@ -34,34 +56,31 @@ tcor = function(A, t=0.99, p=10, cores=detectCores(), ...)
   if(any(s < 10*.Machine$double.eps)) stop("the standard deviation is zero for some columns")
   L  = irlba(A, p, center=mu, scale=s, ...)
   t1 = (proc.time() - t0)[[3]]
-  P  = order(L$v[, 1])  # order the entries of v1 (the permutation in the paper)
-  limit = sqrt(2 * (1 - t)) / L$d[1]
-  v = L$v[P,1]
-# linear time longest run search (A. Poliakov):
-  lower = 1
-  ell = 1
-  for(upper in 2:length(v))
+# Find the projection among the first few with the shortest maximum run length
+# to minimize work in the next step. This is a cheap but usually not very
+# significant optimization.
+  ells = lapply(1:min(2,p), function(N)
   {
-    if(v[upper] - v[lower] <= limit)
-    {
-      ell = max(ell, upper - lower + 1)
-    } else
-    {
-      while(lower < upper && v[upper] - v[lower] > limit) lower = lower + 1
-    }
-  }
+    P = order(L$v[, N])
+    limit = sqrt(2 * (1 - t)) / L$d[N]
+    ell = longrun(L$v[order(L$v[, N]), N], limit)
+    list(P=P, limit=limit, ell=ell)
+  })
+  ellmin = which.min(vapply(ells, function(x) x$ell, 1))
+  P = ells[[ellmin]]$P
+  limit = ells[[ellmin]]$limit
+  ell = ells[[ellmin]]$ell
 
-# This is the big union in step 4 of algorithm 2.1, combined with step 6 to
[…]

### 2015-11-28 threshold/example.R
example.R

@@ -1,25 +1,33 @@
-# EXAMPLE USE of tcor
 source("tcor.R")
 
 library(biclust)  # for the EisenYeast and BicatYeast data
 data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
 A = t(BicatYeast) # to correlate genes to genes
-#data(EisenYeast)  # slightly larger example
-#A = t(EisenYeast) # to correlate genes to genes
+threshold = 0.95
 
-t = 0.95
 
-t1 = proc.time()
-C = cor(A)                     # actual correlation matrix for reference
-C[lower.tri(C,diag=TRUE)] = 0  # cut out symmetric part and ones on diagonal
-C = C[C>t]
-print(proc.time() - t1)
+# Compute using tcor():
+m1  = sum(gc()[,2])  # Memory in megabytes currently in use
+t1  = proc.time()
+tx  = tcor(A, t=threshold, p=10)
+t1  = (proc.time() - t1)[3] # Time in seconds
+m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
 
-t1 = proc.time()
-# Compute candidate pairs with the new algorithm
-x = tcor(A, t=t, p=10)
-print(proc.time() - t1)
+cat("tcor time (s)                      ", t1, "\n")
+cat("tcor peak excess memory (MB)       ", m1, "\n")
 
-# Compare
-print(sum(C > t))
-print(nrow(x$cor))
+# Compute using R's native cor() function:
+m2  = sum(gc()[,2])
+t2  = proc.time()
+cx  = cor(A)                # full correlation matrix
+diag(cx) = -Inf
+pairs = which(cx >= threshold, arr.ind=TRUE)
+t2  = (proc.time() - t2)[3]
+m2  = sum(gc()[,6]) - m2
+
+cat("brute force time (s)               ", t2, "\n")
+cat("brute force peak excess memory (MB)", m2, "\n")
+
+# Compare results!
+cat("Number of vector pairs meeting threshold found by brute force", nrow(pairs)/2, "\n")
+cat("Number of vector pairs meeting threshold found by tcor       ", nrow(tx[[1]]), "\n")

### 2015-11-28 threshold/draft.Rnw
working on the examples in the paper

@@
 James Baglama\\
-Department of Mathematics,\\University of Rhode Island
+University of Rhode Island
 \and
@@
 establishes a relationship between correlation and Euclidean distance between
-two vectors.
+two zero mean, unit norm vectors.
 
@@
 Including more terms from Equation \ref{vdist} increases our ability to prune
-adjacent pairs of vectors
-below the correlation threshold
-with respect to the ordering defined by $P$.
-For example,
-including only one more term finds that 49\% of adjacent vectors fall below the
-a correlation threshold of $0.95$ in the example shown in Figure \ref{fig1} and
-are pruned by identifying indices such that
+adjacent pairs of vectors below the correlation threshold with respect to the
+ordering defined by $P$.  Including only one more term finds that 49\% of
+adjacent vectors fall below the a correlation threshold of $0.95$ in the
+example shown in Figure \ref{fig1} and are pruned by identifying indices such
+that
 \[
@@
 Algorithm \ref{prune} guarantees that no pruned vector pair exceeds the given
-correlation threshold--the pruning does not produce false negatives.
-The algorithm typically prunes the vast majority of pairs of vectors below the
+correlation threshold--the pruning does not produce false negatives.  The
+algorithm typically prunes the vast majority of pairs of vectors below the
 correlation threshold with truncated singular value decompositions of only
-relatively low rank $p$.  The choice of $p$ represents a balance between work in
-computing a truncated SVD plus work required by step 4 and filtering the
-thresholded values in step 7.
+relatively low rank $p$.  The choice of $p$ represents a balance between work
+in computing a truncated SVD and evaluating $\ell$ rank $p$ matrix products in
+Step 4 (each increasing in work as $p$ increases), and the work required to
+filter the thresholded values in step 7 (decreasing work as $p$ increases).
 
@@
 
+A full/naive thresholded correlation computation requires $O(n^2 m)$ flops.
+Let $p$ be the selected IRLBA dimension. Then the proposed method requires
+$O(\ell n p)$ flops, where $\ell$ is the longest run of ordered entries in
+$v_1$ that meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
+plus the flops required by IRLBA and the post-processing step 7 (also $O(np)$).
+
 \section{Numerical experiments}\label{examples}
 
-A full/naive thresholded correlation computation requires $O(n^2 m)$ flops. Let
-$p$ be the selected IRLBA dimension. Then the proposed method requires
-$O(\ell n p)$ flops, where $\ell$ is the longest run of ordered entries in $v_1$
-that
-meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
-plus the flops required by IRLBA and the post-processing step of evaluating the
-correlation of all the candidate pairs.
+We evaluted the algorithm using three publically-available gene expression
+datasets.  The tests were performed on a desktop PC with a single AMD A10-7850K
+3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered RAM running
+the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R version
+3.1.2 using double-precision arithmetic and the OpenBLAS library version
+0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
+OMP\_NUM\_THREADS=4}.  All examples use the reference R implementation
+function, {\tt tcor()}, which can be installed for the R language from the
+development version of IRLBA GitHub repository using the {\tt devtools} package
+with:
+
+<<eval=FALSE,echo=TRUE>>=
+devtools::install_github("bwlewis/irlba")
+@
+
+\subsection{BicatYeast small example}
+
+The first example is a tiny proof of concept that provides the data for Figure
+\ref{fig1} and illustrates our timing and memory-use measurement methodology.
+Anyone can quickly run the expample to experiment with the algorithm and
+compare its output with brute force results. The example uses the {\tt
+BicatYeast} data from the {\tt biclust} R package and related {\tt BicAT}
+software\cite{biclust, bicat}. The data consist of 70 experiments invovling the
+Saccharomyces Cerevisiae yeast measuring gene expression data across 419
+Affymetrix probes, represented as a $70\times 419$ real-valued matrix.  The
+example finds all pairs of columns (Affymetrix probes) that meet or exceed a
+Pearson correlation value of 0.95.  We use a projected dimension $p=10$ in {\tt tcor()}.
+\begin{small}
+<<eval=FALSE,echo=TRUE>>=
+library(irlba)
+library(biclust)  # for the BicatYeast data
+data(BicatYeast)  # in gene by sample orientation
+A = t(BicatYeast) # to correlate genes to genes
+threshold = 0.95
+
+# Compute using tcor():
+m1  = sum(gc()[,2])  # Memory in megabytes currently in use
+t1  = proc.time()
+tx  = tcor(A, t=threshold, p=10)
+t1  = (proc.time() - t1)[3] # Time in seconds
+m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
+
+cat("tcor time (s)                      ", t1, "\n")
+cat("tcor peak excess memory (MB)       ", m1, "\n")
+
+# Compute using R's native cor() function:
+m2  = sum(gc()[,2])
+t2  = proc.time()
+cx  = cor(A)                # full correlation matrix
+diag(cx) = -Inf             # ignore diagonal
+pairs = which(cx >= threshold, arr.ind=TRUE)
+t2  = (proc.time() - t2)[3]
+m2  = sum(gc()[,6]) - m2
+
+cat("brute force time (s)               ", t2, "\n")
+cat("brute force peak excess memory (MB)", m2, "\n")
+
+# Compare results!
+cat("brute force vector pairs found", nrow(pairs)/2, "\n")
+cat("tcor vector pairs found       ", nrow(tx[[1]]), "\n")
+@
+\end{small}
+Brute-force computation outperforms {\tt tcor()} in this tiny first example,
+both in speed and slightly in memory use.  Each algorithm identifies the same
+115 vector pairs that exceed the specified correlation threshold of 0.95, with
+comparative timing and memory use shown in the following table.
+
+\begin{center}
+\begin{tabular}{l|rr}
+ & {\tt tcor()} & native {\tt cor()} \\
+Wall-clock time (s) & 0.26 & 0.02 \\
+Excess peak memory use (MB) & 29.2 & 26.2 \\
+\end{tabular}
+\end{center}
 
@@
 
-\section{NOTES not for publication}
+Brute-force computation benefits substantially from sustained use of BLAS Level
+3 operations (matrix multiplication), achieving high utilization of available
+CPU floating point capacity. The tcor method also benefits from optimized BLAS
+routines, but not as much as it mostly makes use of Level 2 operations and
+Level 3 operations on sequences of smaller problems.
 
-A full/naive thresholded correlation requires $O(n^2 m)$ flops. Let $p$ be the
-selected IRLBA dimension. Then the proposed method requires $O(\ell  n  p)$
-flops, where $\ell$ is the longest run of ordered entries in $v_1$ that meet
-the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$ (not counting
-from flops due to IRLBA and the post-processing step of evaluating the
-correlation of all the candidate pairs).  One problem with the method as
-written is that $\ell$ can a large fraction of n.
+%<<echo=FALSE, results=tex>>=
+%require(xtable)
+%print(
+%    xtable(head(iris),
+%        caption = "blah", label = "tab:five", table.placement = "!h", caption.placement = "top")
+%  , size = "small", table.placement="ht")
+%@
 
-Unnecessarily large values of $\ell$ are due to the loose bound
-$\sqrt{2(1-t)}/\sigma_1$ used in step 3 of Algorithm\ref{prune} which is only
-sharp when $k=1$. General problems may not need to run through differencing
-steps up to $\ell$ in step 4 of the algorithm. We can further improve the
-efficiency of the algorithm by dynamically reducing $\ell$ in step 4 as
-follows.
-
-Equation \ref{vdist} expresses the Euclidean distance between any two columns
-of the matrix $A$ in terms of singular components. Write the smallest squared
-gap between entries of a particular singular vector $v_h$ as:
-\[
-\delta_h = \min_{i,j} (e_i^T v_h - e_j^T v_h)^2.
-\]
-Then note that
-\begin{equation}\label{mindist}
-\|a_i - a_j\|^2 \ge \sigma_1^2 \delta_1 + \sigma_2^2 \delta_2 + \cdots + \sigma_k^2\delta_k,
-\end{equation}
-for any indices $i,j$.
-
-Fix $1\le \tilde{i}\le \ell$ and let $\nu \in R^p$ be the vector of minimum values in
-each column of the $n\times p$ matrix $D^{\tilde{i}}PV_{1:p}$ from step 4 of
-the algorithm.
-It follows from Equation \ref{mindist} that
-\[
-e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \ge \nu^Ts_{1:p}
-\]
-for all $\tilde{j}=1,2,\ldots,n$.
-
-XXX need monotonicity of $\nu^Ts_{1:p}$ XXX
-
-\section{The distribution of $||a_i - a_j||^2$}
-
-\begin{lemma}
-\label{asymdistm1lemma}
-Let $m=1$, $X_i$ and $X_j$ be distributed as $\mathcal{N}(0, \Delta_{ij})$,
-with $\Delta_{ii} = 1$.  Then
-\begin{equation}
-\label{eqasymdist1}
-|| X_i - X_j ||^2 \sim R + \left(\Delta_{ij} + 1 \right) S + 
-  \left(\Delta_{ij} - 1 \right) T
-\end{equation}
-where $R$ is distributed as $\chi_2^2$ and $S$ and $T$ are $\chi_1^2$.
-\end{lemma}
-\begin{proof}
-\begin{equation*}
-||X_i - X_j||^2 = X_i^2 - 2 X_i X_j + X_j^2
-\end{equation*}
-$X_i^2$ and $X_j^2$ are clearly $\chi_1^2$. To derive the distribution
-of the cross-product term realize
-\begin{equation*}
-X_i X_j = \frac{1}{4} \left(X_i + X_j\right)^2 - 
-  \frac{1}{4} \left(X_i - X_j\right)^2.
-\end{equation*}
-$X_i + X_j$ and $X_i - X_j$ are both distributed as normal and they are 
-independent
-since $\mathbf{E}X_1^2 = \mathbf{E}X_2^2$. Therefore, their squares
-are $\chi_1^2$ and the leading coefficients in Equation \ref{eqasymdist1}
-are the respective variances.
-\end{proof}
-
-\begin{lemma}
-\label{asymdistlemma}
-Let $a_i$ and $a_j$ be any two columns of the matrix 
-$B \in \mathcal{R}^{m \times n}$, sampled from 
-$\mathcal{N}(0, \Delta/m)$ where $\Delta_{ii} = 1$ for $1 \leq i \leq n$.
-Then
-\begin{equation*}
-||a_i - a_j||^2 \xrightarrow[m]{D} \mathcal{N} \left(2 - 2 \Delta_{ij},  
-  \frac{6 + 4\Delta_{ij}^2}{m} \right) 
-\end{equation*}
-\end{lemma}
-This result follows by applying the central limit theorem to
-the result in Lemma \ref{asymdistm1lemma} where $m$ is appropriately large.
-
+\section{Extensions}\label{extensions}
 \section{Conclusion}\label{conclusion}
@@
 
+%% BicatYeast data
+\bibitem{bicat} Barkow, S., Bleuler, S., Prelic, A., Zimmermann, P., and E. Zitzler. BicAT: a biclustering analysis toolbox, Bioinformatics, 2006 22(10):1282-1283.
+
 %% Applications in machine learning
@@
 Hall, Mark A. Correlation-based feature selection for machine learning. Diss. The University of Waikato, 1999.
+
+%% R's biclust pacakge
+\bibitem{biclust}
+Kaiser, Sebastian, Santamaria, Rodrigo, Khamiakova, Tatsiana, Sill, Martin,
+Theron, Roberto, Quintales, Luis, Leisch, Friederich, and De Troyer, Ewoud, 
+\url{https://cran.r-project.org/web/packages/biclust}.
 
@@
 Lewis, Bryan,  \url{https://cran.r-project.org/web/packages/irlba}.
+
+\bibitem{tcor}
+Lewis, Bryan,  \url{https://github.com/bwlewis/covar/blob/master/threshold/tcor.R}.
 

### 2015-11-28 threshold/draft.Rnw
More draft notes

@@
 ordering defined by $P$.  Including only one more term finds that 49\% of
-adjacent vectors fall below the a correlation threshold of $0.95$ in the
+adjacent vectors fall below the correlation threshold of $0.95$ in the
 example shown in Figure \ref{fig1} and are pruned by identifying indices such
@@
 function, {\tt tcor()}, which can be installed for the R language from the
-development version of IRLBA GitHub repository using the {\tt devtools} package
-with:
+development GitHub repository using the {\tt devtools} package with:
 
 <<eval=FALSE,echo=TRUE>>=
-devtools::install_github("bwlewis/irlba")
+devtools::install_github("bwlewis/tcor")
 @
 
-\subsection{BicatYeast small example}
+The native R {\tt cor()} function is a general function that can compute
+several different types of correlation matrices including Pearson, Spearman,
+and Kendall and includes a number of options for dealing with missing values.
+Because of its flexibility, the native {\tt cor()} function does not always
+compute Pearson correlation matrices in the most computationally-efficient way.
+
+Correlation matrix computation can benefit substantially from use of optimized
+BLAS Level 3 operations (matrix multiplication), achieving high utilization of
+available CPU floating point capacity. The {\tt tcor()} function also benefits
+from optimized BLAS routines, but not as much as it mostly makes use of Level 2
+operations and Level 3 operations on sequences of smaller problems.
+
+We compare the {\tt tcor()} function with brute-force Pearson correlation
+matrix computation computed in the fastest way we can, directly using matrix
+multiplication instead of using the slower {\tt cor()} function.
+
+\subsection{Small BicatYeast example}
 
 The first example is a tiny proof of concept that provides the data for Figure
-\ref{fig1} and illustrates our timing and memory-use measurement methodology.
+\ref{fig1} and establishes our timing and memory use measurement methodology.
+Wall-clock times are reported in seconds and peak memory use in excess of
+memory required to store the input data matrix is reported in megabytes.
+
 Anyone can quickly run the expample to experiment with the algorithm and
@@
 \begin{small}
-<<eval=FALSE,echo=TRUE>>=
+<<eval=TRUE,echo=TRUE>>=
 library(irlba)
+source("tcor.R")  # XXX Replace later with tcor pacakge
 library(biclust)  # for the BicatYeast data
@@
 
-cat("tcor time (s)                      ", t1, "\n")
-cat("tcor peak excess memory (MB)       ", m1, "\n")
-
-# Compute using R's native cor() function:
+# Fast brute force full correlation:
 m2  = sum(gc()[,2])
 t2  = proc.time()
-cx  = cor(A)                # full correlation matrix
-diag(cx) = -Inf             # ignore diagonal
+mu = colMeans(A)
+s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
+cs = scale(A, center=mu, scale=s)  # scale and center the data
+cx = crossprod(cs)                 # full correlation matrix
+cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
 pairs = which(cx >= threshold, arr.ind=TRUE)
@@
 m2  = sum(gc()[,6]) - m2
-
-cat("brute force time (s)               ", t2, "\n")
-cat("brute force peak excess memory (MB)", m2, "\n")
-
-# Compare results!
-cat("brute force vector pairs found", nrow(pairs)/2, "\n")
-cat("tcor vector pairs found       ", nrow(tx[[1]]), "\n")
 @
@@
 both in speed and slightly in memory use.  Each algorithm identifies the same
-115 vector pairs that exceed the specified correlation threshold of 0.95, with
-comparative timing and memory use shown in the following table.
+115 unique vector pairs that exceed the specified correlation threshold of
+0.95, with comparative timing and memory use shown in the following table.
 
-\begin{center}
-\begin{tabular}{l|rr}
- & {\tt tcor()} & native {\tt cor()} \\
-Wall-clock time (s) & 0.26 & 0.02 \\
-Excess peak memory use (MB) & 29.2 & 26.2 \\
-\end{tabular}
-\end{center}
+<<echo=FALSE, results=tex>>=
+require(xtable)
+ans = data.frame(Method=c("tcor","cor"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
+print(
+    xtable(ans,
+        caption = "BicatYeast results, threshold=0.95", label = "BicatYeast", caption.placement = "top")
+  ,size="small", table.placement="ht", include.rownames=FALSE)
+@
+
+\subsection{TCGA Gene Expression Examples}
+
+Performance benefits from the {\tt tcor()} function become apparent whith large
+data.  The next example uses RPKM-normalized gene expression data measured
+across 20,532 genes from 878 breast cancer tumor samples obtained from the
+Broad GDAC dashboard\cite{gdac} of the Cancer Genome Atlas data. See the
+supplemental code at XXX  (also embedded in this Sweave document) for examples
+of efficiently converting the downloaded sample expression data into an R
+matrix.
+
+Eight columns with constant zero values were removed from the data in a
+pre-processing step, leaving an $878\times 20524$ matrix. The example computes
+pairs of gene expression column vectors with correlation values of 0.99
+or more. We use the same {\tt tcor} projection dimension of 10 and performance
+measurement methodology as used in the first example.
+
+Both methods find the same 3,572 pairs of vectors. The fast brute-force
+computation is slightly faster than the {\tt tcor()} function largely due to
+the still relatively small input matrix and better utilization of optimized
+BLAS routines. But the {\tt tcor()} function uses far less memory.  (For
+comparison, R's {\tt cor()} function took about 372 seconds and about
+3200$\,$MB RAM on the same problem on the same test system.)
+
+<<echo=FALSE, results=tex>>=
+# --------------------------------------------------------------------------------------------------
+# The first time you run this, you need to download the data as outlined in the
+# following commented-out section. This code creates a file named
+# 'rpkm.RData' used subsequently.
+#
+# To run any of the remaining (large) examples, first set the system environment
+# variable RUN_EXAMPLES=TRUE (otherwise they won't run and representative results
+# from Bryan's test machine will be printed).
+#
+#library(doMC)
+# Download and extract this file with, for example:
+# wget  http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/BRCA/20150821/gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
+# tar zxf gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
+# cd gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0
+# R --slave < read.R   # (run this file)
+#f = "BRCA.rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.data.txt"
+#N = 20534  # total number of lines
+#h = 2      # header lines
+# Read the first header line with sampe IDs:
+#id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
+# Read the 2nd header line with sample type (raw, RPKM, etc.)
+#header = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=FALSE, nrows=1))
+# Read the rest of the file in parallel. I prefer doMC/foreach over mclapply
+# for this because foreach can combine results incrementally.
+#cores = 4
+#registerDoMC(cores)
+#block = floor((N - h)/cores)
+#t1 = proc.time()
+#ans = foreach(j=1:cores, .combine=rbind) %dopar%
+#{
+#    skip = block * (j - 1) + h
+#    nrows = ifelse(j == cores, -1, block)
+#    read.table(f, sep="\t", stringsAsFactors=FALSE, skip=skip, header=FALSE, nrows=nrows)
+#}
+#names(ans) = header
+#print(proc.time() - t1)
+# (All timings shown for a quad-core Athlon home PC reading from an older
+# OCZ REVODRIVE3 X2 SSD drive (ca 2014).)
+#   user  system elapsed
+#494.549   4.134 131.232
+# Compare with sequential approach:
+#t1 = proc.time()
+#x = read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=TRUE)
+#print(proc.time() - t1)
+#   user  system elapsed
+#517.430   1.318 519.553
+# Create a samples x gene matrix of RPKM values
+#genes = ans[, 1] # gene labels
+#idx   = grep("RPKM", names(ans))    # Select columns with RPKM values
+#rpkm     = t(as.matrix(ans[, idx]))
+# Label the coordinates by sample ID and gene:
+#rownames(rpkm) = id[idx]
+#colnames(rpkm) = genes
+#save(rpkm, file="rpkm.RData")
+# --------------------------------------------------------------------------------------------------
+# Default values:
+t1=40.34
+t2=35.60
+m1=39.10
+m2=9595.10
+
+if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
+{
+load("rpkm.RData")  # rpkm expression matrix
+
+# The data have a few columns of all zeros that need to be removed
+n = ncol(rpkm)
+rpkm = rpkm[, colMeans(rpkm) > 0]
+
+threshold = 0.99     # correlation threshold
+xx  = gc()           # clear
+m1  = sum(gc()[,2])  # Memory in megabytes currently in use
+t1  = proc.time()
+tx  = tcor(rpkm, threshold, 11)
+t1  = (proc.time() - t1)[3] # Time in seconds
+m1  = sum(gc()[,6] - m1)    # Peak excess memory use during the test
+
+# Compare with the fastest way we know of to compute a thresholded correlation
+# matrix by brute force. This approach makes heavy use of fast multi-threaded
+# BLAS routines. This approach is much faster than R's default `cor` function.
+xx  = gc()
+m2  = sum(gc()[,2])  # Memory in megabytes currently in use
+t2 = proc.time()
+mu = colMeans(rpkm)
+s  = sqrt(apply(rpkm, 2, crossprod) - nrow(rpkm) * mu ^ 2) # col norms of centered matrix
+cs = scale(rpkm, center=mu, scale=s)  # scale and center the data
+bx = crossprod(cs)                    # full correlation matrix
+diag(bx) = 0                          # ignore diagonal (all ones)
+pairs = which(bx >= threshold, arr.ind=TRUE)
+t2  = (proc.time() - t2)[3] # Time in seconds
+m2  = sum(gc()[,6] - m2)    # Peak excess memory use during the test
+} # RUN_EXAMPLES=TRUE
+
+require(xtable)
+ans = data.frame(Method=c("tcor","cor"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
+print(
+    xtable(ans,
+        caption = "TCGA BRCA gene expression results (878 samples x 20,542 genes, threshold=0.99)", label = "TCGA1", caption.placement = "top")
+  ,size="small", table.placement="ht", include.rownames=FALSE)
+@
+
+We simulated a data set with a larger number of samples by simply replicating
+the rows of the gene expression matrix 27 times to obtain a matrix with 23,706
+rows and 20,524 columns, noting that the simulated data have the same rank (of
+at most 878) and the same expected number of vector pairs meeting the
+correlation threshold as the original data. Each method again computed the same
+set of  3,572 pairs of vectors metting the correlation threshold of 0.99, but
+the {\tt tcor()} function outperforms the brute force method in both compute
+time and memory consumption.
+
+<<echo=FALSE, results=tex>>=
+# Default values:
+t1=372.69
+t2=546.06
+m1=768.3
+m2=9600.2
+
+if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
+{
+  # simulated data, replicate rows 27x
+}
+require(xtable)
+ans = data.frame(Method=c("tcor","cor"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
+print(
+    xtable(ans,
+        caption = "TCGA simulated gene expression results (23,706 rows x 20,542 columns, threshold=0.99)", label = "TCGA2", caption.placement = "top")
+  ,size="small", table.placement="ht", include.rownames=FALSE)
+@
 
@@
 
-Brute-force computation benefits substantially from sustained use of BLAS Level
-3 operations (matrix multiplication), achieving high utilization of available
-CPU floating point capacity. The tcor method also benefits from optimized BLAS
-routines, but not as much as it mostly makes use of Level 2 operations and
-Level 3 operations on sequences of smaller problems.
+\section{Extensions}\label{extensions}
 
-%<<echo=FALSE, results=tex>>=
-%require(xtable)
-%print(
-%    xtable(head(iris),
-%        caption = "blah", label = "tab:five", table.placement = "!h", caption.placement = "top")
-%  , size = "small", table.placement="ht")
-%@
+XXX Two basic extensions, include functions/options for this in the package.
 
-\section{Extensions}\label{extensions}
+The main ideas in Algorithm \ref{prune}, although applied to correlation,
+easily extend to any distance metric.
+
+Algorithm \ref{prune} can be adapted to find pairs of highly anti-correlated
+vectors by changing the distance metric...XXX
+
 \section{Conclusion}\label{conclusion}
@@
 Baglama, James, and Lothar Reichel. Augmented implicitly restarted Lanczos bidiagonalization methods. SIAM Journal on Scientific Computing 27.1 (2005): 19-42.
+
 \bibitem{irlbam}
@@
 \bibitem{bicat} Barkow, S., Bleuler, S., Prelic, A., Zimmermann, P., and E. Zitzler. BicAT: a biclustering analysis toolbox, Bioinformatics, 2006 22(10):1282-1283.
+
+%% BROAD GDAC dashboard
+\bibitem{gdac}{\url{http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/BRCA/20150821/gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz}}
 

### 2015-11-28 threshold/draft.Rnw
More examples

@@
 
+Most of the floating point operations occur in step 4 of Algorithm \ref{prune}.
+Fortunately the $\ell$ evaluations in step 4 are independent of each other and
+can easily be computed in parallel. See the reference R
+implementation\cite{sup} for an example.
+
 \section{General matrices and fast truncated SVD}\label{irlba}
@@
 available CPU floating point capacity. The {\tt tcor()} function also benefits
-from optimized BLAS routines, but not as much as it mostly makes use of Level 2
-operations and Level 3 operations on sequences of smaller problems.
+from optimized BLAS routines, but not as much, as it mostly makes use of Level
+2 operations and Level 3 operations on sequences of smaller problems.
 
@@
 require(xtable)
-ans = data.frame(Method=c("tcor","cor"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
+ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
 print(
@@
 
-Performance benefits from the {\tt tcor()} function become apparent whith large
+Performance benefits from the {\tt tcor()} function become apparent with more
 data.  The next example uses RPKM-normalized gene expression data measured
 across 20,532 genes from 878 breast cancer tumor samples obtained from the
-Broad GDAC dashboard\cite{gdac} of the Cancer Genome Atlas data. See the
-supplemental code at XXX  (also embedded in this Sweave document) for examples
-of efficiently converting the downloaded sample expression data into an R
-matrix.
+Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}. See the supplemental
+code\cite{sup} (also embedded in this Sweave document) for examples of
+efficiently converting the downloaded sample expression data into an R matrix.
 
-Eight columns with constant zero values were removed from the data in a
-pre-processing step, leaving an $878\times 20524$ matrix. The example computes
+Eight columns with constant zero values were removed from the data,
+leaving an $878\times 20524$ matrix. The example computes
 pairs of gene expression column vectors with correlation values of 0.99
@@
 correlation threshold as the original data. Each method again computed the same
-set of  3,572 pairs of vectors metting the correlation threshold of 0.99, but
+set of  3,572 pairs of vectors meeting the correlation threshold of 0.99, but
 the {\tt tcor()} function outperforms the brute force method in both compute
@@
         caption = "TCGA simulated gene expression results (23,706 rows x 20,542 columns, threshold=0.99)", label = "TCGA2", caption.placement = "top")
+  ,size="small", table.placement="ht", include.rownames=FALSE)
+@
+
+\subsection{TCGA DNA methylation example}
+
+The following example uses calculated beta values from the Illumina Human
+Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor samples
+obtained from the Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}. See the
+supplemental code\cite{sup} for examples of efficiently converting the
+downloaded sample methylation data into an R matrix.
+
+We removed 91,563 columns with all constant or missing values from the data,
+leaving an $80\times 394014$ matrix. The example computes
+pairs of methylation beta value column vectors with correlation values of 0.99
+or more. We use the same {\tt tcor} projection dimension of 10 and performance
+measurement methodology as used in the first example.
+
+Algorithm \ref{prune} identified 913,601 pairs of vectors with correlation
+values of 0.99 or more in about three hours on our test PC. A brute force
+algorithm would need to evaluate over 155 billion vector pairs; the projection
+of Algorithm \ref{prune} limited the search space to only about 4 million
+candidate pairs, from which the 913,601 correlated pairs were found.
+
+<<echo=FALSE, results=tex>>=
+# Default values:
+t1=10449.670
+t2=NA
+m1=1261.6
+m2=NA
+
+if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
+{
+  # simulated data, replicate rows 27x
+}
+require(xtable)
+ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
+print(
+    xtable(ans,
+        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99)", label = "TCGA3", caption.placement = "top")
   ,size="small", table.placement="ht", include.rownames=FALSE)
@@
 
-XXX Two basic extensions, include functions/options for this in the package.
+XXX
 
-The main ideas in Algorithm \ref{prune}, although applied to correlation,
+1. The main ideas in Algorithm \ref{prune}, although applied to correlation,
 easily extend to any distance metric.
 
-Algorithm \ref{prune} can be adapted to find pairs of highly anti-correlated
-vectors by changing the distance metric...XXX
+2. Anti-correlated pairs? Anyone?
 
@@
 %% BROAD GDAC dashboard
-\bibitem{gdac}{\url{http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/BRCA/20150821/gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz}}
+\bibitem{gdac}{Broad Institute TCGA Genome Data Analysis Center (2015): Analysis-ready standardized TCGA data from Broad GDAC Firehose stddata\_\_2015\_06\_01 run.
+Broad Institute of MIT and Harvard.
+Dataset.
+\url{http://dx.doi.org/10.7908/C1251HBG}}
 
@@
 
-%% IRLBA implementations
+%% IRLBA implementations and supplemental code
 \bibitem{irlbar}
@@
 
-\bibitem{tcor}
-Lewis, Bryan,  \url{https://github.com/bwlewis/covar/blob/master/threshold/tcor.R}.
+\bibitem{sup}
+Lewis, Bryan,  \url{https://github.com/bwlewis/tcor}.
 

### 2015-11-28 threshold/draft.Rnw
fixed a few typos, still working out examples

@@
 filtering out coefficients below a given threshold may not be computationally
-feasable for high-dimensional data in modern genomics and other applications.
+feasible for high-dimensional data in modern genomics and other applications.
 Considerable attention has been devoted to \emph{pruning
@@
 vectors.  For example the TAPER and related algorithms of Xiong et al. develop
-pruning rules based on the frequency of mutually co-ocurring observations (rows)
+pruning rules based on the frequency of mutually co-occurring observations (rows)
 among vector pairs; see \cite{prune1, prune2} and the references therein.
@@
 
-We evaluted the algorithm using three publically-available gene expression
-datasets.  The tests were performed on a desktop PC with a single AMD A10-7850K
-3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered RAM running
-the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R version
-3.1.2 using double-precision arithmetic and the OpenBLAS library version
-0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
-OMP\_NUM\_THREADS=4}.  All examples use the reference R implementation
-function, {\tt tcor()}, which can be installed for the R language from the
-development GitHub repository using the {\tt devtools} package with:
+We evaluated the algorithm using three public gene expression and methylation
+datasets.  Most tests were performed on a desktop PC with a single AMD
+A10-7850K 3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered
+RAM running the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R
+version 3.1.2 using double-precision arithmetic and the OpenBLAS library
+version 0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
+OMP\_NUM\_THREADS=4}. The last experiment includes a test run on a Linux
+cluster as described in the text.
+
+All examples use the reference R implementation, {\tt tcor()}, which can be
+installed for the R language from the development GitHub repository using the
+{\tt devtools} package with:
 
@@
 
-Anyone can quickly run the expample to experiment with the algorithm and
+Anyone can quickly run the example to experiment with the algorithm and
 compare its output with brute force results. The example uses the {\tt
 BicatYeast} data from the {\tt biclust} R package and related {\tt BicAT}
-software\cite{biclust, bicat}. The data consist of 70 experiments invovling the
+software\cite{biclust, bicat}. The data consist of 70 experiments involving the
 Saccharomyces Cerevisiae yeast measuring gene expression data across 419
@@
 #h = 2      # header lines
-# Read the first header line with sampe IDs:
+# Read the first header line with sample IDs:
 #id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
@@
 
+We illustrate that step 4 of Algorithm \ref{prune} can be computed in parallel
+by comparing the above results run on our reference test PC with results run on
+a modest GNU/Linux cluster at Paradigm4, Inc. The cluster consisted of four
+computers connected by 10 gigabit ethernet. Each computer contained two Intel
+Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per computer) and 64$\,$GB
+ECC registered RAM running at 1,600$\,$MHz running CentOS 6.6 and 64-bit R
+version 3.2.2 with the reference R BLAS library.
+
+No code change was required to run in parallel on the cluster. The reference R
+{\tt tcor()} implementation uses R's {\tt foreach}\cite{foreach} framework to
+run on arbitrary parallel configurations (``back ends''). We simply registered
+a {\tt doRedis}\cite{doredis} parallel back end with 16 R workers per computer
+prior to invoking {\tt tcor()}.
+
+The results computed on the Linux cluster are marked ``tcor (64 core cluster)''
+below.
+
 <<echo=FALSE, results=tex>>=
@@
 t2=NA
+t3=NA
 m1=1261.6
 m2=NA
+m3=NA
 
@@
 require(xtable)
-ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
+ans = data.frame(Method=c("tcor (4 core PC)","brute force","tcor (64 core cluster)"), `Wall-clock time (s)`=c(t1, t2, t3), `Peak excess memory use (MB)`=c(m1, m2, m3), check.names=FALSE)
 print(
@@
         caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99)", label = "TCGA3", caption.placement = "top")
-  ,size="small", table.placement="ht", include.rownames=FALSE)
+  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="NA")
 @

### 2015-11-29 threshold/tcor.R
just register whatever foreach backend you want to use

@@ -1,5 +1,5 @@
 require(irlba)
-require(doMC)
+require(foreach)
 
 #' linear time longest run search (A. Poliakov), find the longest
 #' run of values in the vector within the specified distance
@@ -31,8 +31,6 @@ longrun = function(v, limit)
 #' @param A an m by n real-valued dense or sparse matrix
 #' @param t a threshold value for correlation
 #' @param p projected subspace dimension
-#' @param cores number of CPU cores to use in explicitly parallel code sections
-#'  (note that available BLAS libraries may implicitly use multiple threads)
 #' @param additional arguments passed to \code{\link{irlba}}
 #'
 #' @return A list with three elements:
@@ -47,7 +45,7 @@ longrun = function(v, limit)
 #'     the correlation threshold identified by the algorithm.
 #'   \item \code{total_time} Total run time.
 #' }
-tcor = function(A, t=0.99, p=10, cores=detectCores(), ...)
+tcor = function(A, t=0.99, p=10, ...)
 {
   if(ncol(A) < p) p = max(1, floor(ncol(A) / 2 - 1))
   t0 = proc.time()
@@ -74,13 +72,11 @@ tcor = function(A, t=0.99, p=10, cores=detectCores(), ...)
 # The big union in step 4 of algorithm 2.1 follows, combined with step 6 to
 # convert back to original indices, and step 7 to evaluate the candiadtes.
 # Each step from 1 to ell is independent of the others; the steps can run
-# in parallel. Warning, use a fork-safe BLAS library with the doMC package.
-# Alternatively use the doParallel pacakge with a socket-based cluster.
+# in parallel.
   combine = function(x, y)
   {
     list(idx=rbind(x$idx, y$idx), n=x$n + y$n)
   }
-  registerDoMC(cores)
 
   indices = foreach(i=1:ell, .combine=combine, .inorder=FALSE) %dopar%
   {

### 2015-11-30 threshold/draft.Rnw
sequential filter option

@@
-%%
+%
 %% Instructions for processing this document into a PDF document
@@
 
-We illustrate that step 4 of Algorithm \ref{prune} can be computed in parallel
-by comparing the above results run on our reference test PC with results run on
-a modest GNU/Linux cluster at Paradigm4, Inc. The cluster consisted of four
-computers connected by 10 gigabit ethernet. Each computer contained two Intel
-Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per computer) and 64$\,$GB
-ECC registered RAM running at 1,600$\,$MHz running CentOS 6.6 and 64-bit R
-version 3.2.2 with the reference R BLAS library.
+We illustrate that steps 4--7 of Algorithm \ref{prune} can be computed in
+parallel by comparing the above results run on our reference test PC with
+results run on a modest GNU/Linux cluster at Paradigm4, Inc. The cluster
+consisted of four computers connected by 10 gigabit ethernet. Each computer
+contained two Intel Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per
+computer) and 64$\,$GB ECC registered RAM running at 1,600$\,$MHz running
+CentOS 6.6 and 64-bit R version 3.2.2 with the reference R BLAS library.
 
@@
 a {\tt doRedis}\cite{doredis} parallel back end with 16 R workers per computer
-prior to invoking {\tt tcor()}.
+(64 total R worker processes) prior to invoking {\tt tcor()}.
 
 The results computed on the Linux cluster are marked ``tcor (64 core cluster)''
-below.
+below. The cluster example computed the filtering step 7 of Algorithm
+\ref{prune} in parallel across the 64 remote R processes. That approach incurs,
+in this example, an additional memory overhead of about 16$\,$GB because each
+remote R process requires access to a copy of the data matrix $A$. More thrifty
+adaptations of the algorithm are possible and could, for example, share the
+memory required by $A$ between worker R processes on each computer.
 
@@
 t1=10449.670
-t2=NA
-t3=NA
+t3=482
+t4=NA
 m1=1261.6
-m2=NA
-m3=NA
+m3=994.3
+m4=42282.4
 
@@
 {
-  # simulated data, replicate rows 27x
+ # methylation data
 }
 require(xtable)
-ans = data.frame(Method=c("tcor (4 core PC)","brute force","tcor (64 core cluster)"), `Wall-clock time (s)`=c(t1, t2, t3), `Peak excess memory use (MB)`=c(m1, m2, m3), check.names=FALSE)
+ans = data.frame(Method=c("tcor (4 core PC)", "tcor (64 core cluster)", ""), `Wall-clock time (s)`=c(t1, t3, t4), `Memory use (MB)`=c(m1, m3, m4), " "=c("","(local)","(remote)"), check.names=FALSE)
 print(
     xtable(ans,
-        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99)", label = "TCGA3", caption.placement = "top")
-  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="NA")
+        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99).
+Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).", label = "TCGA3", caption.placement = "top")
+  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="")
 @
+From a set of almost 400 thousand vectors, Algorithm \ref{prune} finds 913,601
+pairs of vectors with correlation values of 0.99 or more on a modest Linux
+cluster in about 8 minutes. The same algorithm also runs well in constrained
+memory settings.
 
@@
 
+%% doredis
+\bibitem{doredis}
+Lewis, Bryan. The doRedis package for R. \url{https://cran.r-project.org/web/packages/doRedis}.
+
 %% IRLBA implementations and supplemental code
 \bibitem{irlbar}
-Lewis, Bryan,  \url{https://cran.r-project.org/web/packages/irlba}.
+Lewis, Bryan. The irlba package for R.  \url{https://cran.r-project.org/web/packages/irlba}.
 
 \bibitem{sup}
-Lewis, Bryan,  \url{https://github.com/bwlewis/tcor}.
+Lewis, Bryan. The tcor package for R.  \url{https://github.com/bwlewis/tcor}.
 
@@
 Song, Lin, Peter Langfelder, and Steve Horvath. Comparison of co-expression measures: mutual information, correlation, and model based indices. BMC bioinformatics 13.1 (2012): 328.
+
+%% foreach
+\bibitem{foreach}
+Weston, Steve. The foreach package for R. \url{https://cran.r-project.org/web/packages/foreach/vignettes/foreach.pdf}.
 

### 2015-11-30 threshold/tcor.R

@@ -31,7 +31,10 @@ longrun = function(v, limit)
 #' @param A an m by n real-valued dense or sparse matrix
 #' @param t a threshold value for correlation
 #' @param p projected subspace dimension
-#' @param additional arguments passed to \code{\link{irlba}}
+#' @param filter "local" filters candidate set sequentially,
+#'  "distributed" computes thresholded correlations in a parallel code section which can be
+#'  faster but requires that the data matrix is available (see notes).
+#' @param ... additional arguments passed to \code{\link{irlba}}
 #'
 #' @return A list with three elements:
 #' \enumerate{
@@ -45,8 +48,9 @@ longrun = function(v, limit)
 #'     the correlation threshold identified by the algorithm.
 #'   \item \code{total_time} Total run time.
 #' }
-tcor = function(A, t=0.99, p=10, ...)
+tcor = function(A, t=0.99, p=10, filter=c("distributed", "local"), ...)
 {
+  filter = match.arg(filter)
   if(ncol(A) < p) p = max(1, floor(ncol(A) / 2 - 1))
   t0 = proc.time()
   mu = colMeans(A)
@@ -78,7 +82,34 @@ tcor = function(A, t=0.99, p=10, ...)
     list(idx=rbind(x$idx, y$idx), n=x$n + y$n)
   }
 
-  indices = foreach(i=1:ell, .combine=combine, .inorder=FALSE) %dopar%
+  if(filter == "distributed")
+  {
+    indices = foreach(i=1:ell, .combine=combine, .inorder=FALSE) %dopar%
+    {
+      d = diff(L$v[P, 1:p, drop=FALSE], lag=i) ^ 2 %*% L$d[1:p] ^ 2
+      # These ordered indices meet the projected threshold:
+      j = which(d <= 2 * (1 - t))
+      n = length(j)
+      # return original un-permuted column indices that meet true threshold
+      # (step 7), including the number of possible candidates for info.:
+      if(n == 0)
+      {
+        ans = vector("list", 2)
+        names(ans) = c("idx", "n")
+        ans$n = n
+        return(ans)
+      }
+      v = vapply(j, function(k) cor(A[, P[k]], A[, P[k+i]]), 1)
+      h = v >= t
+      j = j[h]
+      v = v[h]
[…]

### 2015-12-09 threshold/draft.Rnw
switched to bibtex, changed examples
Dropped rnaseq example and using a different small example

@@
 The following lemma expresses the Euclidean distance between any two columns of
-the matrix $A$ as a weighted sum of projected interval distances along the SVD
+the matrix $A$ as a weighted sum of projected distances along the SVD
 basis vectors forming the columns of $V$.
 \begin{lemma}
+\label{sumlemma}
 Let $a_i$ and $a_j$
@@
 of $v_1$ in increasing order.  For example, Figure \ref{fig1} displays the
-ordered entries of $Pv_1$ for the small ``BicatYeast'' example (where
-$A\in\mathcal{R}^{70\times 419}$) from Section \ref{examples}.  The lines in
+ordered entries of $Pv_1$ for the small ``EisenYeast'' example (where
+$A\in\mathcal{R}^{80 \times 6221}$) from Section \ref{examples}.  The lines in
 the plot illustrate the interval associated with the correlation threshold
-$t=0.95$ placed at an arbitrary vertical axis location.
+$t=0.99$ placed at an arbitrary vertical axis location.
 
@@
 library(biclust)
-data(BicatYeast)
-A=t(BicatYeast)
-s=svd(A)
-i=order(s$v[,1])
+data(EisenYeast)
+A = t(EisenYeast)
+A = sweep(A, 2, apply(A,2,mean), FUN=`-`)
+A = sweep(A, 2, sqrt(apply(A,2,crossprod)), FUN=`/`)
+s = svd(A)
+i = order(s$v[,1])
 plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
-t=0.95
-abline(h=c(-0.05,-0.05+sqrt(2*(1-t))/s$d[1]),col=2,lwd=0.5)
+t=0.99
+abline(h=c(0, sqrt(2*(1-t))/s$d[1]), col=2, lwd=1)
 @
 \caption{
-Example plot of the ordered entries of $v_1$ for the BicatYeast example from
+Example plot of the ordered entries of $v_1$ for the EisenYeast example from
 Section \ref{examples}. The red lines illustrate the interval corresponding
-threshold $t=0.95$ placed at an arbitrary location on the vertical axis. Points
+threshold $t=0.99$ placed at an arbitrary location on the vertical axis. Points
 farther apart than this interval correspond to column vectors that do not meet
@@
 adjacent pairs of vectors below the correlation threshold with respect to the
-ordering defined by $P$.  Including only one more term finds that 49\% of
-adjacent vectors fall below the correlation threshold of $0.95$ in the
+ordering defined by $P$.  Including only one more term finds that 96\% of
+adjacent vectors fall below the correlation threshold of $0.99$ in the
 example shown in Figure \ref{fig1} and are pruned by identifying indices such
@@
 Note that we use the permutation $P$ defined by the order of vector $v_1$
-throughout. Including ten terms prunes over 96\% of adjacent vector pairs,
-leaving only 15 candidate pairs of adjacent vectors that might possibly
-meet the correlation threshold.
-In general, using $p\le k$ terms to prune pairs of vectors below the
-correlation threshold boils down to evaluating the expression
+throughout.  In general, using $p\le k$ terms to prune pairs of vectors
+below the correlation threshold boils down to evaluating the expression
 \begin{equation}\label{proj}
@@
 
-A full/naive thresholded correlation computation requires $O(n^2 m)$ flops.
+A full/na\"ive thresholded correlation computation requires $O(n^2 m)$ flops.
 Let $p$ be the selected IRLBA dimension. Then the proposed method requires
-$O(\ell n p)$ flops, where $\ell$ is the longest run of ordered entries in
+$O(n p \ell)$ flops, where $\ell$ is the longest run of ordered entries in
 $v_1$ that meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
@@
 
-We evaluated the algorithm using three public gene expression and methylation
+We evaluated the algorithm using two public gene expression and methylation
 datasets.  Most tests were performed on a desktop PC with a single AMD
@@
 
-\subsection{Small BicatYeast example}
+\subsection{Small EisenYeast example}
 
-The first example is a tiny proof of concept that provides the data for Figure
+The first example is a small proof of concept that provides the data for Figure
 \ref{fig1} and establishes our timing and memory use measurement methodology.
@@
 compare its output with brute force results. The example uses the {\tt
-BicatYeast} data from the {\tt biclust} R package and related {\tt BicAT}
-software\cite{biclust, bicat}. The data consist of 70 experiments involving the
-Saccharomyces Cerevisiae yeast measuring gene expression data across 419
-Affymetrix probes, represented as a $70\times 419$ real-valued matrix.  The
-example finds all pairs of columns (Affymetrix probes) that meet or exceed a
-Pearson correlation value of 0.95.  We use a projected dimension $p=10$ in {\tt tcor()}.
+EisenYeast} data from the {\tt biclust} R package\cite{biclust}.
+The data consist of 80 experiments involving the
+Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
+Affymetrix probes, represented as a $80\times 6221$ real-valued matrix.  The
+example finds all pairs of columns (gene expression measurements) that meet or exceed a
+Pearson correlation value of 0.95.  We use a projected dimension $p=10$
+in {\tt tcor()}.
 \begin{small}
@@
 library(irlba)
-source("tcor.R")  # XXX Replace later with tcor pacakge
+source("tcor.R")  # XXX Replace later with tcor pacakge XXX XXX
 library(biclust)  # for the BicatYeast data
-data(BicatYeast)  # in gene by sample orientation
-A = t(BicatYeast) # to correlate genes to genes
+data(EisenYeast)  # in gene by sample orientation
+A = t(EisenYeast) # to correlate genes to genes
 threshold = 0.95
@@
 \end{small}
-Brute-force computation outperforms {\tt tcor()} in this tiny first example,
-both in speed and slightly in memory use.  Each algorithm identifies the same
-115 unique vector pairs that exceed the specified correlation threshold of
-0.95, with comparative timing and memory use shown in the following table.
+Brute-force computation outperforms {\tt tcor()} in this small first example in
+speed but consumes substantially more memory.  Each algorithm identifies the
+same set of 125 gene expression pairs that exceed the specified correlation
+threshold of 0.95, with comparative timing and memory use shown in the
+following table.
 
@@
 require(xtable)
-ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
+ans = data.frame(Method=c("tcor","brute force"), `Wall clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
 print(
     xtable(ans,
-        caption = "BicatYeast results, threshold=0.95", label = "BicatYeast", caption.placement = "top")
-  ,size="small", table.placement="ht", include.rownames=FALSE)
-@
-
-\subsection{TCGA Gene Expression Examples}
-
-Performance benefits from the {\tt tcor()} function become apparent with more
-data.  The next example uses RPKM-normalized gene expression data measured
-across 20,532 genes from 878 breast cancer tumor samples obtained from the
-Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}. See the supplemental
-code\cite{sup} (also embedded in this Sweave document) for examples of
-efficiently converting the downloaded sample expression data into an R matrix.
-
-Eight columns with constant zero values were removed from the data,
-leaving an $878\times 20524$ matrix. The example computes
-pairs of gene expression column vectors with correlation values of 0.99
-or more. We use the same {\tt tcor} projection dimension of 10 and performance
-measurement methodology as used in the first example.
-
-Both methods find the same 3,572 pairs of vectors. The fast brute-force
-computation is slightly faster than the {\tt tcor()} function largely due to
-the still relatively small input matrix and better utilization of optimized
-BLAS routines. But the {\tt tcor()} function uses far less memory.  (For
-comparison, R's {\tt cor()} function took about 372 seconds and about
-3200$\,$MB RAM on the same problem on the same test system.)
-
-<<echo=FALSE, results=tex>>=
-# --------------------------------------------------------------------------------------------------
-# The first time you run this, you need to download the data as outlined in the
-# following commented-out section. This code creates a file named
-# 'rpkm.RData' used subsequently.
-#
-# To run any of the remaining (large) examples, first set the system environment
-# variable RUN_EXAMPLES=TRUE (otherwise they won't run and representative results
-# from Bryan's test machine will be printed).
-#
-#library(doMC)
-# Download and extract this file with, for example:
-# wget  http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/BRCA/20150821/gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
-# tar zxf gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
-# cd gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0
-# R --slave < read.R   # (run this file)
-#f = "BRCA.rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.data.txt"
-#N = 20534  # total number of lines
-#h = 2      # header lines
-# Read the first header line with sample IDs:
-#id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
-# Read the 2nd header line with sample type (raw, RPKM, etc.)
-#header = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=FALSE, nrows=1))
-# Read the rest of the file in parallel. I prefer doMC/foreach over mclapply
-# for this because foreach can combine results incrementally.
-#cores = 4
-#registerDoMC(cores)
-#block = floor((N - h)/cores)
-#t1 = proc.time()
-#ans = foreach(j=1:cores, .combine=rbind) %dopar%
-#{
-#    skip = block * (j - 1) + h
-#    nrows = ifelse(j == cores, -1, block)
-#    read.table(f, sep="\t", stringsAsFactors=FALSE, skip=skip, header=FALSE, nrows=nrows)
-#}
-#names(ans) = header
-#print(proc.time() - t1)
-# (All timings shown for a quad-core Athlon home PC reading from an older
-# OCZ REVODRIVE3 X2 SSD drive (ca 2014).)
-#   user  system elapsed
-#494.549   4.134 131.232
-# Compare with sequential approach:
-#t1 = proc.time()
-#x = read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=TRUE)
-#print(proc.time() - t1)
-#   user  system elapsed
-#517.430   1.318 519.553
-# Create a samples x gene matrix of RPKM values
-#genes = ans[, 1] # gene labels
-#idx   = grep("RPKM", names(ans))    # Select columns with RPKM values
-#rpkm     = t(as.matrix(ans[, idx]))
-# Label the coordinates by sample ID and gene:
-#rownames(rpkm) = id[idx]
-#colnames(rpkm) = genes
-#save(rpkm, file="rpkm.RData")
-# --------------------------------------------------------------------------------------------------
-# Default values:
-t1=40.34
-t2=35.60
-m1=39.10
-m2=9595.10
-
-if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
-{
-load("rpkm.RData")  # rpkm expression matrix
-
-# The data have a few columns of all zeros that need to be removed
-n = ncol(rpkm)
-rpkm = rpkm[, colMeans(rpkm) > 0]
-
-threshold = 0.99     # correlation threshold
-xx  = gc()           # clear
-m1  = sum(gc()[,2])  # Memory in megabytes currently in use
-t1  = proc.time()
-tx  = tcor(rpkm, threshold, 11)
-t1  = (proc.time() - t1)[3] # Time in seconds
-m1  = sum(gc()[,6] - m1)    # Peak excess memory use during the test
-
-# Compare with the fastest way we know of to compute a thresholded correlation
-# matrix by brute force. This approach makes heavy use of fast multi-threaded
-# BLAS routines. This approach is much faster than R's default `cor` function.
-xx  = gc()
-m2  = sum(gc()[,2])  # Memory in megabytes currently in use
-t2 = proc.time()
-mu = colMeans(rpkm)
-s  = sqrt(apply(rpkm, 2, crossprod) - nrow(rpkm) * mu ^ 2) # col norms of centered matrix
-cs = scale(rpkm, center=mu, scale=s)  # scale and center the data
-bx = crossprod(cs)                    # full correlation matrix
-diag(bx) = 0                          # ignore diagonal (all ones)
-pairs = which(bx >= threshold, arr.ind=TRUE)
-t2  = (proc.time() - t2)[3] # Time in seconds
-m2  = sum(gc()[,6] - m2)    # Peak excess memory use during the test
-} # RUN_EXAMPLES=TRUE
-
-require(xtable)
-ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
-print(
-    xtable(ans,
-        caption = "TCGA BRCA gene expression results (878 samples x 20,542 genes, threshold=0.99)", label = "TCGA1", caption.placement = "top")
-  ,size="small", table.placement="ht", include.rownames=FALSE)
-@
-
-We simulated a data set with a larger number of samples by simply replicating
-the rows of the gene expression matrix 27 times to obtain a matrix with 23,706
-rows and 20,524 columns, noting that the simulated data have the same rank (of
-at most 878) and the same expected number of vector pairs meeting the
-correlation threshold as the original data. Each method again computed the same
-set of  3,572 pairs of vectors meeting the correlation threshold of 0.99, but
-the {\tt tcor()} function outperforms the brute force method in both compute
-time and memory consumption.
-
-<<echo=FALSE, results=tex>>=
-# Default values:
-t1=372.69
-t2=546.06
-m1=768.3
-m2=9600.2
-
-if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
-{
-  # simulated data, replicate rows 27x
-}
-require(xtable)
-ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
-print(
-    xtable(ans,
-        caption = "TCGA simulated gene expression results (23,706 rows x 20,542 columns, threshold=0.99)", label = "TCGA2", caption.placement = "top")
+        caption = "EisenYeast results, threshold=0.95", label = "EisenYeast", caption.placement = "top")
   ,size="small", table.placement="ht", include.rownames=FALSE)
@@
 
-The following example uses calculated beta values from the Illumina Human
-Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor samples
-obtained from the Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}. See the
-supplemental code\cite{sup} for examples of efficiently converting the
-downloaded sample methylation data into an R matrix.
+Performance benefits from the {\tt tcor()} function become apparent with more
+data.  The following example uses calculated beta values from the Illumina
+Human Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor
+samples obtained from the Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}.
+See the supplemental code\cite{sup} for examples of efficiently converting the
+downloaded sample methylation data into an R matrix. We chose these data for
+their public availability as a general example of computing thresholded
+correlation among sets of large numbers of vectors.
 
 We removed 91,563 columns with all constant or missing values from the data,
-leaving an $80\times 394014$ matrix. The example computes
-pairs of methylation beta value column vectors with correlation values of 0.99
-or more. We use the same {\tt tcor} projection dimension of 10 and performance
-measurement methodology as used in the first example.
+leaving an $80\times 394014$ matrix. The example computes pairs of methylation
+beta value column vectors with correlation values of 0.99 or more. We used a
+{\tt tcor} projection dimension of 10 and the same performance measurement
+methodology used in the first example. Despite its low rank and modest data
+size, the problem is too large to compute using a na\"ive brute force
+correlation method.
 
@@
 values of 0.99 or more in about three hours on our test PC. A brute force
-algorithm would need to evaluate over 155 billion vector pairs; the projection
-of Algorithm \ref{prune} limited the search space to only about 4 million
-candidate pairs, from which the 913,601 correlated pairs were found.
+algorithm would need to evaluate the threshold condition for more than 155
+billion vector pairs; the projection of Algorithm \ref{prune} limited the
+search space to only about 4 million candidate pairs, from which the 913,601
+correlated pairs were found.
 
@@
 contained two Intel Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per
-computer) and 64$\,$GB ECC registered RAM running at 1,600$\,$MHz running
-CentOS 6.6 and 64-bit R version 3.2.2 with the reference R BLAS library.
+computer) and 64$\,$GB/1,600$\,$MHz ECC registered RAM running CentOS 6.6 and
+64-bit R version 3.2.2 with the reference R BLAS library.
 
@@
 require(xtable)
-ans = data.frame(Method=c("tcor (4 core PC)", "tcor (64 core cluster)", ""), `Wall-clock time (s)`=c(t1, t3, t4), `Memory use (MB)`=c(m1, m3, m4), " "=c("","(local)","(remote)"), check.names=FALSE)
+ans = data.frame(Method=c("tcor (4 core PC)", "tcor (64 core cluster)", ""), `Wall clock time (s)`=c(t1, t3, t4), `Memory use (MB)`=c(m1, m3, m4), " "=c("","(local)","(remote)"), check.names=FALSE)
 print(
@@
 @
-From a set of almost 400 thousand vectors, Algorithm \ref{prune} finds 913,601
-pairs of vectors with correlation values of 0.99 or more on a modest Linux
-cluster in about 8 minutes. The same algorithm also runs well in constrained
-memory settings.
+From a set of almost 400 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
+pairs with correlation values of 0.99 or more on a modest Linux
+cluster in about 8 minutes. The same algorithm also runs, although slower, in
+the constrained memory setting of a desktop PC.
 
@@
 
-XXX
-
-1. The main ideas in Algorithm \ref{prune}, although applied to correlation,
-easily extend to any distance metric.
-
-2. Anti-correlated pairs? Anyone?
+The idea of evaluating threshold conditions by projection into low dimensional
+subspaces applies to distance metrics similarly to correlation. Lemma
+\ref{sumlemma} decomposes Euclidean distance between two vectors in terms of
+singular basis vectors and Algorithm \ref{prune} can be applied directly to the
+problem of finding all pairs of vectors that meet a thresholded Euclidean
+distance with small changes in steps 3, 4, and 7.  And, by equivalence of
+norms, thresholded Euclidean distance can be extended to other distance
+metrices similarly. The reference R implementation\cite{sup} includes functions
+for general thresholded distance computation.
 
@@
 
-\begin{thebibliography}{99}
+\bibliography{bibliography}{}
+\bibliographystyle{plain}
 
-%% Related use of SVD in other methods
-%\bibitem{spca}
-%Bair, Eric, et al. Prediction by supervised principal components. Journal of the American Statistical Association 101.473 (2006).
-
-%% irlba
-\bibitem{irlba}
-Baglama, James, and Lothar Reichel. Augmented implicitly restarted Lanczos bidiagonalization methods. SIAM Journal on Scientific Computing 27.1 (2005): 19-42.
-
-\bibitem{irlbam}
-Baglama, James,  {\url{http://www.math.uri.edu/~jbaglama/software/irlba.m}}.
-
-%% BicatYeast data
-\bibitem{bicat} Barkow, S., Bleuler, S., Prelic, A., Zimmermann, P., and E. Zitzler. BicAT: a biclustering analysis toolbox, Bioinformatics, 2006 22(10):1282-1283.
-
-%% BROAD GDAC dashboard
-\bibitem{gdac}{Broad Institute TCGA Genome Data Analysis Center (2015): Analysis-ready standardized TCGA data from Broad GDAC Firehose stddata\_\_2015\_06\_01 run.
-Broad Institute of MIT and Harvard.
-Dataset.
-\url{http://dx.doi.org/10.7908/C1251HBG}}
-
-%% Applications in machine learning
-\bibitem{ml1}
-Hall, Mark A. Correlation-based feature selection for machine learning. Diss. The University of Waikato, 1999.
-
-%% R's biclust pacakge
-\bibitem{biclust}
-Kaiser, Sebastian, Santamaria, Rodrigo, Khamiakova, Tatsiana, Sill, Martin,
-Theron, Roberto, Quintales, Luis, Leisch, Friederich, and De Troyer, Ewoud, 
-\url{https://cran.r-project.org/web/packages/biclust}.
-
-%% IRLBA implementations
-\bibitem{irlbap}
-Kane, Michael and Lewis, Bryan,  \url{https://github.com/bwlewis/irlbpy}.
-
-%% doredis
-\bibitem{doredis}
-Lewis, Bryan. The doRedis package for R. \url{https://cran.r-project.org/web/packages/doRedis}.
-
-%% IRLBA implementations and supplemental code
-\bibitem{irlbar}
-Lewis, Bryan. The irlba package for R.  \url{https://cran.r-project.org/web/packages/irlba}.
-
-\bibitem{sup}
-Lewis, Bryan. The tcor package for R.  \url{https://github.com/bwlewis/tcor}.
-
-%% Approximate correlation
-\bibitem{timeseries}
-Mueen, Abdullah, Suman Nath, and Jie Liu. Fast approximate correlation for massive time-series data. Proceedings of the 2010 ACM SIGMOD International Conference on Management of data. ACM, 2010.
-
-%% Applications in genomics
-\bibitem{genomics}
-Song, Lin, Peter Langfelder, and Steve Horvath. Comparison of co-expression measures: mutual information, correlation, and model based indices. BMC bioinformatics 13.1 (2012): 328.
-
-%% foreach
-\bibitem{foreach}
-Weston, Steve. The foreach package for R. \url{https://cran.r-project.org/web/packages/foreach/vignettes/foreach.pdf}.
-
-%% SVD similarity -- extremely similar to the approach in this paper!
-%% Applied to images
-\bibitem{svd-similarity}
-Wu, Daniel, et al. Efficient retrieval for browsing large image databases. Proceedings of the fifth international conference on Information and knowledge management. ACM, 1996.
-
-%% Other correlation pruning methods
-\bibitem{prune1}
-Xiong, Hui, et al. TAPER: A two-step approach for all-strong-pairs correlation query in large databases. Knowledge and Data Engineering, IEEE Transactions on 18.4 (2006): 493-508.
-\bibitem{prune2}
-Xiong, Hui, Mark Brodie, and Sheng Ma. Top-cop: Mining top-k strongly correlated pairs in large databases. Data Mining, 2006. ICDM'06. Sixth International Conference on. IEEE, 2006.
-
-%% Applications in machine learning
-\bibitem{ml2}
-Yu, Lei, and Huan Liu. Feature selection for high-dimensional data: A fast correlation-based filter solution. ICML. Vol. 3. 2003.
-
-%% Other correlation pruning methods
-\bibitem{prune3}
-Zhang, Jian, and Joan Feigenbaum.
-Finding highly correlated pairs efficiently with powerful pruning.
-Proceedings of the 15th ACM international conference on Information and knowledge management.
-ACM, 2006.
-
-\end{thebibliography}
 \end{document}

### 2015-12-17 threshold/draft.Rnw
Updated draft; about to rebuild examples

@@
 above a specified threshold is important to many practical applications.
-Real-world problems in genomics, machine learning, financial and other time
-series and other applications can produce correlation matrices too large to
-explicitly form and tractably compute. Often, only values corresponding to
-highly-correlated vectors are of interest, and those values typically make up a
-small fraction of the overall correlation matrix. We present a method based on
-the singular value decomposition (SVD) and its relationship to the data
-covariance structure that can efficiently compute thresholded subsets of very
-large correlation matrices.
+Real-world problems in genomics, machine learning, finance other applications
+can produce correlation matrices too large to explicitly form and tractably
+compute. Often, only values corresponding to highly-correlated vectors are of
+interest, and those values typically make up a small fraction of the overall
+correlation matrix. We present a method based on the singular value
+decomposition (SVD) and its relationship to the data covariance structure that
+can efficiently compute thresholded subsets of very large correlation matrices.
 
@@
 \end{lemma}
-where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
+where, here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
 vector consisting of $1$ in position $j$ and zeros otherwise.
@@
 difference matrix with $-1$ along the diagonal, $1$ along the $i$th
-super-diagonal, and zeros elsewhere.  Consider just adjacent points of $P v_1$.
-Then $D^1P v_1$ consists of the differences of adjacent entries in $P v_1$.
-Entries such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$ correspond to pairs of
-vectors that do not meet the correlation threshold, where here and below
-an exponent applied to a vector denotes element-wise exponentiation.
+super-diagonal, and zeros elsewhere.  Then $D^1P v_1$ consists of the
+differences of adjacent entries in $P v_1$.  Entries such that $\sigma_1^2 (D^1
+P v_1)^2 > 2(1-t)$ correspond to pairs of vectors that do not meet the
+correlation threshold, where here and below an exponent applied to a vector
+denotes element-wise exponentiation.
 
@@
 additional computational or storage overhead as follows. Assume $A$ is an
-$m\times n$ real-valued matrix without constant-valued columns. Let
+$m\times n$ real-valued matrix without any constant-valued columns. Let
 $z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
@@
 A full/na\"ive thresholded correlation computation requires $O(n^2 m)$ flops.
-Let $p$ be the selected IRLBA dimension. Then the proposed method requires
+Let $p$ be the selected IRLBA dimension. Then Algorithm \ref{prune} requires
 $O(n p \ell)$ flops, where $\ell$ is the longest run of ordered entries in
 $v_1$ that meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
-plus the flops required by IRLBA and the post-processing step 7 (also $O(np)$).
+including the flops required by IRLBA and final thresholding in step 7.  The
+main computational savings of $\frac{nm}{lp}$ depends on the rate of decay in
+the singular values of $A$, the subspace dimension $p$, and the desired
+threshold value. We remark that this estimate of computational savings is
+available already in step 3 of Algorithm \ref{prune} after only modest
+computational effort. Users might use this cheap estimate to, for example,
+alter their threshold in some cases.
 
@@
 version 0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
-OMP\_NUM\_THREADS=4}. The last experiment includes a test run on a Linux
+OMP\_NUM\_THREADS=1}. The last experiment includes a test run on a Linux
 cluster as described in the text.

### 2015-12-19 threshold/draft.Rnw
JSS style update

@@
-%
 %% Instructions for processing this document into a PDF document
 %%
-%% FIRST NOTE, you need the 'biclust' package installed in R.
+%% FIRST NOTE, you need the 'xtable' and 'biclust' packages installedi in R.
 %% Some of the examples will come from data sets in this package.
@@
 %% install.packages("biclust",repos="http://cran.case.edu")
+%% install.packages("xtable",repos="http://cran.case.edu")
 %%
@@
 %% LaTeX generally takes a few runs to get reference numbers right.
-%%
-\documentclass[10pt]{article}
+
+\documentclass[article]{jss}
+%% need no \usepackage{Sweave.sty}
 
@@
 \newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
 
@@
 University of Rhode Island
-\and
+\And
 Michael Kane\\
 Yale University
-\and
+\AND
 Bryan Lewis\\
 Paradigm4, Inc.\\
-\and
+\And
 Alex Poliakov\\
@@
 }
+\Plainauthor{James Baglama, Michael Kane, Bryan Lewis, Alex Poliakov}
+\Plaintitle{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
+\Shorttitle{Efficient Thresholded Correlation}
 
-\date{}
-
-\begin{document}
-\SweaveOpts{concordance=TRUE}
-\maketitle
-
-\begin{abstract}
-
+\Abstract{
 Efficiently computing a subset of a correlation matrix consisting of values
@@
 can efficiently compute thresholded subsets of very large correlation matrices.
+}
+\Keywords{thresholded correlation, SVD, \proglang{R}}
+\Plainkeywords{thresholded correlation, SVD, R}
+\Address{
+James Baglama, Chair of Mathematics\\
+Department of Mathematics\\
+University of Rhode Island\\
+Kingston, Rhode Island 02881\\
+E-mail: \email{jbaglama@math.uri.edu}\\
+URL: \url{http://www.math.uri.edu/~jbaglama}\\
+}
 
-\end{abstract}
+\begin{document}
 
@@
 of gene expression values may be used in the discovery of networks of genes
-relevant to particular biological processes\cite{genomics}.  Identification of
+relevant to particular biological processes \citep{genomics}.  Identification of
 highly-correlated pairs may be used in feature selection algorithms for machine
-learning applications\cite{ml1, ml2} and are important to time series and
-image processing applications\cite{timeseries, svd-similarity}.
+learning applications \citep{ml1, ml2} and are important to time series and
+image processing applications \citep{timeseries, svd-similarity}.
 
@@
 pruning rules based on the frequency of mutually co-occurring observations (rows)
-among vector pairs; see \cite{prune1, prune2} and the references therein.
+among vector pairs; see \citep{prune1, prune2} and the references therein.
 Related methods approximate a set containing the most highly-correlated
-pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
+pairs using hashing algorithms \citep{prune3}. A number of other methods arrive
 at approximate sets of the most highly-correlated pairs using
 dimensionality-reduction techniques like the discrete Fourier
-transform\cite{timeseries}. The method of Wu, et al.\cite{svd-similarity},
+transform \citep{timeseries}. The method of Wu, et al. \citep{svd-similarity},
 conceptually very similar to our approach, uses the singular value
@@
 Fortunately the $\ell$ evaluations in step 4 are independent of each other and
-can easily be computed in parallel. See the reference R
-implementation\cite{sup} for an example.
+can easily be computed in parallel. See the reference \proglang{R}
+implementation \citep{sup} for an example.
 
@@
 The augmented implicitly restarted Lanczos bidiagonalization algorithm\break(IRLBA)
-of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value
+of Baglama and Reichel \citep{irlba} efficiently computes truncated singular value
 decompositions of large dense or sparse matrices. Reference implementations are
-available for R\cite{irlbar}, Matlab\cite{irlbam}, and Python\cite{irlbap}.
+available for \proglang{R} \citep{irlbar}, \proglang{Matlab} \citep{irlbam},
+and \proglang{Python} \citep{irlbap}.
 
@@
 at most computing two additional vector inner products per matrix-vector product.
-The R implementation\cite{irlbar} of IRLBA includes arguments for centering and
+The \proglang{R} implementation \citep{irlbar} of IRLBA includes arguments for centering and
 scaling the input matrix.
 
-A full/na\"ive thresholded correlation computation requires $O(n^2 m)$ flops.
+A na\"ive brute force thresholded correlation computation requires $O(n^2 m)$ flops.
 Let $p$ be the selected IRLBA dimension. Then Algorithm \ref{prune} requires
@@
 
-We evaluated the algorithm using two public gene expression and methylation
-datasets.  Most tests were performed on a desktop PC with a single AMD
-A10-7850K 3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered
-RAM running the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R
-version 3.1.2 using double-precision arithmetic and the OpenBLAS library
-version 0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
-OMP\_NUM\_THREADS=1}. The last experiment includes a test run on a Linux
-cluster as described in the text.
+We evaluated the algorithm using public gene expression and DNA methylation
+datasets from the \proglang{R} {\tt biclust} \citep{biclust} package and from
+the Cancer Genome Atlas \citep{gdac}, referred to as TCGA.  Most tests were
+performed on a desktop PC with a single AMD A10-7850K 3.7$\,$GHz Athlon
+quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered RAM running the Ubuntu 15.10
+GNU/Linux OS. Tests were performed with 64-bit \proglang{R} version 3.2.3 using
+double-precision arithmetic and the OpenBLAS library version 0.2.14-1ubuntu1
+(based on Goto's BLAS version 1.13) with {\tt OMP\_NUM\_THREADS=1}. The last
+experiment includes a test run on a Linux cluster described in that section.
 
-All examples use the reference R implementation, {\tt tcor()}, which can be
-installed for the R language from the development GitHub repository using the
+All examples use the reference \proglang{R} implementation, {\tt tcor()}, which can be
+installed from the development GitHub repository using the
 {\tt devtools} package with:
@@
 
-The native R {\tt cor()} function is a general function that can compute
+The native \proglang{R} {\tt cor()} function is a general function that can compute
 several different types of correlation matrices including Pearson, Spearman,
@@
 
-\subsection{Small EisenYeast example}
+\subsection{Small gene expression example}
 
@@
 compare its output with brute force results. The example uses the {\tt
-EisenYeast} data from the {\tt biclust} R package\cite{biclust}.
+EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
 The data consist of 80 experiments involving the
@@
 <<eval=TRUE,echo=TRUE>>=
-library(irlba)
-source("tcor.R")  # XXX Replace later with tcor pacakge XXX XXX
+library(tcor)
+library(doMC)     # local parallel computation
 library(biclust)  # for the BicatYeast data
@@
 threshold = 0.95
+
+# Register a four-core multicore parallel backend
+registerDoMC(4)
 
@@
 Human Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor
-samples obtained from the Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}.
-See the supplemental code\cite{sup} for examples of efficiently converting the
-downloaded sample methylation data into an R matrix. We chose these data for
-their public availability as a general example of computing thresholded
-correlation among sets of large numbers of vectors.
+samples obtained from the Broad GDAC Cancer Genome Atlas dashboard \citep{gdac}.
+See the \proglang{R} package implementation \citep{sup} for examples of
+efficiently converting the downloaded sample methylation data into an R matrix.
+We chose these data for their public availability as a general example of
+computing thresholded correlation among sets of large numbers of vectors,
+rather than for any specific biological meaning.
 
@@
 correlation method.
+
+Our simple brute force implementation was not possible due to memory constraints.
+We remark that the matrix multiplication central to the brute force
+implementation can be decomposed into a sequence of smaller steps to
+fit within given memory constraints. However, the brute force algorithm would still
+need to evaluate all of the nearly 160 billion possible vector pair correlation
+values.
 
@@
 computer) and 64$\,$GB/1,600$\,$MHz ECC registered RAM running CentOS 6.6 and
-64-bit R version 3.2.2 with the reference R BLAS library.
+64-bit \proglang{R} version 3.2.2 with the reference \proglang{R} BLAS library.
 
-No code change was required to run in parallel on the cluster. The reference R
-{\tt tcor()} implementation uses R's {\tt foreach}\cite{foreach} framework to
+No code change was required to run in parallel on the cluster. The reference \proglang{R}
+{\tt tcor()} implementation uses \proglang{R}'s {\tt foreach} \citep{foreach} framework to
 run on arbitrary parallel configurations (``back ends''). We simply registered
-a {\tt doRedis}\cite{doredis} parallel back end with 16 R workers per computer
-(64 total R worker processes) prior to invoking {\tt tcor()}.
+a {\tt doRedis} \citep{doredis} parallel back end with 16 \proglang{R} workers per computer
+(64 total \proglang{R} worker processes) prior to invoking {\tt tcor()}.
 
@@
 below. The cluster example computed the filtering step 7 of Algorithm
-\ref{prune} in parallel across the 64 remote R processes. That approach incurs,
+\ref{prune} in parallel across the 64 remote \proglang{R} processes. That approach incurs,
 in this example, an additional memory overhead of about 16$\,$GB because each
-remote R process requires access to a copy of the data matrix $A$. More thrifty
+remote \proglang{R} process requires access to a copy of the data matrix $A$. More thrifty
 adaptations of the algorithm are possible and could, for example, share the
-memory required by $A$ between worker R processes on each computer.
+memory required by $A$ between worker \proglang{R} processes on each computer.
 
 <<echo=FALSE, results=tex>>=
-# Default values:
-t1=10449.670
+# Test system values:
+t1=10309
 t3=482
 t4=NA
-m1=1261.6
+m1=389.1
 m3=994.3
@@
 {
- # methylation data
+  library(doMC)
+  system("wget http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/ACC/20150821/gdac.broadinstitute.org_ACC.Merge_methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.Level_3.2015082100.0.0.tar.gz")
+  cmd = "cat ACC.methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.data.txt  | tail -n +%s | cut -d '       ' -f 2,6,10,14,18,22,26,30,34,38,42,46,50,54,58,62,66,70,74,78,82,86,90,94,98,102,106,110,114,118,122,126,130,134,138,142,146,150,154,158,162,166,170,174,178,182,186,190,194,198,202,206,210,214,218,222,226,230,234,238,242,246,250,254,258,262,266,270,274,278,282,286,290,294,298,302,306,310,314,318"
+  N = 485579  # total number of lines
+  h = 2       # total header lines
+  # Read the first header line with sample IDs:
+  f = pipe(cmd, open="r")
+  id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
+  close(f)
+  # Read the rest of the file in parallel. I prefer doMC/foreach over mclapply for this because foreach can combine results incrementally.
+  registerDoMC(4)  # four cores on our test machine
+  block = floor((N - h)/cores)
+  acc = foreach(j=1:cores, .combine=rbind) %dopar%
+  {
+    skip = block * (j - 1) + h + 1
+    nrows = ifelse(j == cores, -1, block)
+    f = pipe(sprintf(cmd, skip), open="r")
+    on.exit(close(f))
+    read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=nrows)
   }
+  acc = t(acc)
+  # OK, the example data are parsed and in acc. proceed as in the smaller example.
+  threshold = 0.99
+  # Compute using tcor():
+  m1  = sum(gc()[,2])  # Memory in megabytes currently in use
+  t1  = proc.time()
+  tx  = tcor(acc, t=threshold, p=10)
+  t1  = (proc.time() - t1)[3] # Time in seconds
+  m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
+
+  # The cluster example proceeds similarly, but registers a doRedis backend
+  # instead of the doMC backend before invoking tcor...
+} # end if run examples
 require(xtable)
@@
 @
-From a set of almost 400 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
+From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
 pairs with correlation values of 0.99 or more on a modest Linux
-cluster in about 8 minutes. The same algorithm also runs, although slower, in
-the constrained memory setting of a desktop PC.
-
-%%<<eval=TRUE,echo=FALSE>>=
-%%library(irlba)
-%%
-%%print("R code here")
-%%@
+cluster in about 8 minutes. The same algorithm works, more slowly, in
+the constrained memory setting of a desktop PC without modification.
 
@@
 problem of finding all pairs of vectors that meet a thresholded Euclidean
-distance with small changes in steps 3, 4, and 7.  And, by equivalence of
-norms, thresholded Euclidean distance can be extended to other distance
-metrices similarly. The reference R implementation\cite{sup} includes functions
-for general thresholded distance computation.
+distance with small changes in steps 3, 4, and 7.  By equivalence of norms,
+thresholded Euclidean distance can be extended to other distance metrics. The
+reference \proglang{R} implementation \citep{sup} includes functions for thresholded
+distance and correlation.
+
+The ideas presented here can be adapted to efficiently find the top N most
+correlated vector pairs for a given constant N similarly to the work of
+Xiong et al. \citep{prune2}.
 
@@
 
+We present a simple algorithm and reference \proglang{R} implementation based on a
+truncated singular value decomposition (SVD) that efficiently computes
+thresholded subsets of large correlation and distance matrices.  We show that
+the natural connection between data covariance and the SVD can result in
+substantial computational and memory savings, and illustrate these savings with
+a few genomics data experiments.  Moreover, the main computational work in the
+algorithm is easily parallelized and easily deployed across computational
+clusters.
+
 \bibliography{bibliography}{}
-\bibliographystyle{plain}
 

### 2015-12-19 threshold/draft.Rnw
Fixed numerous typos

@@
 matrix with nonzero eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$.
-The span of the columns of $V$ form an orthonormal basis of the range of the
+The columns of $V$ form an orthonormal basis of the range of the
 correlation matrix $A^TA$.
@@
 chosen in practice such that $p<<k$. Setting $p=10$ for the example shown in
-Figure \ref{fig1} prunes all but $431$ possible vector pairs out of a total
-$419(419-1)/2 = 87,571$ with $\ell=138$.
+Figure \ref{fig1} prunes all but $12,924$ possible vector pairs out of a total
+$6221(6221 - 1) = 193,47,310$ with $\ell=1,715$.
 
@@
 $z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
-of the $m\times n$ matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
+of $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
 be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
@@
 matrix computation computed in the fastest way we can, directly using matrix
-multiplication instead of using the slower {\tt cor()} function.
+multiplication instead of using the more general {\tt cor()} function.
 
@@
 Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
-Affymetrix probes, represented as a $80\times 6221$ real-valued matrix.  The
+Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
 example finds all pairs of columns (gene expression measurements) that meet or exceed a
@@
     xtable(ans,
-        caption = "EisenYeast results, threshold=0.95", label = "EisenYeast", caption.placement = "top")
+        caption = "Gene expression results, threshold=0.95", label = "EisenYeast", caption.placement = "top")
   ,size="small", table.placement="ht", include.rownames=FALSE)
@@
 We chose these data for their public availability as a general example of
-computing thresholded correlation among sets of large numbers of vectors,
+computing thresholded correlation among large numbers of vectors,
 rather than for any specific biological meaning.
@@
 size, the problem is too large to compute using a na\"ive brute force
-correlation method.
-
-Our simple brute force implementation was not possible due to memory constraints.
-We remark that the matrix multiplication central to the brute force
-implementation can be decomposed into a sequence of smaller steps to
-fit within given memory constraints. However, the brute force algorithm would still
-need to evaluate all of the nearly 160 billion possible vector pair correlation
-values.
+correlation method.  (We remark that the matrix multiplication central to the
+brute force implementation can be decomposed into a sequence of smaller steps
+to fit within given memory constraints. However, the brute force algorithm
+would still need to evaluate all of the nearly 160 billion possible vector pair
+correlation values.)
 
@@
 a few genomics data experiments.  Moreover, the main computational work in the
-algorithm is easily parallelized and easily deployed across computational
+algorithm is easily parallelized and deployed across computational
 clusters.

### 2015-12-20 threshold/draft.Rnw
Addresses

@@
 URL: \url{http://www.math.uri.edu/~jbaglama}\\
+\\
+Michael Kane\\
+Yale University\\
+300 Geroge Street\\
+New Haven, Connecticut 06511\\
+E-mail: \email{michael.kane@yale.edu}\\
+\\
+Bryan Lewis\\
+Paradigm4, Inc.\\
+281 Winter Street, Suite 360\\
+Waltham, MA 02451\\
+E-mail: \email{blewis@illposed.net}\\
+URL: \url{http://illposed.net}\\
+\\
+Alex Poliakov\\
+Paradigm4, Inc.\\
+281 Winter Street, Suite 360\\
+Waltham, MA 02451\\
+E-mail: \email{apoliakov@paradigm4.com}\\
 }

### 2015-12-20 threshold/draft.tex
Added raw tex/image files for jss submission
old R files/example moving to package

--- 2015-12-20 threshold/draft.Rnw
@@
 \begin{center}
-<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
-library(biclust)
-data(EisenYeast)
-A = t(EisenYeast)
-A = sweep(A, 2, apply(A,2,mean), FUN=`-`)
-A = sweep(A, 2, sqrt(apply(A,2,crossprod)), FUN=`/`)
-s = svd(A)
-i = order(s$v[,1])
-plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
-t=0.99
-abline(h=c(0, sqrt(2*(1-t))/s$d[1]), col=2, lwd=1)
-@
+\includegraphics{draft-001}
 \caption{
@@
 
-<<eval=FALSE,echo=TRUE>>=
-devtools::install_github("bwlewis/tcor")
-@
+\begin{Schunk}
+\begin{Sinput}
+> devtools::install_github("bwlewis/tcor")
+\end{Sinput}
+\end{Schunk}
 
@@
 \begin{small}
-<<eval=TRUE,echo=TRUE>>=
-library(tcor)
-library(doMC)     # local parallel computation
-library(biclust)  # for the BicatYeast data
-data(EisenYeast)  # in gene by sample orientation
-A = t(EisenYeast) # to correlate genes to genes
-threshold = 0.95
-
-# Register a four-core multicore parallel backend
-registerDoMC(4)
-
-# Compute using tcor():
-m1  = sum(gc()[,2])  # Memory in megabytes currently in use
-t1  = proc.time()
-tx  = tcor(A, t=threshold, p=10)
-t1  = (proc.time() - t1)[3] # Time in seconds
-m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
-
-# Fast brute force full correlation:
-m2  = sum(gc()[,2])
-t2  = proc.time()
-mu = colMeans(A)
-s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
-cs = scale(A, center=mu, scale=s)  # scale and center the data
-cx = crossprod(cs)                 # full correlation matrix
-cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
-pairs = which(cx >= threshold, arr.ind=TRUE)
-t2  = (proc.time() - t2)[3]
-m2  = sum(gc()[,6]) - m2
-@
+\begin{Schunk}
+\begin{Sinput}
+> library(tcor)
+> library(doMC)     # local parallel computation
+> library(biclust)  # for the BicatYeast data
+> data(EisenYeast)  # in gene by sample orientation
+> A = t(EisenYeast) # to correlate genes to genes
+> threshold = 0.95
+> # Register a four-core multicore parallel backend
+> registerDoMC(4)
+> # Compute using tcor():
+> m1  = sum(gc()[,2])  # Memory in megabytes currently in use
+> t1  = proc.time()
+> tx  = tcor(A, t=threshold, p=10)
+> t1  = (proc.time() - t1)[3] # Time in seconds
+> m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
+> # Fast brute force full correlation:
+> m2  = sum(gc()[,2])
+> t2  = proc.time()
+> mu = colMeans(A)
+> s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
+> cs = scale(A, center=mu, scale=s)  # scale and center the data
+> cx = crossprod(cs)                 # full correlation matrix
+> cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
+> pairs = which(cx >= threshold, arr.ind=TRUE)
+> t2  = (proc.time() - t2)[3]
+> m2  = sum(gc()[,6]) - m2
+\end{Sinput}
+\end{Schunk}
 \end{small}
@@
 
-<<echo=FALSE, results=tex>>=
-require(xtable)
-ans = data.frame(Method=c("tcor","brute force"), `Wall clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
-print(
-    xtable(ans,
-        caption = "Gene expression results, threshold=0.95", label = "EisenYeast", caption.placement = "top")
-  ,size="small", table.placement="ht", include.rownames=FALSE)
-@
+% latex table generated in R 3.2.3 by xtable 1.8-0 package
+% Sun Dec 20 09:52:31 2015
+\begin{table}[ht]
+\centering
+{\small
+\begin{tabular}{lrr}
+  \hline
+Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
+  \hline
+tcor & 3.19 & 59.60 \\ 
+  brute force & 1.74 & 1075.10 \\ 
+   \hline
+\end{tabular}
+}
+\caption{Gene expression results, threshold=0.95} 
+\label{EisenYeast}
+\end{table}
 
@@
 
-<<echo=FALSE, results=tex>>=
-# Test system values:
-t1=10309
-t3=482
-t4=NA
-m1=389.1
-m3=994.3
-m4=42282.4
-
-if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
-{
-  library(doMC)
-  system("wget http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/ACC/20150821/gdac.broadinstitute.org_ACC.Merge_methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.Level_3.2015082100.0.0.tar.gz")
-  cmd = "cat ACC.methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.data.txt  | tail -n +%s | cut -d '       ' -f 2,6,10,14,18,22,26,30,34,38,42,46,50,54,58,62,66,70,74,78,82,86,90,94,98,102,106,110,114,118,122,126,130,134,138,142,146,150,154,158,162,166,170,174,178,182,186,190,194,198,202,206,210,214,218,222,226,230,234,238,242,246,250,254,258,262,266,270,274,278,282,286,290,294,298,302,306,310,314,318"
-  N = 485579  # total number of lines
-  h = 2       # total header lines
-  # Read the first header line with sample IDs:
-  f = pipe(cmd, open="r")
-  id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
-  close(f)
-  # Read the rest of the file in parallel. I prefer doMC/foreach over mclapply for this because foreach can combine results incrementally.
-  registerDoMC(4)  # four cores on our test machine
-  block = floor((N - h)/cores)
-  acc = foreach(j=1:cores, .combine=rbind) %dopar%
-  {
-    skip = block * (j - 1) + h + 1
-    nrows = ifelse(j == cores, -1, block)
-    f = pipe(sprintf(cmd, skip), open="r")
-    on.exit(close(f))
-    read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=nrows)
+% latex table generated in R 3.2.3 by xtable 1.8-0 package
+% Sun Dec 20 09:52:31 2015
+\begin{table}[ht]
+\centering
+{\small
+\begin{tabular}{lrrl}
+  \hline
+Method & Wall clock time (s) & Memory use (MB) &   \\ 
+  \hline
+tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
+  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
+   &  & 42282.40 & (remote) \\ 
+   \hline
+\end{tabular}
 }
-  acc = t(acc)
-  # OK, the example data are parsed and in acc. proceed as in the smaller example.
-  threshold = 0.99
-  # Compute using tcor():
-  m1  = sum(gc()[,2])  # Memory in megabytes currently in use
-  t1  = proc.time()
-  tx  = tcor(acc, t=threshold, p=10)
-  t1  = (proc.time() - t1)[3] # Time in seconds
-  m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
-
-  # The cluster example proceeds similarly, but registers a doRedis backend
-  # instead of the doMC backend before invoking tcor...
-} # end if run examples
-require(xtable)
-ans = data.frame(Method=c("tcor (4 core PC)", "tcor (64 core cluster)", ""), `Wall clock time (s)`=c(t1, t3, t4), `Memory use (MB)`=c(m1, m3, m4), " "=c("","(local)","(remote)"), check.names=FALSE)
-print(
-    xtable(ans,
-        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99).
-Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).", label = "TCGA3", caption.placement = "top")
-  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="")
-@
-From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
+\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99).
+Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).}
+\label{TCGA3}
+\end{table}From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
 pairs with correlation values of 0.99 or more on a modest Linux

### 2015-12-20 threshold/example.R

@@ -1,33 +1,0 @@
-source("tcor.R")
-
-library(biclust)  # for the EisenYeast and BicatYeast data
-data(BicatYeast)  # in gene by sample orientation (a tiny example for testing)
-A = t(BicatYeast) # to correlate genes to genes
-threshold = 0.95
-
-
-# Compute using tcor():
-m1  = sum(gc()[,2])  # Memory in megabytes currently in use
-t1  = proc.time()
-tx  = tcor(A, t=threshold, p=10)
-t1  = (proc.time() - t1)[3] # Time in seconds
-m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
-
-cat("tcor time (s)                      ", t1, "\n")
-cat("tcor peak excess memory (MB)       ", m1, "\n")
-
-# Compute using R's native cor() function:
-m2  = sum(gc()[,2])
-t2  = proc.time()
-cx  = cor(A)                # full correlation matrix
-diag(cx) = -Inf
-pairs = which(cx >= threshold, arr.ind=TRUE)
-t2  = (proc.time() - t2)[3]
-m2  = sum(gc()[,6]) - m2
-
-cat("brute force time (s)               ", t2, "\n")
-cat("brute force peak excess memory (MB)", m2, "\n")
-
-# Compare results!
-cat("Number of vector pairs meeting threshold found by brute force", nrow(pairs)/2, "\n")
-cat("Number of vector pairs meeting threshold found by tcor       ", nrow(tx[[1]]), "\n")

### 2015-12-20 threshold/tcor.R

@@ -1,133 +1,0 @@
-require(irlba)
-require(foreach)
-
-#' linear time longest run search (A. Poliakov), find the longest
-#' run of values in the vector within the specified distance
-#' @param v a vector with entries ordered in increasing order
-#' @param limit distance interval
-#' @return run length
-longrun = function(v, limit)
-{
-  lower = 1
-  ell = 1
-  for(upper in 2:length(v))
-  {
-    if(v[upper] - v[lower] <= limit)
-    {
-      ell = max(ell, upper - lower + 1)
-    } else
-    {
-      while(lower < upper && v[upper] - v[lower] > limit) lower = lower + 1
-    }
-  }
-  ell
-}
-
-#' Compute the thresholded correlations between columns of a matrix.
-#'
-#' Increase p to cut down the total number of candidate pairs evaluated,
-#' at the expense of costlier truncated SVDs.
-#'
-#' @param A an m by n real-valued dense or sparse matrix
-#' @param t a threshold value for correlation
-#' @param p projected subspace dimension
-#' @param filter "local" filters candidate set sequentially,
-#'  "distributed" computes thresholded correlations in a parallel code section which can be
-#'  faster but requires that the data matrix is available (see notes).
-#' @param ... additional arguments passed to \code{\link{irlba}}
-#'
-#' @return A list with three elements:
-#' \enumerate{
-#'   \item \code{indices} A three-column matrix such that the  first two columns contain
-#'         indices of vectors meeting the correlation threshold \code{t},
-#'         and the third column contains the corresponding correlation value.
-#'   \item \code{longest_run} The largest number of successive entries in the
-#'     ordered first singular vector within a projected distance defined by the
-#'     correlation threshold.
-#'   \item \code{n} The total number of _possible_ vectors that meet
-#'     the correlation threshold identified by the algorithm.
-#'   \item \code{total_time} Total run time.
-#' }
-tcor = function(A, t=0.99, p=10, filter=c("distributed", "local"), ...)
-{
-  filter = match.arg(filter)
-  if(ncol(A) < p) p = max(1, floor(ncol(A) / 2 - 1))
-  t0 = proc.time()
[…]

### 2015-12-22 threshold/draft.Rnw
Updated draft

@@
 \end{algorithm}
-Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
-decrease in proportion to variability in the data, it's often reasonable to
-judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
-projection $\tilde{i}=1$ in step 4 of the algorithm.
+One can cheaply estimate whether there are ``too many'' candidate pairs for a
+given dimension $p$ in step 5 by evaluating step 4 for only adjacent pairs of
+vectors, $\tilde{i}=1$.
 
@@
 
-The native \proglang{R} {\tt cor()} function is a general function that can compute
-several different types of correlation matrices including Pearson, Spearman,
-and Kendall and includes a number of options for dealing with missing values.
-Because of its flexibility, the native {\tt cor()} function does not always
-compute Pearson correlation matrices in the most computationally-efficient way.
+The native \proglang{R} {\tt cor()} function is a general function that can
+compute several different types of correlation matrices including Pearson,
+Spearman, and Kendall and includes a number of options for dealing with missing
+values.  Because of its flexibility, the native {\tt cor()} function does not
+always compute Pearson correlation matrices in the most
+computationally-efficient way.
 
@@
 example finds all pairs of columns (gene expression measurements) that meet or exceed a
-Pearson correlation value of 0.95.  We use a projected dimension $p=10$
-in {\tt tcor()}.
+Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
+the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
+generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
+to $10$, the number of candidate vector pairs drops to less than 13,000.
+The reference \proglang{R} implementation includes options for
+tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
+inexpensively extend an existing truncated SVD. The following example uses $p=10$.
 \begin{small}
 <<eval=TRUE,echo=TRUE>>=
+threshold = 0.95
+
+# Test system values:
+t1=3.13
+t2=1.75
+m1=59.60
+m2=1075.10
+
+if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
+{
 library(tcor)
@@
 A = t(EisenYeast) # to correlate genes to genes
-threshold = 0.95
-
 # Register a four-core multicore parallel backend
@@
 m2  = sum(gc()[,6]) - m2
+}
 @
@@
   acc = t(acc)
+
+  # remove columns containing missing values, all zero
+  n = ncol(acc)
+  acc = acc[, colMeans(acc) > 0]
+  cat("Removed", n - ncol(acc), "columns of all zeros.\n")
+  n = ncol(acc)
+  acc = acc[, apply(acc,2,function(x) !any(is.na(x)))]
+  cat("Removed", n - ncol(acc), "columns with missing values.\n")
+  cat("Data matrix size ", nrow(acc), "x", ncol(acc), "\n")
+
   # OK, the example data are parsed and in acc. proceed as in the smaller example.

### 2015-12-22 threshold/draft.tex
Updated draft

@@
 example finds all pairs of columns (gene expression measurements) that meet or exceed a
-Pearson correlation value of 0.95.  We use a projected dimension $p=10$
-in {\tt tcor()}.
+Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
+the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
+generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
+to $10$, the number of candidate vector pairs drops to less than 13,000.
+The reference \proglang{R} implementation includes options for
+tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
+inexpensively extend an existing truncated SVD. The following example uses $p=10$.
 \begin{small}
@@
 \begin{Sinput}
-> library(tcor)
-> library(doMC)     # local parallel computation
-> library(biclust)  # for the BicatYeast data
-> data(EisenYeast)  # in gene by sample orientation
-> A = t(EisenYeast) # to correlate genes to genes
 > threshold = 0.95
-> # Register a four-core multicore parallel backend
-> registerDoMC(4)
-> # Compute using tcor():
-> m1  = sum(gc()[,2])  # Memory in megabytes currently in use
-> t1  = proc.time()
-> tx  = tcor(A, t=threshold, p=10)
-> t1  = (proc.time() - t1)[3] # Time in seconds
-> m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
-> # Fast brute force full correlation:
-> m2  = sum(gc()[,2])
-> t2  = proc.time()
-> mu = colMeans(A)
-> s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
-> cs = scale(A, center=mu, scale=s)  # scale and center the data
-> cx = crossprod(cs)                 # full correlation matrix
-> cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
-> pairs = which(cx >= threshold, arr.ind=TRUE)
-> t2  = (proc.time() - t2)[3]
-> m2  = sum(gc()[,6]) - m2
+> # Test system values:
+> t1=3.13
+> t2=1.75
+> m1=59.60
+> m2=1075.10
+> if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
++ {
++ library(tcor)
++ library(doMC)     # local parallel computation
++ library(biclust)  # for the BicatYeast data
++ data(EisenYeast)  # in gene by sample orientation
++ A = t(EisenYeast) # to correlate genes to genes
++ # Register a four-core multicore parallel backend
++ registerDoMC(4)
++ 
++ # Compute using tcor():
++ m1  = sum(gc()[,2])  # Memory in megabytes currently in use
++ t1  = proc.time()
++ tx  = tcor(A, t=threshold, p=10)
++ t1  = (proc.time() - t1)[3] # Time in seconds
++ m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
++ 
++ # Fast brute force full correlation:
++ m2  = sum(gc()[,2])
++ t2  = proc.time()
++ mu = colMeans(A)
++ s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
++ cs = scale(A, center=mu, scale=s)  # scale and center the data
++ cx = crossprod(cs)                 # full correlation matrix
++ cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
++ pairs = which(cx >= threshold, arr.ind=TRUE)
++ t2  = (proc.time() - t2)[3]
++ m2  = sum(gc()[,6]) - m2
++ }
 \end{Sinput}
@@
 % latex table generated in R 3.2.3 by xtable 1.8-0 package
-% Sun Dec 20 09:52:31 2015
+% Mon Dec 21 19:50:10 2015
 \begin{table}[ht]
@@
   \hline
-tcor & 3.19 & 59.60 \\ 
-  brute force & 1.74 & 1075.10 \\ 
+tcor & 3.13 & 59.60 \\ 
+  brute force & 1.75 & 1075.10 \\ 
    \hline

### 2015-12-22 threshold/draft.Rnw
Minor changes to conclusion

@@
 The idea of evaluating threshold conditions by projection into low dimensional
-subspaces applies to distance metrics similarly to correlation. Lemma
-\ref{sumlemma} decomposes Euclidean distance between two vectors in terms of
-singular basis vectors and Algorithm \ref{prune} can be applied directly to the
-problem of finding all pairs of vectors that meet a thresholded Euclidean
-distance with small changes in steps 3, 4, and 7.  By equivalence of norms,
-thresholded Euclidean distance can be extended to other distance metrics. The
-reference \proglang{R} implementation \citep{sup} includes functions for thresholded
-distance and correlation.
+subspaces applies to distance metrics similarly to correlation, see
+\citep{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
+Euclidean distance between two vectors in terms of singular basis vectors and
+Algorithm \ref{prune} can be applied directly to the problem of finding all
+pairs of vectors that meet a thresholded Euclidean distance with small changes
+in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
+can be extended to other distance metrics. The reference \proglang{R}
+implementation \citep{sup} includes functions for thresholded distance and
+correlation.
 
@@
 
-We present a simple algorithm and reference \proglang{R} implementation based on a
-truncated singular value decomposition (SVD) that efficiently computes
-thresholded subsets of large correlation and distance matrices.  We show that
-the natural connection between data covariance and the SVD can result in
-substantial computational and memory savings, and illustrate these savings with
-a few genomics data experiments.  Moreover, the main computational work in the
-algorithm is easily parallelized and deployed across computational
-clusters.
+We present a simple algorithm and reference \proglang{R} implementation based
+on a truncated singular value decomposition (SVD) that efficiently computes
+thresholded subsets of large correlation  matrices.  We show that the natural
+connection between data covariance and the SVD can result in substantial
+computational and memory savings, and illustrate these savings with a few
+genomics data experiments.  The main computational work in the algorithm is
+easily parallelized and deployed across computational clusters.
 

### 2015-12-22 threshold/draft.tex
Minor changes to conclusion

@@
 % latex table generated in R 3.2.3 by xtable 1.8-0 package
-% Mon Dec 21 19:50:10 2015
+% Tue Dec 22 09:08:06 2015
 \begin{table}[ht]

### 2015-12-22 threshold/arxiv.tex
Added arxiv version

--- 2015-12-22 threshold/draft.tex
@@
-%% Instructions for processing this document into a PDF document
-%%
-%% FIRST NOTE, you need the 'xtable' and 'biclust' packages installedi in R.
-%% Some of the examples will come from data sets in this package.
-%% You can install it from the R console with, for example:
-%%
-%% install.packages("biclust",repos="http://cran.case.edu")
-%% install.packages("xtable",repos="http://cran.case.edu")
-%%
-%% Now you can prepare the PDF of the paper with any of:
-%%
-%% From a shell:
-%% R CMD Sweave --pdf draft.Rnw
-%%
-%% From an R prompt:
-%% Sweave("draft.Rnw")
-%% tools::texi2pdf("draft.tex")
-%%
-%% From RStudio:
-%% click on the little "Compile PDF" button
-%%
-%% Note that you should probably run the above a few times because
-%% LaTeX generally takes a few runs to get reference numbers right.
+\documentclass{article}
 
-\documentclass[article]{jss}
-%% need no \usepackage{Sweave.sty}
+\usepackage{Sweave}
+\usepackage{cite}
+\usepackage{authblk}
 
@@
 \usepackage{hyperref}
+\usepackage{graphicx}
+
+\let\proglang=\textsf
 
@@
 \title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
-\author{
-James Baglama\\
+%%\author{
+%%James Baglama\\
+%%University of Rhode Island
+%%\And
+%%Michael Kane\\
+%%Yale University
+%%\AND
+%%Bryan Lewis\\
+%%Paradigm4, Inc.\\
+%%\And
+%%Alex Poliakov\\
+%%Paradigm4, Inc.
+%%}
+%%\Plainauthor{James Baglama, Michael Kane, Bryan Lewis, Alex Poliakov}
+%%\Plaintitle{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
+%%\Shorttitle{Efficient Thresholded Correlation}
+\author[1]{James Baglama \thanks{jbaglama@math.uri.edu}}
+\author[2]{Michael Kane \thanks{michael.kane@yale.edu}}
+\author[3]{Bryan Lewis \thanks{blewis@illposed.net}}
+\author[3]{Alex Poliakov \thanks{apoliakov@paradigm4.com}}
+\affil[1]{
 University of Rhode Island
-\And
-Michael Kane\\
+}
+\affil[2]{
 Yale University
-\AND
-Bryan Lewis\\
-Paradigm4, Inc.\\
-\And
-Alex Poliakov\\
+}
+\affil[3]{
 Paradigm4, Inc.
 }
-\Plainauthor{James Baglama, Michael Kane, Bryan Lewis, Alex Poliakov}
-\Plaintitle{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
-\Shorttitle{Efficient Thresholded Correlation}
-
-\Abstract{
+\renewcommand\Authands{ and }
+%\Abstract{
+\begin{document}
+\maketitle
+\begin{abstract}
 Efficiently computing a subset of a correlation matrix consisting of values
@@
 can efficiently compute thresholded subsets of very large correlation matrices.
-}
-\Keywords{thresholded correlation, SVD, \proglang{R}}
-\Plainkeywords{thresholded correlation, SVD, R}
-\Address{
-James Baglama, Chair of Mathematics\\
-Department of Mathematics\\
-University of Rhode Island\\
-Kingston, Rhode Island 02881\\
-E-mail: \email{jbaglama@math.uri.edu}\\
-URL: \url{http://www.math.uri.edu/~jbaglama}\\
-\\
-Michael Kane\\
-Yale University\\
-300 Geroge Street\\
-New Haven, Connecticut 06511\\
-E-mail: \email{michael.kane@yale.edu}\\
-\\
-Bryan Lewis\\
-Paradigm4, Inc.\\
-281 Winter Street, Suite 360\\
-Waltham, MA 02451\\
-E-mail: \email{blewis@illposed.net}\\
-URL: \url{http://illposed.net}\\
-\\
-Alex Poliakov\\
-Paradigm4, Inc.\\
-281 Winter Street, Suite 360\\
-Waltham, MA 02451\\
-E-mail: \email{apoliakov@paradigm4.com}\\
-}
-
-\begin{document}
+\end{abstract}
+%%\Keywords{thresholded correlation, SVD, \proglang{R}}
+%%\Plainkeywords{thresholded correlation, SVD, R}
+%%\Address{
+%%James Baglama,
+%%Chair of Mathematics\\
+%%Department of Mathematics\\
+%%University of Rhode Island\\
+%%Kingston, Rhode Island 02881\\
+%%E-mail: \email{jbaglama@math.uri.edu}\\
+%%URL: \url{http://www.math.uri.edu/~jbaglama}\\
+%%\\
+%%Michael Kane\\
+%%Yale University\\
+%%300 Geroge Street\\
+%%New Haven, Connecticut 06511\\
+%%E-mail: \email{michael.kane@yale.edu}\\
+%%\\
+%%Bryan Lewis\\
+%%Paradigm4, Inc.\\
+%%281 Winter Street, Suite 360\\
+%%Waltham, MA 02451\\
+%%E-mail: \email{blewis@illposed.net}\\
+%%URL: \url{http://illposed.net}\\
+%%\\
+%%Alex Poliakov\\
+%%Paradigm4, Inc.\\
+%%281 Winter Street, Suite 360\\
+%%Waltham, MA 02451\\
+%%E-mail: \email{apoliakov@paradigm4.com}\\
+%%}
 
@@
 of gene expression values may be used in the discovery of networks of genes
-relevant to particular biological processes \citep{genomics}.  Identification of
+relevant to particular biological processes \cite{genomics}.  Identification of
 highly-correlated pairs may be used in feature selection algorithms for machine
-learning applications \citep{ml1, ml2} and are important to time series and
-image processing applications \citep{timeseries, svd-similarity}.
+learning applications \cite{ml1, ml2} and are important to time series and
+image processing applications \cite{timeseries, svd-similarity}.
 
@@
 pruning rules based on the frequency of mutually co-occurring observations (rows)
-among vector pairs; see \citep{prune1, prune2} and the references therein.
+among vector pairs; see \cite{prune1, prune2} and the references therein.
 Related methods approximate a set containing the most highly-correlated
-pairs using hashing algorithms \citep{prune3}. A number of other methods arrive
+pairs using hashing algorithms \cite{prune3}. A number of other methods arrive
 at approximate sets of the most highly-correlated pairs using
 dimensionality-reduction techniques like the discrete Fourier
-transform \citep{timeseries}. The method of Wu, et al. \citep{svd-similarity},
+transform \cite{timeseries}. The method of Wu, et al. \cite{svd-similarity},
 conceptually very similar to our approach, uses the singular value
@@
 can easily be computed in parallel. See the reference \proglang{R}
-implementation \citep{sup} for an example.
+implementation \cite{sup} for an example.
 
@@
 The augmented implicitly restarted Lanczos bidiagonalization algorithm\break(IRLBA)
-of Baglama and Reichel \citep{irlba} efficiently computes truncated singular value
+of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value
 decompositions of large dense or sparse matrices. Reference implementations are
-available for \proglang{R} \citep{irlbar}, \proglang{Matlab} \citep{irlbam},
-and \proglang{Python} \citep{irlbap}.
+available for \proglang{R} \cite{irlbar}, \proglang{Matlab} \cite{irlbam},
+and \proglang{Python} \cite{irlbap}.
 
@@
 at most computing two additional vector inner products per matrix-vector product.
-The \proglang{R} implementation \citep{irlbar} of IRLBA includes arguments for centering and
+The \proglang{R} implementation \cite{irlbar} of IRLBA includes arguments for centering and
 scaling the input matrix.
@@
 We evaluated the algorithm using public gene expression and DNA methylation
-datasets from the \proglang{R} {\tt biclust} \citep{biclust} package and from
-the Cancer Genome Atlas \citep{gdac}, referred to as TCGA.  Most tests were
+datasets from the \proglang{R} {\tt biclust} \cite{biclust} package and from
+the Cancer Genome Atlas \cite{gdac}, referred to as TCGA.  Most tests were
 performed on a desktop PC with a single AMD A10-7850K 3.7$\,$GHz Athlon
@@
 compare its output with brute force results. The example uses the {\tt
-EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
+EisenYeast} data from the {\tt biclust} \proglang{R} package \cite{biclust}.
 The data consist of 80 experiments involving the
@@
 Human Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor
-samples obtained from the Broad GDAC Cancer Genome Atlas dashboard \citep{gdac}.
-See the \proglang{R} package implementation \citep{sup} for examples of
+samples obtained from the Broad GDAC Cancer Genome Atlas dashboard \cite{gdac}.
+See the \proglang{R} package implementation \cite{sup} for examples of
 efficiently converting the downloaded sample methylation data into an R matrix.
@@
 No code change was required to run in parallel on the cluster. The reference \proglang{R}
-{\tt tcor()} implementation uses \proglang{R}'s {\tt foreach} \citep{foreach} framework to
+{\tt tcor()} implementation uses \proglang{R}'s {\tt foreach} \cite{foreach} framework to
 run on arbitrary parallel configurations (``back ends''). We simply registered
-a {\tt doRedis} \citep{doredis} parallel back end with 16 \proglang{R} workers per computer
+a {\tt doRedis} \cite{doredis} parallel back end with 16 \proglang{R} workers per computer
 (64 total \proglang{R} worker processes) prior to invoking {\tt tcor()}.
@@
 subspaces applies to distance metrics similarly to correlation, see
-\citep{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
+\cite{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
 Euclidean distance between two vectors in terms of singular basis vectors and
@@
 can be extended to other distance metrics. The reference \proglang{R}
-implementation \citep{sup} includes functions for thresholded distance and
+implementation \cite{sup} includes functions for thresholded distance and
 correlation.
@@
 correlated vector pairs for a given constant N similarly to the work of
-Xiong et al. \citep{prune2}.
+Xiong et al. \cite{prune2}.
 
@@
 \bibliography{bibliography}{}
+\bibliographystyle{plain}
 

### 2015-12-22 threshold/draft.Rnw
Clarified discussion of pruning efficiency, indicating t=0.99

@@
 chosen in practice such that $p<<k$. Setting $p=10$ for the example shown in
-Figure \ref{fig1} prunes all but $12,924$ possible vector pairs out of a total
-$6221(6221 - 1) = 193,47,310$ with $\ell=1,715$.
+Figure \ref{fig1} with $t=0.99$ prunes all but $149$ possible vector pairs out of a total
+$6221(6221 - 1) = 193,47,310$ with $\ell=787$.
 

### 2015-12-22 threshold/draft.tex
Clarified discussion of pruning efficiency, indicating t=0.99

@@
 
-% latex table generated in R 3.2.3 by xtable 1.8-0 package
-% Tue Dec 22 09:08:06 2015
+% latex table generated in R 3.2.2 by xtable 1.7-4 package
+% Tue Dec 22 14:32:06 2015
 \begin{table}[ht]

### 2015-12-24 threshold/arxiv.tex
Typos
unfortunately caught after submission

@@
 chosen in practice such that $p<<k$. Setting $p=10$ for the example shown in
-Figure \ref{fig1} prunes all but $12,924$ possible vector pairs out of a total
-$6221(6221 - 1) = 193,47,310$ with $\ell=1,715$.
+Figure \ref{fig1} prunes all but $149$ possible vector pairs out of a total
+$6221(6221 - 1)/2 = 19,347,310$ with $\ell=787$.
 
@@
 to fit within given memory constraints. However, the brute force algorithm
-would still need to evaluate all of the nearly 160 billion possible vector pair
+would still need to evaluate all possible vector pair
 correlation values.)
@@
 values of 0.99 or more in about three hours on our test PC. A brute force
-algorithm would need to evaluate the threshold condition for more than 155
+algorithm would need to evaluate the threshold condition for more than 77
 billion vector pairs; the projection of Algorithm \ref{prune} limited the
@@
 \label{TCGA3}
-\end{table}From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
+\end{table}From a set of almost 80billion pairs of vectors, Algorithm \ref{prune} finds 913,601
 pairs with correlation values of 0.99 or more on a modest Linux

### 2015-12-24 threshold/draft.Rnw
Typos
unfortunately caught after submission

--- 2015-12-22 threshold/draft.Rnw
@@
 chosen in practice such that $p<<k$. Setting $p=10$ for the example shown in
-Figure \ref{fig1} prunes all but $12,924$ possible vector pairs out of a total
-$6221(6221 - 1) = 193,47,310$ with $\ell=1,715$.
+Figure \ref{fig1} with $t=0.99$ prunes all but $149$ possible vector pairs out of a total
+$6221(6221 - 1)/2 = 19,347,310$ with $\ell=787$.
 
@@
 @
-From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
+From a set of almost 80 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
 pairs with correlation values of 0.99 or more on a modest Linux

### 2015-12-24 threshold/draft.tex
Typos
unfortunately caught after submission

@@
 Figure \ref{fig1} with $t=0.99$ prunes all but $149$ possible vector pairs out of a total
-$6221(6221 - 1) = 193,47,310$ with $\ell=787$.
+$6221(6221 - 1)/2 = 19,347,310$ with $\ell=787$.
 
@@
 % latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Tue Dec 22 14:32:06 2015
+% Thu Dec 24 16:53:32 2015
 \begin{table}[ht]
@@
 \label{TCGA3}
-\end{table}From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
+\end{table}From a set of almost 80 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
 pairs with correlation values of 0.99 or more on a modest Linux

### 2016-03-02 threshold/draft.Rnw
Remove jss format

@@
 
-\documentclass[article]{jss}
-%% need no \usepackage{Sweave.sty}
+\documentclass{article}
 
+\usepackage{Sweave}
 \usepackage{amsmath}
 \usepackage{hyperref}
+\usepackage{cite}
+\usepackage{authblk}
+\usepackage{graphicx}
+\let\proglang=\textsf
 
@@
 
-\newcommand{\qed}{\nobreak \ifvmode \relax \else
-      \ifdim\lastskip<1.5em \hskip-\lastskip
-      \hskip1.5em plus0em minus0.5em \fi \nobreak
-      \vrule height0.75em width0.5em depth0.25em\fi}
-
 \title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
-\author{
-James Baglama\\
+%\author{
+%James Baglama\\
+%University of Rhode Island
+%\And
+%Michael Kane\\
+%Yale University
+%\AND
+%Bryan Lewis\\
+%Paradigm4, Inc.\\
+%\And
+%Alex Poliakov\\
+%Paradigm4, Inc.
+%}
+%\Plainauthor{James Baglama, Michael Kane, Bryan Lewis, Alex Poliakov}
+%\Plaintitle{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
+%\Shorttitle{Efficient Thresholded Correlation}
+\author[1]{James Baglama \thanks{jbaglama@math.uri.edu}}
+\author[2]{Michael Kane \thanks{michael.kane@yale.edu}}
+\author[3]{Bryan Lewis \thanks{blewis@illposed.net}}
+\author[3]{Alex Poliakov \thanks{apoliakov@paradigm4.com}}
+\affil[1]{
 University of Rhode Island
-\And
-Michael Kane\\
+}
+\affil[2]{
 Yale University
-\AND
-Bryan Lewis\\
-Paradigm4, Inc.\\
-\And
-Alex Poliakov\\
+}
+\affil[3]{
 Paradigm4, Inc.
 }
-\Plainauthor{James Baglama, Michael Kane, Bryan Lewis, Alex Poliakov}
-\Plaintitle{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
-\Shorttitle{Efficient Thresholded Correlation}
+\renewcommand\Authands{ and }
 
-\Abstract{
+%\Abstract{
+\begin{document}
+\maketitle
+\begin{abstract}
 Efficiently computing a subset of a correlation matrix consisting of values
@@
 can efficiently compute thresholded subsets of very large correlation matrices.
-}
-\Keywords{thresholded correlation, SVD, \proglang{R}}
-\Plainkeywords{thresholded correlation, SVD, R}
-\Address{
-James Baglama, Chair of Mathematics\\
-Department of Mathematics\\
-University of Rhode Island\\
-Kingston, Rhode Island 02881\\
-E-mail: \email{jbaglama@math.uri.edu}\\
-URL: \url{http://www.math.uri.edu/~jbaglama}\\
-\\
-Michael Kane\\
-Yale University\\
-300 Geroge Street\\
-New Haven, Connecticut 06511\\
-E-mail: \email{michael.kane@yale.edu}\\
-\\
-Bryan Lewis\\
-Paradigm4, Inc.\\
-281 Winter Street, Suite 360\\
-Waltham, MA 02451\\
-E-mail: \email{blewis@illposed.net}\\
-URL: \url{http://illposed.net}\\
-\\
-Alex Poliakov\\
-Paradigm4, Inc.\\
-281 Winter Street, Suite 360\\
-Waltham, MA 02451\\
-E-mail: \email{apoliakov@paradigm4.com}\\
-}
-
-\begin{document}
+\end{abstract}
+%\Keywords{thresholded correlation, SVD, \proglang{R}}
+%\Plainkeywords{thresholded correlation, SVD, R}
+%\Address{
+%James Baglama, Chair of Mathematics\\
+%Department of Mathematics\\
+%University of Rhode Island\\
+%Kingston, Rhode Island 02881\\
+%E-mail: \email{jbaglama@math.uri.edu}\\
+%URL: \url{http://www.math.uri.edu/~jbaglama}\\
+%\\
+%Michael Kane\\
+%Yale University\\
+%300 Geroge Street\\
+%New Haven, Connecticut 06511\\
+%E-mail: \email{michael.kane@yale.edu}\\
+%\\
+%Bryan Lewis\\
+%Paradigm4, Inc.\\
+%281 Winter Street, Suite 360\\
+%Waltham, MA 02451\\
+%E-mail: \email{blewis@illposed.net}\\
+%URL: \url{http://illposed.net}\\
+%\\
+%Alex Poliakov\\
+%Paradigm4, Inc.\\
+%281 Winter Street, Suite 360\\
+%Waltham, MA 02451\\
+%E-mail: \email{apoliakov@paradigm4.com}\\
+%}
 

### 2016-03-02 threshold/draft.tex
Updated the bibliography.
Remove jss format

@@
 
-% latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Thu Dec 24 16:53:32 2015
+% latex table generated in R 3.2.2 by xtable 1.8-2 package
+% Wed Mar  2 12:56:08 2016
 \begin{table}[ht]
 \centering
-{\small
+\begingroup\small
 \begin{tabular}{lrr}
@@
 \end{tabular}
-}
+\endgroup
 \caption{Gene expression results, threshold=0.95} 
@@
 
-% latex table generated in R 3.2.2 by xtable 1.7-4 package
-% Thu Dec 24 16:53:32 2015
+% latex table generated in R 3.2.2 by xtable 1.8-2 package
+% Wed Mar  2 12:56:08 2016
 \begin{table}[ht]
 \centering
-{\small
+\begingroup\small
 \begin{tabular}{lrrl}
@@
 \end{tabular}
-}
+\endgroup
 \caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99).

### 2016-09-04 threshold/draft.Rnw
new arxiv copy; Hero reference; generic formatting
Anti-correlation section

@@
 %%
-%% FIRST NOTE, you need the 'xtable' and 'biclust' packages installedi in R.
+%% FIRST NOTE, you need the 'xtable' and 'biclust' packages installed in R.
 %% Some of the examples will come from data sets in this package.
@@
 feasible for high-dimensional data in modern genomics and other applications.
-Considerable attention has been devoted to \emph{pruning
-methods} that cheaply prune away all but the most highly-correlated pairs of
-vectors.  For example the TAPER and related algorithms of Xiong et al. develop
-pruning rules based on the frequency of mutually co-occurring observations (rows)
-among vector pairs; see \cite{prune1, prune2} and the references therein.
-Related methods approximate a set containing the most highly-correlated
-pairs using hashing algorithms \cite{prune3}. A number of other methods arrive
-at approximate sets of the most highly-correlated pairs using
-dimensionality-reduction techniques like the discrete Fourier
-transform \cite{timeseries}. The method of Wu, et al. \cite{svd-similarity},
-conceptually very similar to our approach, uses the singular value
-decomposition (SVD) to find all pairs of close vectors with respect to a
-distance metric (instead of correlation).  Our method finds all of the most
-highly-correlated vector pairs with respect to a given threshold by pruning
-along an intuitive path of decreasing variance defined by the data and computed
-by a truncated SVD.
+Considerable attention has been devoted to \emph{pruning methods} that cheaply
+prune away all but the most highly-correlated pairs of vectors.  For example
+the TAPER and related algorithms of Xiong et al. develop pruning rules based on
+the frequency of mutually co-occurring observations (rows) among vector pairs;
+see \cite{prune1, prune2} and the references therein.  Related methods
+approximate a set containing the most highly-correlated pairs using hashing
+algorithms \cite{prune3}. A number of other methods arrive at approximate sets
+of the most highly-correlated pairs using dimensionality-reduction techniques
+like the discrete Fourier transform \cite{timeseries}. The method of Wu, et al.
+\cite{svd-similarity}, conceptually very similar to our approach, uses the
+singular value decomposition (SVD) to find all pairs of close vectors with
+respect to a distance metric (instead of correlation). Hero and
+Rajaratnam\cite{hero2011large} apply projections to the thresholded correlation
+problem with the additional objective of determining a reasonable threshold
+value; their method follows a probabilistic approach.  Our method finds all of
+the most highly-correlated vector pairs with respect to a given threshold by
+pruning along an intuitive path of decreasing variance defined by the data and
+computed by a truncated SVD.
 
@@
 i = order(s$v[,1])
-plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
+plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Permuted index", pch='.')
 t=0.99
@@
 
-\section{Extensions}\label{extensions}
+\section{Practical extensions}\label{extensions}
+
+This section presents a number of practical extensions to the main thresholded
+correlation idea, most of which are available in the reference R package
+implementation. Extensions include finding anti-correlated vectors, finding
+thresholded correlations between column vectors from two different matrices,
+and application to other distance metrics in place of correlation. All of the
+discussed extensions are simple to implement.
+
+\subsection{Anti-correlation}
+
+Some applications are interested in finding both highly correlated and
+anti-correlated vectors--that is, for a given threshold $t > 0$, find column
+vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
+Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
+similarly to Lemma \ref{sumlemma},
+\begin{equation*}
+\|a_i + a_j\|^2 =
+\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
+\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
+\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
+\end{equation*}
+where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
+Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
+anti-correlation threshold as follows:
+\begin{enumerate}
+\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
+$P$ that orders points in $\hat{v}_1$.
+\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
+\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
+\[
+\bigcup_{\tilde{i}=1}^\ell
+\left\{
+(\tilde{j},\tilde{j}+\tilde{i}) :
+\begin{array}{l}
+e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
+\\
+e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
+\end{array}
+\right\},
+\]
+where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
+absolute value of the matrix $D^{\tilde{i}}$.
+\end{enumerate}
+The modified step 3 produces a run length $\ell$ very likely to be larger than
+in the default case, and step 4 above requires more arithmetic. Thus, including
+anti-correlated vectors in the result is more expensive than simply finding
+only highly correlated vectors.
+
+\subsection{Thresholded correlation of columns between two different matrices}
+
+WRITE ME
+
+\subsection{Alternative metrics}
 

### 2016-09-04 threshold/draft.tex
new arxiv copy; Hero reference; generic formatting
Anti-correlation section

@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Wed Mar  2 12:56:08 2016
+% Sat Sep  3 20:12:12 2016
 \begin{table}[ht]
@@
 
-\section{Extensions}\label{extensions}
+\section{Practical extensions}\label{extensions}
+
+This section presents a number of practical extensions to the main thresholded
+correlation idea, most of which are available in the reference R package
+implementation. Extensions include finding anti-correlated vectors, finding
+thresholded correlations between column vectors from two different matrices,
+and application to other distance metrics in place of correlation. All of the
+discussed extensions are simple to implement.
+
+\subsection{Anti-correlation}
+
+Some applications are interested in finding both highly correlated and
+anti-correlated vectors--that is, for a given threshold $t > 0$, find column
+vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
+Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
+similarly to Lemma \ref{sumlemma},
+\begin{equation*}
+\|a_i + a_j\|^2 =
+\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
+\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
+\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
+\end{equation*}
+where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
+Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
+anti-correlation threshold as follows:
+\begin{enumerate}
+\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
+$P$ that orders points in $\hat{v}_1$.
+\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
+\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
+\[
+\bigcup_{\tilde{i}=1}^\ell
+\left\{
+(\tilde{j},\tilde{j}+\tilde{i}) :
+\begin{array}{l}
+e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
+\\
+e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
+\end{array}
+\right\},
+\]
+where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
+absolute value of the matrix $D^{\tilde{i}}$.
+\end{enumerate}
+The modified step 3 produces a run length $\ell$ very likely to be larger than
+in the default case, and step 4 above requires more arithmetic. Thus, including
+anti-correlated vectors in the result is more expensive than simply finding
+only highly correlated vectors.
+
+\subsection{Thresholded correlation of columns between two different matrices}
+
+Arbitrary vectors
+
+\subsection{Alternative metrics}
 

### 2017-06-06 threshold/draft.Rnw
Added generic parallel correlation example

@@
 The modified step 3 produces a run length $\ell$ very likely to be larger than
-in the default case, and step 4 above requires more arithmetic. Thus, including
-anti-correlated vectors in the result is more expensive than simply finding
-only highly correlated vectors.
+in the default case and step 4 above requires more arithmetic. Thus, including
+anti-correlated vectors in the result is more expensive than only finding
+highly correlated vectors.
 

### 2017-06-06 threshold/draft.tex
Added generic parallel correlation example

@@
 % latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Sat Sep  3 20:12:12 2016
+% Sat Sep  3 20:14:31 2016
 \begin{table}[ht]
@@
 The modified step 3 produces a run length $\ell$ very likely to be larger than
-in the default case, and step 4 above requires more arithmetic. Thus, including
-anti-correlated vectors in the result is more expensive than simply finding
+in the default case and step 4 above requires more arithmetic. Thus, including
+anti-correlated vectors in the result is more expensive than finding
 only highly correlated vectors.
@@
 
-Arbitrary vectors
+WRITE ME
 

### 2017-06-06 correlation.R

@@ -1,0 +1,129 @@
+# Example parallel computation of correlation matrix entries
+#
+# The test here does not do anything with the results. See the note
+# below within the foreach loop.
+#
+library('foreach')
+library('doSNOW')
+nodelist = c("n1","n2","n3","n4")
+
+m = 200     # Number of rows
+n = 100000  # Number of columns
+
+A = matrix(0,nrow=m, ncol=n)
+t1 = proc.time()
+set.seed(1)
+A[,] = runif(m*n)
+dt = (proc.time() - t1)[[3]]
+cat("Time to create random entries",dt,"\n")
+
+t1 = proc.time()
+x = apply(A,2,mean)
+s = apply(A,2,sd)
+for(j in 1:length(x)) A[,j] = A[,j] - x[j]
+
+# In practice, we'll want a copy of the transpose and the original matrix for
+# speed. A 1M x 1K matrix consumes only about 8GB memory.  So, with this
+# extra copy we need at least about 16GB for that sized problem.
+TA = t(A)
+# We also scale the columns of A (but not TA) in advance for efficiency's sake
+A = t(t(A)/s)
+dt = (proc.time() - t1)[[3]]
+cat("Time to center matrix and compute sd",dt,"\n")
+
+# Save A and distribute to other nodes
+t1 = proc.time()
+save(A, file="/dev/shm/A.mat")
+save(TA, file="/dev/shm/TA.mat")
+for(j in nodelist[-1]) {
+  cmd = paste("scp -r /dev/shm/A.mat ",j,":/dev/shm/",sep="")
+  system(cmd)
+  cmd = paste("scp -r /dev/shm/TA.mat ",j,":/dev/shm/",sep="")
+  system(cmd)
+}
+dt = (proc.time() - t1)[[3]]
+cat("Time to distribute data",dt,"\n")
+
+
+matinit = function()
+{
+  if(!exists("A",envir=globalenv())) {
+    load("/dev/shm/A.mat", envir=globalenv())
+    load("/dev/shm/TA.mat", envir=globalenv())
+  }
+}
+
+# The subcor function computes a subset of the correlation matrix of A, where
+# A is a matrix and assumes columns are centered and scaled by sd,
+# TA is the transpose of unscaled A (either matrix or big.matrix),
+# s is a vector of columnwise standard deviations of A,
[…]

### 2017-06-07 pearson.R
extra correlation codes

@@ -1,0 +1,20 @@
+# Example efficient Pearson's sample correlation matrix computation
+#
+# You should have R linked against a high-performance BLAS library for best
+# results. See XXX
+
+#' @param x input m by n data matrix
+#' @result n x n symmetric Pearson's sample correlation matrix
+#' @notes compare with cor(x, method="pearson")
+pearson = function(x)
+{
+  m = nrow(x)
+  z = .colMeans(x, nrow(x), ncol(x))
+  w = apply(x, 2, crossprod)
+  W = 1 / sqrt(w - m * z * z)
+  e = rep(1.0, m)
+  r = (2 * z) %*% crossprod(e, x)
+  r = m * tcrossprod(z) - r
+  ans = crossprod(x) + r
+  sweep((W * ans), 2, W, '*')
+}

### 2017-06-07 pearson.R
notest

@@ -1,7 +1,7 @@
 # Example efficient Pearson's sample correlation matrix computation
 #
 # You should have R linked against a high-performance BLAS library for best
-# results. See XXX
+# results. See http://illposed.net/r-on-linux.html
 
 #' @param x input m by n data matrix
 #' @result n x n symmetric Pearson's sample correlation matrix

### 2017-06-07 pearson.R
notes

@@ -6,6 +6,11 @@
 #' @param x input m by n data matrix
 #' @result n x n symmetric Pearson's sample correlation matrix
 #' @notes compare with cor(x, method="pearson")
+#' @examples
+#' x = matrix(rnorm(20), nrow=5, ncol=4)
+#' cor(x)
+#' pearson(x)
+#'
 pearson = function(x)
 {
   m = nrow(x)

### 2017-08-10 threshold/draft.Rnw
Wordsmiting and reorganization.

--- 2016-09-04 threshold/draft.Rnw
@@
 \begin{document}
+\SweaveOpts{concordance=TRUE}
 \maketitle
 \begin{abstract}
-Efficiently computing a subset of a correlation matrix consisting of values
-above a specified threshold is important to many practical applications.
-Real-world problems in genomics, machine learning, finance other applications
-can produce correlation matrices too large to explicitly form and tractably
-compute. Often, only values corresponding to highly-correlated vectors are of
-interest, and those values typically make up a small fraction of the overall
-correlation matrix. We present a method based on the singular value
-decomposition (SVD) and its relationship to the data covariance structure that
-can efficiently compute thresholded subsets of very large correlation matrices.
+Efficiently computing a subset of a correlation matrix consisting of values above 
+a specified threshold is important to many practical applications of large-scale 
+statistics and machine learning. Real-world problems in genomics, machine 
+learning, finance, and others can produce correlation matrices too large to 
+explicitly form and to tractably compute. However, in most applications, only 
+values corresponding to highly-correlated (or anti-correlated) vectors are of 
+interest, and those corresponding values typically make up a small fraction of 
+the overall correlation matrix. For this situation, we present a method based on 
+the singular value decomposition (SVD) and its relationship to the data 
+covariance structure for efficiently computing thresholded subsets of very large correlation matrices.
 \end{abstract}
@@
 
-Finding highly-correlated pairs among a large set of vectors is an important
-part of many applications. For instance, subsets of highly-correlated vectors
-of gene expression values may be used in the discovery of networks of genes
-relevant to particular biological processes \cite{genomics}.  Identification of
-highly-correlated pairs may be used in feature selection algorithms for machine
-learning applications \cite{ml1, ml2} and are important to time series and
+Finding highly-correlated pairs among a large set of vectors is an important part 
+of many statistics and machine learning applications. For instance, subsets of 
+highly-correlated vectors of gene expression values may be used in the discovery 
+of networks of genes relevant to particular biological processes \cite{genomics}. 
+Identification of highly-correlated pairs may be used in feature selection 
+algorithms for model fitting \cite{ml1, ml2} and are important to time series and 
 image processing applications \cite{timeseries, svd-similarity}.
 
-The number of correlation coefficients grows quadratically with the number of
-vectors. Simply computing all pairs of correlation coefficients and then
-filtering out coefficients below a given threshold may not be computationally
-feasible for high-dimensional data in modern genomics and other applications.
-Considerable attention has been devoted to \emph{pruning methods} that cheaply
-prune away all but the most highly-correlated pairs of vectors.  For example
-the TAPER and related algorithms of Xiong et al. develop pruning rules based on
-the frequency of mutually co-occurring observations (rows) among vector pairs;
-see \cite{prune1, prune2} and the references therein.  Related methods
-approximate a set containing the most highly-correlated pairs using hashing
-algorithms \cite{prune3}. A number of other methods arrive at approximate sets
-of the most highly-correlated pairs using dimensionality-reduction techniques
-like the discrete Fourier transform \cite{timeseries}. The method of Wu, et al.
-\cite{svd-similarity}, conceptually very similar to our approach, uses the
-singular value decomposition (SVD) to find all pairs of close vectors with
-respect to a distance metric (instead of correlation). Hero and
-Rajaratnam\cite{hero2011large} apply projections to the thresholded correlation
-problem with the additional objective of determining a reasonable threshold
-value; their method follows a probabilistic approach.  Our method finds all of
-the most highly-correlated vector pairs with respect to a given threshold by
-pruning along an intuitive path of decreasing variance defined by the data and
+Because the number of correlation coefficients grows quadratically with the 
+number of vectors, simply computing all pairs of correlation coefficients and 
+then filtering out coefficients below a given threshold may not be 
+computationally feasible for large-scale data applications. As a result, 
+considerable attention has been devoted to \emph{pruning methods} that cheaply 
+prune away all but the most highly-correlated pairs of vectors.  For example the 
+TAPER and related algorithms of Xiong et al. develop pruning rules based on the 
+frequency of mutually co-occurring observations (rows) among vector pairs; see 
+\cite{prune1, prune2} and the references therein.  Related methods approximate a 
+set containing the most highly-correlated pairs using hashing algorithms 
+\cite{prune3}. Other methods arrive at approximate sets of the most 
+highly-correlated pairs using dimensionality-reduction techniques like the 
+discrete Fourier transform \cite{timeseries}.
+
+The method of Wu, et al. \cite{svd-similarity}, conceptually very similar to our 
+approach, uses the singular value decomposition (SVD) to find all pairs of close 
+vectors with respect to a distance metric (instead of correlation). Similarly, 
+Hero and Rajaratnam\cite{hero2011large} apply projections to the thresholded 
+correlation problem with the additional objective of determining a reasonable 
+threshold value; their method follows a probabilistic approach.  Our method finds 
+all of the most highly-correlated vector pairs with respect to a given threshold 
+by pruning along an intuitive path of decreasing variance defined by the data and
 computed by a truncated SVD.
@@
 
-\section{Efficient pruning using truncated SVD}
+\section{Efficient pruning using truncated SVD} \label{sect:algo}
 
@@
 
+\section{Extensions}\label{extensions}
+
+This section presents a number of practical extensions to the main thresholded
+correlation idea, most of which are available in the reference R package
+implementation. Extensions include finding anti-correlated vectors, finding
+thresholded correlations between column vectors from two different matrices,
+and application to other distance metrics in place of correlation. All of the
+discussed extensions are simple to implement.
+
+\subsection{Anti-correlation}
+
+Some applications are interested in finding both highly correlated and
+anti-correlated vectors--that is, for a given threshold $t > 0$, find column
+vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
+Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
+similarly to Lemma \ref{sumlemma},
+\begin{equation*}
+\|a_i + a_j\|^2 =
+\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
+\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
+\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
+\end{equation*}
+where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
+Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
+anti-correlation threshold as follows:
+\begin{enumerate}
+\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
+$P$ that orders points in $\hat{v}_1$.
+\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
+\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
+\[
+\bigcup_{\tilde{i}=1}^\ell
+\left\{
+(\tilde{j},\tilde{j}+\tilde{i}) :
+\begin{array}{l}
+e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
+\\
+e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
+\end{array}
+\right\},
+\]
+where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
+absolute value of the matrix $D^{\tilde{i}}$.
+\end{enumerate}
+The modified step 3 produces a run length $\ell$ very likely to be larger than
+in the default case, and step 4 above requires more arithmetic. Thus, including
+anti-correlated vectors in the result is more expensive than simply finding
+only highly correlated vectors.
+
+\subsection{Thresholded correlation of columns between two different matrices}
+
+WRITE ME
+
+\subsection{Alternative metrics}
+
+The idea of evaluating threshold conditions by projection into low dimensional
+subspaces applies to distance metrics similarly to correlation, see
+\cite{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
+Euclidean distance between two vectors in terms of singular basis vectors and
+Algorithm \ref{prune} can be applied directly to the problem of finding all
+pairs of vectors that meet a thresholded Euclidean distance with small changes
+in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
+can be extended to other distance metrics. The reference \proglang{R}
+implementation \cite{sup} includes functions for thresholded distance and
+correlation.
+
+The ideas presented here can be adapted to efficiently find the top N most
+correlated vector pairs for a given constant N similarly to the work of
+Xiong et al. \cite{prune2}.
+
 \section{General matrices and fast truncated SVD}\label{irlba}
 
-We have so far assumed that the mean of each column vector of the $m\times n$
-matrix $A$ is zero and the columns are scaled to have unit norm, but we really
-want a method that works with general matrices $A$.  We also want a way to
-efficiently compute a relatively low-rank truncated SVD of potentially large
-matrices required by step 1 of Algorithm \ref{prune}.  And the truncated SVD
-method should be able to cheaply restart to compute additional singular vectors
-as required in step 5 of Algorithm \ref{prune}. Fortunately all desires are met
-by one algorithm.
+For Algorithm \ref{prune} to be used in practice, three challenges need to be 
+addressed. First, the algorithm requires an efficient method for computing a 
+relatively low-rank truncated SVD of potentially large matrices required by step 
+1. Second, the truncated SVD method should be able to cheaply restart to compute 
+additional singular vectors as required in step 5 of Algorithm \ref{prune}. 
+Third, while the algorithm assumes a standardized matrix, it should work for 
+general matrices. Fortunately, each of these challenges is met with a single 
+algorithm.
 
-The augmented implicitly restarted Lanczos bidiagonalization algorithm\break(IRLBA)
-of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value
-decompositions of large dense or sparse matrices. Reference implementations are
-available for \proglang{R} \cite{irlbar}, \proglang{Matlab} \cite{irlbam},
-and \proglang{Python} \cite{irlbap}.
+The first two challenges are addressed directly by the augmented implicitly 
+restarted Lanczos bidiagonalization algorithm\break(IRLBA) of Baglama and Reichel 
+\cite{irlba}. It efficiently computes truncated singular value decompositions of 
+either large dense or sparse matrices and the result from the $k$ rank approximati
+on can be efficiently used to create an approximation with higher rank. Reference 
+implementations are available for \proglang{R} \cite{irlbar}, \proglang{Matlab} 
+\cite{irlbam}, and \proglang{Python} \cite{irlbap}.
 
-We can relax the column mean and scale assumptions without introducing much
-additional computational or storage overhead as follows. Assume $A$ is an
-$m\times n$ real-valued matrix without any constant-valued columns. Let
-$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
-of $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
+The third challenge is addressed by relaxing the column mean and scale 
+assumptions without introducing significant additional computational or storage
+overhead as follows. Assume $A$ is an $m\times n$ real-valued matrix without any 
+constant-valued columns. Let $z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent 
+the vector of column means of $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
 be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
@@
 \]
-The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
-products. The main idea behind efficient application of IRLB to correlation
-problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
-the iterations, implicitly working with a scaled and centered matrix without
-forming it.  This comes at the cost of storing two additional length $n$ vectors and
+%The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
+%products. The main idea behind efficient application of IRLB to correlation
+%problems replaces 
+Matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ may be used in
+the iterations of the algrithm, implicitly working with a scaled and 
+centered matrix without explicit standardization. This comes at the cost 
+of storing two additional length $n$ vectors and
 at most computing two additional vector inner products per matrix-vector product.
-The \proglang{R} implementation \cite{irlbar} of IRLBA includes arguments for centering and
-scaling the input matrix.
+The \proglang{R} implementation \cite{irlbar} of IRLBA includes arguments for centering and scaling the input matrix.
+
+\section{Numerical experiments}\label{examples}
 
@@
 alter their threshold in some cases.
-
-\section{Numerical experiments}\label{examples}
 
@@
 
-\section{Practical extensions}\label{extensions}
-
-This section presents a number of practical extensions to the main thresholded
-correlation idea, most of which are available in the reference R package
-implementation. Extensions include finding anti-correlated vectors, finding
-thresholded correlations between column vectors from two different matrices,
-and application to other distance metrics in place of correlation. All of the
-discussed extensions are simple to implement.
-
-\subsection{Anti-correlation}
-
-Some applications are interested in finding both highly correlated and
-anti-correlated vectors--that is, for a given threshold $t > 0$, find column
-vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
-Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
-similarly to Lemma \ref{sumlemma},
-\begin{equation*}
-\|a_i + a_j\|^2 =
-\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
-\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
-\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
-\end{equation*}
-where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
-Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
-anti-correlation threshold as follows:
-\begin{enumerate}
-\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
-$P$ that orders points in $\hat{v}_1$.
-\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
-\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
-\[
-\bigcup_{\tilde{i}=1}^\ell
-\left\{
-(\tilde{j},\tilde{j}+\tilde{i}) :
-\begin{array}{l}
-e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
-\\
-e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
-\end{array}
-\right\},
-\]
-where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
-absolute value of the matrix $D^{\tilde{i}}$.
-\end{enumerate}
-The modified step 3 produces a run length $\ell$ very likely to be larger than
-in the default case, and step 4 above requires more arithmetic. Thus, including
-anti-correlated vectors in the result is more expensive than simply finding
-only highly correlated vectors.
-
-\subsection{Thresholded correlation of columns between two different matrices}
-
-WRITE ME
-
-\subsection{Alternative metrics}
-
-The idea of evaluating threshold conditions by projection into low dimensional
-subspaces applies to distance metrics similarly to correlation, see
-\cite{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
-Euclidean distance between two vectors in terms of singular basis vectors and
-Algorithm \ref{prune} can be applied directly to the problem of finding all
-pairs of vectors that meet a thresholded Euclidean distance with small changes
-in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
-can be extended to other distance metrics. The reference \proglang{R}
-implementation \cite{sup} includes functions for thresholded distance and
-correlation.
-
-The ideas presented here can be adapted to efficiently find the top N most
-correlated vector pairs for a given constant N similarly to the work of
-Xiong et al. \cite{prune2}.
-
 \section{Conclusion}\label{conclusion}

### 2017-08-10 threshold/draft.tex
Wordsmiting and reorganization.

--- 2016-09-04 threshold/draft.tex
@@
 \begin{document}
+\input{draft-concordance}
 \maketitle
 \begin{abstract}
-Efficiently computing a subset of a correlation matrix consisting of values
-above a specified threshold is important to many practical applications.
-Real-world problems in genomics, machine learning, finance other applications
-can produce correlation matrices too large to explicitly form and tractably
-compute. Often, only values corresponding to highly-correlated vectors are of
-interest, and those values typically make up a small fraction of the overall
-correlation matrix. We present a method based on the singular value
-decomposition (SVD) and its relationship to the data covariance structure that
-can efficiently compute thresholded subsets of very large correlation matrices.
+Efficiently computing a subset of a correlation matrix consisting of values above 
+a specified threshold is important to many practical applications of large-scale 
+statistics and machine learning. Real-world problems in genomics, machine 
+learning, finance, and others can produce correlation matrices too large to 
+explicitly form and to tractably compute. However, in most applications, only 
+values corresponding to highly-correlated (or anti-correlated) vectors are of 
+interest, and those corresponding values typically make up a small fraction of 
+the overall correlation matrix. For this situation, we present a method based on 
+the singular value decomposition (SVD) and its relationship to the data 
+covariance structure for efficiently computing thresholded subsets of very large correlation matrices.
 \end{abstract}
@@
 
-% latex table generated in R 3.2.2 by xtable 1.8-2 package
-% Sat Sep  3 20:12:12 2016
+% latex table generated in R 3.4.1 by xtable 1.8-2 package
+% Thu Aug 10 12:49:00 2017
 \begin{table}[ht]
@@
 
-\section{Practical extensions}\label{extensions}
-
-This section presents a number of practical extensions to the main thresholded
-correlation idea, most of which are available in the reference R package
-implementation. Extensions include finding anti-correlated vectors, finding
-thresholded correlations between column vectors from two different matrices,
-and application to other distance metrics in place of correlation. All of the
-discussed extensions are simple to implement.
-
-\subsection{Anti-correlation}
-
-Some applications are interested in finding both highly correlated and
-anti-correlated vectors--that is, for a given threshold $t > 0$, find column
-vectors $a_i$ and $a_j$ such that $|\mbox{cor}(a_i, a_j)| \ge t$.
-Note that $-\mbox{cor}(a_i, a_j) = \mbox{cor}(a_i, -a_j)$, and that
-similarly to Lemma \ref{sumlemma},
-\begin{equation*}
-\|a_i + a_j\|^2 =
-\sigma_1^2 (e_i^Tv_{1} + e_j^Tv_{1})^2 + 
-\sigma_2^2 (e_i^Tv_{2} + e_j^Tv_{2})^2 + \cdots + 
-\sigma_k^2 (e_i^Tv_{k} + e_j^Tv_{k})^2,
-\end{equation*}
-where $e_j, v_j, \sigma_j$ are defined as in Lemma \ref{sumlemma}.
-Then steps 2--4 of Algorithm \ref{prune} can be modified to include vector pairs meeting an
-anti-correlation threshold as follows:
-\begin{enumerate}
-\item[2.] Let $\hat{v}_1\in\mathcal{R}^{2n}=[v_1; -v_1]$ (concatenate the vectors). Compute permutation
-$P$ that orders points in $\hat{v}_1$.
-\item[3.] Compute $\ell$, the longest run of successive points in $P \hat{v}_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$. If $\ell > n-1$, then set $\ell = n-1$.
-\item[4.] Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
-\[
-\bigcup_{\tilde{i}=1}^\ell
-\left\{
-(\tilde{j},\tilde{j}+\tilde{i}) :
-\begin{array}{l}
-e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t) \mbox{ or }
-\\
-e_{\tilde{j}}^T(|D^{\tilde{i}}| P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
-\end{array}
-\right\},
-\]
-where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
-absolute value of the matrix $D^{\tilde{i}}$.
-\end{enumerate}
-The modified step 3 produces a run length $\ell$ very likely to be larger than
-in the default case, and step 4 above requires more arithmetic. Thus, including
-anti-correlated vectors in the result is more expensive than simply finding
-only highly correlated vectors.
-
-\subsection{Thresholded correlation of columns between two different matrices}
-
-Arbitrary vectors
-
-\subsection{Alternative metrics}
-
-The idea of evaluating threshold conditions by projection into low dimensional
-subspaces applies to distance metrics similarly to correlation, see
-\cite{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
-Euclidean distance between two vectors in terms of singular basis vectors and
-Algorithm \ref{prune} can be applied directly to the problem of finding all
-pairs of vectors that meet a thresholded Euclidean distance with small changes
-in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
-can be extended to other distance metrics. The reference \proglang{R}
-implementation \cite{sup} includes functions for thresholded distance and
-correlation.
-
-The ideas presented here can be adapted to efficiently find the top N most
-correlated vector pairs for a given constant N similarly to the work of
-Xiong et al. \cite{prune2}.
-
 \section{Conclusion}\label{conclusion}

### 2017-08-10 threshold/draft.Rnw
General Matrices is now it's own section.

@@
 
-\subsection{Thresholded correlation of columns between two different matrices}
+\subsection{Spearman correlation}
 
-WRITE ME
-
-\subsection{Alternative metrics}
-
+TODO: WRITE THIS
 The idea of evaluating threshold conditions by projection into low dimensional
@@
 correlation.
+
+\subsection{Most correlated pairs}
+
+TODO: WRITE THIS
 

### 2017-08-10 threshold/draft.tex
General Matrices is now it's own section.

@@
 % latex table generated in R 3.4.1 by xtable 1.8-2 package
-% Thu Aug 10 12:49:00 2017
+% Thu Aug 10 15:41:27 2017
 \begin{table}[ht]

### 2017-08-14 threshold/draft.tex
Updated draft following Mike's revision
Switching to the tex version only--no longer using Sweave (draft.tex is the working document now)

@@
-%% Instructions for processing this document into a PDF document
-%%
-%% FIRST NOTE, you need the 'xtable' and 'biclust' packages installed in R.
-%% Some of the examples will come from data sets in this package.
-%% You can install it from the R console with, for example:
-%%
-%% install.packages("biclust",repos="http://cran.case.edu")
-%% install.packages("xtable",repos="http://cran.case.edu")
-%%
-%% Now you can prepare the PDF of the paper with any of:
-%%
-%% From a shell:
-%% R CMD Sweave --pdf draft.Rnw
-%%
-%% From an R prompt:
-%% Sweave("draft.Rnw")
-%% tools::texi2pdf("draft.tex")
-%%
-%% From RStudio:
-%% click on the little "Compile PDF" button
-%%
-%% Note that you should probably run the above a few times because
-%% LaTeX generally takes a few runs to get reference numbers right.
-
 \documentclass{article}
@@
 then filtering out coefficients below a given threshold may not be 
-computationally feasible for large-scale data applications. As a result, 
+computationally feasible for high-dimensional data. As a result, 
 considerable attention has been devoted to \emph{pruning methods} that cheaply 
@@
 and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
-diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
-later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$
-to be the vector of the first $p\le k$ singular values along the diagonal of
-$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
+diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.
+%%For
+%%later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$
+%%to be the vector of the first $p\le k$ singular values along the diagonal of
+%%$\Sigma$.
+Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
 $v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
@@
 super-diagonal, and zeros elsewhere.  Then $D^1P v_1$ consists of the
-differences of adjacent entries in $P v_1$.  Entries such that $\sigma_1^2 (D^1
+differences of sequential entries in $P v_1$.  Entries such that $\sigma_1^2 (D^1
 P v_1)^2 > 2(1-t)$ correspond to pairs of vectors that do not meet the
@@
 Even a 1-d projected interval distance threshold may prune many possible
-non-adjacent vector pairs with respect to the ordering defined by $P$ as
+non-sequential vector pairs with respect to the ordering defined by $P$ as
 illustrated by the example in Figure \ref{fig1}.  However, this observation is
-unlikely to rule out many adjacent pairs of vectors in most problems.  For
-instance, the maximum distance between adjacent points shown in the example in
+unlikely to rule out many sequential pairs of vectors in most problems.  For
+instance, the maximum distance between sequential points shown in the example in
 Figure \ref{fig1} is less than $\sqrt{2(1-t)}/\sigma_1$, and no pruning of
-adjacent pairs of vectors with respect to the ordering $P$ occurs.
+sequential pairs of vectors with respect to the ordering $P$ occurs.
 
 Including more terms from Equation \ref{vdist} increases our ability to prune
-adjacent pairs of vectors below the correlation threshold with respect to the
+sequential pairs of vectors below the correlation threshold with respect to the
 ordering defined by $P$.  Including only one more term finds that 96\% of
-adjacent vectors fall below the correlation threshold of $0.99$ in the
+sequential vectors fall below the correlation threshold of $0.99$ in the
 example shown in Figure \ref{fig1} and are pruned by identifying indices such
@@
 where $\tilde{j}=1,2,\ldots,n-\tilde{i}$.
-\item If there are judged to be  ``too many'' candidate pairs, increase $p$,
+\item If there are judged to be  too many candidate pairs, increase $p$,
 compute $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
@@
 \end{algorithm}
-One can cheaply estimate whether there are ``too many'' candidate pairs for a
-given dimension $p$ in step 5 by evaluating step 4 for only adjacent pairs of
+One can cheaply estimate whether there are too many candidate pairs for a
+given dimension $p$ in step 5 by evaluating step 4 for only sequential pairs of
 vectors, $\tilde{i}=1$.
@@
 
+\subsection{General matrices and fast truncated SVD}\label{irlba}
+
+Three practical challenges need to be addressed for Algorithm \ref{prune} to be
+used in practice:
+\begin{enumerate}
+\item The algorithm needs a computationally efficient way to compute
+a low-rank truncated SVD of potentially large sparse or dense matrices.
+\item The truncated SVD method should be able to cheaply restart to
+compute additional singular vectors as needed by step 5 of Algorithm
+\ref{prune}.
+\item Algorithm \ref{prune} as stated assumes a standardized matrix;
+it needs to be extended to efficiently apply to general matrices.
+\end{enumerate}
+%For Algorithm \ref{prune} to be used in practice, three challenges need to be 
+%addressed. First, the algorithm requires an efficient method for computing a 
+%relatively low-rank truncated SVD of potentially large matrices required by step 
+%1. Second, the truncated SVD method should be able to cheaply restart to compute 
+%additional singular vectors as required in step 5 of Algorithm \ref{prune}. 
+%Third, while the algorithm assumes a standardized matrix, it should work for 
+%general matrices. Fortunately, each of these challenges is met with a single 
+%algorithm.
+
+The first two challenges are addressed directly by the augmented implicitly
+restarted Lanczos bidiagonalization algorithm (IRLBA) of Baglama and Reichel
+\cite{irlba}. It efficiently computes truncated singular value decompositions
+of either large dense or sparse matrices and the result from an existing rank
+$k$ approximation can be efficiently extended to higher-rank approximations by
+restarting.  Reference implementations are available for \proglang{R}
+\cite{irlbar}, \proglang{Matlab} \cite{irlbam}, and \proglang{Python}
+\cite{irlbap}.
+
+The third challenge is addressed by relaxing the column mean and scale
+assumptions without introducing significant additional computational or storage
+overhead as follows. Assume $A$ is an $m\times n$ real-valued matrix without
+any constant-valued columns. Let $z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$
+represent the vector of column means of $A$,
+$w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector of squared column
+norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal matrix with diagonal
+entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let $e\in\mathcal{R}^m$ be a vector of all
+ones.  Then $(A - ez^T)W$ is a centered matrix with zero column means and
+scaled to have unit column norms, and
+\[
+\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
+\]
+%The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
+%products.
+IRLBA does not directly require the matrix $A$, but instead only needs to
+compute matrix-vector products with it.  The main idea behind efficient
+application of IRLBA to correlation problems replaces matrix-vector products of
+the form $Ax$ with $AWx - ez^TWx$, implicitly working with a scaled and
+centered matrix without explicit standardization. This comes at the cost of
+storing two additional length $n$ vectors and at most computing two additional
+vector inner products per matrix-vector product.  The \proglang{R}
+implementation \cite{irlbar} of IRLBA includes arguments for centering and
+scaling the input matrix.
+
+A na\"ive brute force thresholded correlation computation requires $O(n^2 m)$
+flops.  Let $p$ be the selected IRLBA dimension. Then Algorithm \ref{prune}
+requires $O(n p \ell)$ flops, where $\ell$ is the longest run of ordered
+entries in $v_1$ that meet the 1-d projected distance threshold
+$\sqrt{2(1-t)}/\sigma_1$, including the flops required by IRLBA and final
+thresholding in step 7.  The main computational savings of $\frac{nm}{lp}$
+depends on the rate of decay in the singular values of $A$, the subspace
+dimension $p$, and the desired threshold value. We remark that this estimate of
+computational savings is available already in step 3 of Algorithm \ref{prune}
+after only modest computational effort. Users might use this cheap estimate to,
+for example, alter their threshold in some cases.
+
 \section{Extensions}\label{extensions}
 
-This section presents a number of practical extensions to the main thresholded
-correlation idea, most of which are available in the reference R package
-implementation. Extensions include finding anti-correlated vectors, finding
-thresholded correlations between column vectors from two different matrices,
-and application to other distance metrics in place of correlation. All of the
-discussed extensions are simple to implement.
+A number of practical extensions apply to the main thresholded correlation
+idea, most of which are available in the reference R package implementation.
+Extensions include finding anti-correlated vectors, finding thresholded
+pairwise correlations between column vectors from two different matrices,
+application to other distance metrics in place of correlation, and, similarly
+to the work of Xiong et al. \cite{prune2}, efficiently finding the top N most
+correlated vector pairs for a given constant N.
+
+All of the discussed extensions are simple to implement. This section
+outlines key ideas behind the extensions.
+
+\subsection{Spearman correlation and application to other distance metrics}
+
+The idea of evaluating threshold conditions by projection into low dimensional
+subspaces applies directly to the Euclidean distance metric, see
+\cite{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
+Euclidean distance between two vectors in terms of singular basis vectors and
+Algorithm \ref{prune} can be applied directly to the problem of finding all
+pairs of vectors that meet a thresholded Euclidean distance with small changes
+in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
+can be extended to other distance metrics with increased computational cost.
+The reference \proglang{R} implementation \cite{sup} includes functions for
+thresholded distance and correlation.
+
+Spearman rank correlation may be found by first transforming the data to ranked
+values followed by thresholded correlation using Algorithm \ref{prune} on the
+ranked values.
 
@@
 \]
-where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates the entry-wise
-absolute value of the matrix $D^{\tilde{i}}$.
+where $\tilde{j}=1,2,\ldots,n-\tilde{i}$, and $|D^{\tilde{i}}|$ indicates
+the entry-wise absolute value of the matrix $D^{\tilde{i}}$.
 \end{enumerate}
@@
 
-\subsection{Spearman correlation}
+Very similarly, we can compute pairwise thresholded correlation between pairs
+of column vectors of two distinct matrices by a truncated SVD of a matrix
+formed by coulmn-wise concatenation of the two matrices. The resulting
+thresholded pairs are then grouped by original column location to consider only
+pairs consisting of column vectors from each original matrix.
 
-TODO: WRITE THIS
-The idea of evaluating threshold conditions by projection into low dimensional
-subspaces applies to distance metrics similarly to correlation, see
-\cite{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
-Euclidean distance between two vectors in terms of singular basis vectors and
-Algorithm \ref{prune} can be applied directly to the problem of finding all
-pairs of vectors that meet a thresholded Euclidean distance with small changes
-in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
-can be extended to other distance metrics. The reference \proglang{R}
-implementation \cite{sup} includes functions for thresholded distance and
-correlation.
+\subsection{Top $N$ most correlated pairs}
 
-\subsection{Most correlated pairs}
+A procedure that finds the top $N$ most correlated pairs for a fixed integer
+$N$ proceeds iteratively starting with a small initial value of $p$, for
+instance $p = 0.999$.  Recall that the maximum number of pairs meeting a
+threshold is available by step 4 of Algorithm \ref{prune}.  Then:
 
-TODO: WRITE THIS
+\begin{enumerate}
+\item[a.] Reduce $p$ and repeat steps 1 -- 4 of Algorithm \ref{prune} until
+the maximum number of pairs meething the threshold $p$ is greater than $N$.
+\item[b.] Complete steps 5 -- 7 of Algorithm \ref{prune}. If the number of
+pairs meeting the threshold is less than $N$, reduce $p$ and goto step a.
+\end{enumerate}
 
-The ideas presented here can be adapted to efficiently find the top N most
-correlated vector pairs for a given constant N similarly to the work of
-Xiong et al. \cite{prune2}.
-
-\section{General matrices and fast truncated SVD}\label{irlba}
-
-For Algorithm \ref{prune} to be used in practice, three challenges need to be 
-addressed. First, the algorithm requires an efficient method for computing a 
-relatively low-rank truncated SVD of potentially large matrices required by step 
-1. Second, the truncated SVD method should be able to cheaply restart to compute 
-additional singular vectors as required in step 5 of Algorithm \ref{prune}. 
-Third, while the algorithm assumes a standardized matrix, it should work for 
-general matrices. Fortunately, each of these challenges is met with a single 
-algorithm.
-
-The first two challenges are addressed directly by the augmented implicitly 
-restarted Lanczos bidiagonalization algorithm\break(IRLBA) of Baglama and Reichel 
-\cite{irlba}. It efficiently computes truncated singular value decompositions of 
-either large dense or sparse matrices and the result from the $k$ rank approximati
-on can be efficiently used to create an approximation with higher rank. Reference 
-implementations are available for \proglang{R} \cite{irlbar}, \proglang{Matlab} 
-\cite{irlbam}, and \proglang{Python} \cite{irlbap}.
-
-The third challenge is addressed by relaxing the column mean and scale 
-assumptions without introducing significant additional computational or storage
-overhead as follows. Assume $A$ is an $m\times n$ real-valued matrix without any 
-constant-valued columns. Let $z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent 
-the vector of column means of $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
-be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
-a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
-$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
-matrix with zero column means and scaled to have unit column norms, and
-\[
-\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
-\]
-%The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
-%products. The main idea behind efficient application of IRLB to correlation
-%problems replaces 
-Matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ may be used in
-the iterations of the algrithm, implicitly working with a scaled and 
-centered matrix without explicit standardization. This comes at the cost 
-of storing two additional length $n$ vectors and
-at most computing two additional vector inner products per matrix-vector product.
-The \proglang{R} implementation \cite{irlbar} of IRLBA includes arguments for centering and scaling the input matrix.
+The procedure then relies on the ability to cheaply restart IRLBA to expand an
+existing truncated SVD.
 
 \section{Numerical experiments}\label{examples}
-
-A na\"ive brute force thresholded correlation computation requires $O(n^2 m)$ flops.
-Let $p$ be the selected IRLBA dimension. Then Algorithm \ref{prune} requires
-$O(n p \ell)$ flops, where $\ell$ is the longest run of ordered entries in
-$v_1$ that meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
-including the flops required by IRLBA and final thresholding in step 7.  The
-main computational savings of $\frac{nm}{lp}$ depends on the rate of decay in
-the singular values of $A$, the subspace dimension $p$, and the desired
-threshold value. We remark that this estimate of computational savings is
-available already in step 3 of Algorithm \ref{prune} after only modest
-computational effort. Users might use this cheap estimate to, for example,
-alter their threshold in some cases.
 
@@
 \begin{Sinput}
-> threshold = 0.95
-> # Test system values:
-> t1=3.13
-> t2=1.75
-> m1=59.60
-> m2=1075.10
-> if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
-+ {
-+ library(tcor)
-+ library(doMC)     # local parallel computation
-+ library(biclust)  # for the BicatYeast data
-+ data(EisenYeast)  # in gene by sample orientation
-+ A = t(EisenYeast) # to correlate genes to genes
-+ # Register a four-core multicore parallel backend
-+ registerDoMC(4)
-+ 
-+ # Compute using tcor():
-+ m1  = sum(gc()[,2])  # Memory in megabytes currently in use
-+ t1  = proc.time()
-+ tx  = tcor(A, t=threshold, p=10)
-+ t1  = (proc.time() - t1)[3] # Time in seconds
-+ m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
-+ 
-+ # Fast brute force full correlation:
-+ m2  = sum(gc()[,2])
-+ t2  = proc.time()
-+ mu = colMeans(A)
-+ s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
-+ cs = scale(A, center=mu, scale=s)  # scale and center the data
-+ cx = crossprod(cs)                 # full correlation matrix
-+ cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
-+ pairs = which(cx >= threshold, arr.ind=TRUE)
-+ t2  = (proc.time() - t2)[3]
-+ m2  = sum(gc()[,6]) - m2
-+ }
+threshold = 0.95
+library(tcor)
+library(doMC)     # local parallel computation
+library(biclust)  # for the BicatYeast data
+data(EisenYeast)  # in gene by sample orientation
+A = t(EisenYeast) # to correlate genes to genes
+# Register a four-core multicore parallel backend
+registerDoMC(4)
+
+# Compute using tcor():
+m1  = sum(gc()[, 2])  # Memory in megabytes currently in use
+t1  = proc.time()
+tx  = tcor(A, t=threshold, p=10)
+t1  = (proc.time() - t1)[3] # Time in seconds
+m1  = sum(gc()[, 6]) - m1    # Peak excess memory use during the test
+
+# Fast brute force full correlation:
+m2  = sum(gc()[, 2])
+t2  = proc.time()
+mu = colMeans(A)
+s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
+cs = scale(A, center=mu, scale=s)  # scale and center the data
+cx = crossprod(cs)                 # full correlation matrix
+cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
+pairs = which(cx >= threshold, arr.ind=TRUE)
+t2  = (proc.time() - t2)[3]
+m2  = sum(gc()[, 6]) - m2
 \end{Sinput}
@@
 
-% latex table generated in R 3.4.1 by xtable 1.8-2 package
-% Thu Aug 10 15:41:27 2017
+% latex table generated in R 3.4.0 by xtable 1.8-2 package
+% Sat Aug 12 09:35:30 2017
 \begin{table}[ht]
@@
 
-% latex table generated in R 3.4.1 by xtable 1.8-2 package
-% Thu Aug 10 15:41:27 2017
+% latex table generated in R 3.4.0 by xtable 1.8-2 package
+% Sat Aug 12 09:35:32 2017
 \begin{table}[ht]

### 2017-08-14 threshold/draft.tex
Typos and other minor miscellaneous edits

@@
 \item Algorithm \ref{prune} as stated assumes a standardized matrix;
-it needs to be extended to efficiently apply to general matrices.
+it needs to be extended to general matrices.
 \end{enumerate}
@@
 \cite{irlba}. It efficiently computes truncated singular value decompositions
-of either large dense or sparse matrices and the result from an existing rank
+of either large dense or sparse matrices. And, the result from an existing rank
 $k$ approximation can be efficiently extended to higher-rank approximations by
@@
 $N$ proceeds iteratively starting with a small initial value of $p$, for
-instance $p = 0.999$.  Recall that the maximum number of pairs meeting a
+instance $p = 0.999$.  The maximum number of pairs meeting a
 threshold is available by step 4 of Algorithm \ref{prune}.  Then:
@@
 
-The procedure then relies on the ability to cheaply restart IRLBA to expand an
+The procedure relies on the ability to cheaply restart IRLBA to expand an
 existing truncated SVD.
@@
 \begin{Sinput}
-> devtools::install_github("bwlewis/tcor")
+devtools::install_github("bwlewis/tcor")
 \end{Sinput}
@@
 library(biclust)  # for the BicatYeast data
-data(EisenYeast)  # in gene by sample orientation
-A = t(EisenYeast) # to correlate genes to genes
-# Register a four-core multicore parallel backend
-registerDoMC(4)
+data(EisenYeast)
+A = t(EisenYeast) # (to correlate genes to genes)
+registerDoMC(4)   # register a four-core multicore parallel backend
 
@@
 
-No code change was required to run in parallel on the cluster. The reference \proglang{R}
-{\tt tcor()} implementation uses \proglang{R}'s {\tt foreach} \cite{foreach} framework to
-run on arbitrary parallel configurations (``back ends''). We simply registered
-a {\tt doRedis} \cite{doredis} parallel back end with 16 \proglang{R} workers per computer
-(64 total \proglang{R} worker processes) prior to invoking {\tt tcor()}.
+No code change was required to run in parallel on the cluster. The reference
+\proglang{R} {\tt tcor()} implementation uses \proglang{R}'s {\tt foreach}
+\cite{foreach} framework to run on arbitrary parallel configurations (``back
+ends''). We simply registered a {\tt doRedis} \cite{doredis} parallel back end
+with 16 \proglang{R} workers per computer (64 total \proglang{R} worker
+processes) prior to invoking {\tt tcor()}.
 
@@
 below. The cluster example computed the filtering step 7 of Algorithm
-\ref{prune} in parallel across the 64 remote \proglang{R} processes. That approach incurs,
-in this example, an additional memory overhead of about 16$\,$GB because each
-remote \proglang{R} process requires access to a copy of the data matrix $A$. More thrifty
-adaptations of the algorithm are possible and could, for example, share the
-memory required by $A$ between worker \proglang{R} processes on each computer.
+\ref{prune} in parallel across the 64 remote \proglang{R} processes. That
+approach incurs, in this example, an additional memory overhead because each
+remote \proglang{R} process requires access to a copy of the data matrix $A$.
+More thrifty adaptations of the algorithm are possible and could, for example,
+share the memory required by $A$ between worker \proglang{R} processes on each
+computer.
 
